---
title: "Lab 5 - Team 4 - Dynamic Linear Models"
author: "Nick Chambers, Emma Timmins-Schiffman, Miranda Mudge"
date: "May 11, 2023"
output: 
 html_document:
  code-folding: true
  toc: true
  toc_float: true
---

```{r setup, echo = FALSE, message = FALSE}
options(dplyr.summarise.inform = FALSE)
```


Load the data and look at the time series.
```{r}
library(MARSS)
library(ggplot2)
## get data
data(KvichakSockeye, package="atsalibrary")
SR_data <- KvichakSockeye
plot.ts(x=SR_data$brood_year, SR_data$spawners, ylab="Number of Spawners and Recruits", xlab='Time')
lines(x=SR_data$brood_year,SR_data$recruits, col='blue')

```

Subset the data to include the time range 1956-1998 so that there is spawner and recruit data for each time point.
```{r }
#look at 1956-1998
SR_56to98<-subset(SR_data, brood_year > 1955 & brood_year < 1999)

plot.ts(x=SR_56to98$brood_year, SR_56to98$spawners, ylim=c(min(SR_56to98$recruits), max(SR_56to98$recruits)),ylab="Number of Spawners and Recruits", xlab='Time')
lines(x=SR_56to98$brood_year,SR_56to98$recruits, col='blue')
```

Create the metric representing reproductive rate: log(Rt/St)
```{r}
#alpha is intercept and beta is slope
#log(Rt/St) = at - bSt + vt
#where at is a RW
#need to estimate a and b (as hidden states)
TT<-length(SR_56to98$brood_year)
SR_56to98$RtovSt<-log(SR_56to98$recruits/SR_56to98$spawners)
dat <- matrix(SR_56to98$RtovSt, nrow = 1)
dat.z<-zscore(dat)

plot(x=SR_56to98$brood_year, y=SR_56to98$RtovSt, type='l', ylab='log(Rt/St)', xlab='Time')
```

Create covariate variables
```{r}
#covariate/predictor variable
#start with pdo summer
pdosum <- SR_56to98$pdo_summer_t2
## z-score the CUI
pdosum_z <- matrix(zscore(pdosum), nrow = 1)
## number of regr params (slope + intercept)
m <- dim(pdosum_z)[1] + 1

# Now do winter
pdowin <- SR_56to98$pdo_winter_t2
## z-score the CUI
pdowin_z <- matrix(zscore(pdowin), nrow = 1)
## number of regr params (slope + intercept)
m <- dim(pdowin_z)[1] + 1 # same as previous m so don't need to change

```

# Part 1: Estimate 1 regression parameter
Estimate the alpha regression parameter (the intercept).
```{r}
B <- 'identity'  
U <- matrix(0, nrow = 1, ncol = 1)  ## 2x1; both elements = 0
Q <- matrix(list(0), 1, 1)  ## 2x2; all 0 for now
diag(Q) <- c("q.alpha")

Z<-"identity"
A <- matrix(0)  ## 1x1; scalar = 0
R <- matrix("r")  ## 1x1; scalar = r

## only need starting values for regr parameters
inits_list <- list(x0 = 0)

## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)

#DLM without covariates
dlm_1 <- MARSS(dat.z, inits = inits_list, model = mod_list, silent = TRUE)

#plot alpha(t)
plot(x=SR_56to98$brood_year, y=dlm_1$states, type='l', main = "Time series of alpha", xlab = "Year", ylab = "alpha parameter")
lines(x=SR_56to98$brood_year, y=zscore(SR_56to98$RtovSt), col='blue')

dlm_1$AICc
#121.0344

plot1 <- autoplot(dlm_1,  plot.type = "fitted.ytT")
plot2 <- autoplot(dlm_1,  plot.type = "model.resids.ytT")
plot3 <- autoplot(dlm_1,  plot.type = "acf.std.model.resids.ytT")

```
The autocorrelation at 10 years may indicate a PDO effect.

# Part 2 - Estimate 2 regression parameters
Estimate alpha (intercept) and beta (slope)
```{r}
#### Estimate 2 regression parameters
B <- diag(2)  ## 2x2; Identity
U <- matrix(0, nrow = 2, ncol = 1)  ## 2x1; both elements = 0
#Q <- matrix(list(0), 1, 1)
Q <- matrix(list(0), 2, 2)  ## 2x2; all 0 for now
diag(Q) <- c("q.alpha", "q.beta")

Z<-matrix(1, nrow = 1, ncol = 2)
A <- matrix(0)  ## 1x1; scalar = 0
R <- matrix("r")  ## 1x1; scalar = r

## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0), nrow = 2))

## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)

#DLM without covariates
dlm_2 <- MARSS(dat.z, inits = inits_list, model = mod_list, silent = TRUE)

# plot alpha(t)
plot(x=SR_56to98$brood_year, y=dlm_2$states[1,], type='l', main = "Time series of alpha with 2 regression parameters", xlab = "Year", ylab = "alpha parameter")
lines(x=SR_56to98$brood_year, y=zscore(SR_56to98$RtovSt), col='blue')

plota <- autoplot(dlm_2,  plot.type = "fitted.ytT")
plotb <- autoplot(dlm_2,  plot.type = "model.resids.ytT")
plotc <- autoplot(dlm_2,  plot.type = "acf.std.model.resids.ytT")
#AICc: 126.0491
```
When two regression parameters are used the model produces a fit that appears to revert to the mean and appears less overfit than model 1. 


Model a time-varying alpha (intercept) and static beta (slope, or, effect of spawners). This means the process variance (q) of beta will be 0.
```{r}
#model 
###### time-varying alpha; static beta ####
B <- diag(2)  ## 2x2; Identity
U <- matrix(0, nrow = 2, ncol = 1)  ## 2x1; both elements = 0

Q <- matrix(list(0), 2, 2)  ## 2x2; all 0 for now
diag(Q) <- list("q.alpha", 0)

Z<-matrix(1, nrow = 1, ncol = 2)
A <- matrix(0)  ## 1x1; scalar = 0
R <- matrix("r")  ## 1x1; scalar = r

## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0), nrow = 2))

## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)

#DLM without covariates
dlm_3 <- MARSS(dat.z, inits = inits_list, model = mod_list, silent = TRUE)

#get alpha and beta
states <- dlm_3$states
states <- t(states)

plot(states[,1],ylab='Stock productivity (alpha)', xlab='time')
abline(h=mean(states[,1]), col='magenta')

autoplot(dlm_3,  plot.type = "fitted.ytT")
autoplot(dlm_3,  plot.type = "model.resids.ytT")
autoplot(dlm_3,  plot.type = "acf.std.model.resids.ytT")
```

# Part 3 - Fit expanded model with covariate summer PDO 
The residuals show a significant lag at 10 years for each model. This suggests that we are not capturing a source of variation in our model, which is perhaps an effect of the PDO.
```{r}
B <- diag(3)  ## 3x3; Identity
U <- matrix(0, nrow = 3, ncol = 1)  ## 3x1; elements = 0


Z <- array(NA, c(1, 3, TT))  ## NxMxT; pdo is effecting alpha
Z[1, 1, ] <- pdosum_z
Z[1, 2, ] <- rep(1,TT)
Z[1, 3, ] <- rep(1,TT)

Q <- matrix(list(0), 3, 3)  ## 3x3; 
diag(Q) <- list("q.alpha", 0, "g.alpha")

A <- matrix("a")  ## 1x1; scalar = 0 # a matrix should be the intercept
R <- matrix("r")  ## 1x1; scalar = r

## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0,0), nrow = 3))

## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)

#DLM without covariates
dlm_4 <- MARSS(dat.z, inits = inits_list, model = mod_list, silent = TRUE)

# calculate the mean productivity
mean_prod.4 <- mean(dlm_4$states[1,])

plot(x=SR_56to98$brood_year, y=dlm_4$states[3,], type='l', main = "Time series of delta with summer PDO",xlab = "Year", ylab = "delta parameter")
lines(x=SR_56to98$brood_year, y=zscore(SR_56to98$RtovSt), col='blue')

autoplot(dlm_4,  plot.type = "fitted.ytT")
autoplot(dlm_4,  plot.type = "model.resids.ytT")
autoplot(dlm_4,  plot.type = "acf.std.model.resids.ytT")
```

# Part 4 - Fit expanded model with covariate winter PDO 
```{r}
# Model winter PDO instead of summer PDO

B <- diag(3)  ## 3x3; Identity
U <- matrix(0, nrow = 3, ncol = 1)  ## 3x1; elements = 0


Z <- array(NA, c(1, 3, TT))  ## NxMxT; pdo is effecting alpha
Z[1, 1, ] <- pdowin_z
Z[1, 2, ] <- rep(1,TT)
Z[1, 3, ] <- rep(1,TT)

Q <- matrix(list(0), 3, 3)  ## 3x3; 
diag(Q) <- list("q.alpha", 0, "g.alpha")

A <- matrix("a")  ## 1x1; scalar = 0 # a matrix should be the intercept
R <- matrix("r")  ## 1x1; scalar = r

## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0,0), nrow = 3))

## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)

#DLM without covariates
dlm_5 <- MARSS(dat.z, inits = inits_list, model = mod_list, silent = TRUE)

# calculate the mean productivity
mean_prod.5 <- mean(dlm_5$states[1,])

# plot time series of delta(t)
plot(x=SR_56to98$brood_year, y=dlm_5$states[3,], type='l', main = "Time series of delta with winter PDO", xlab = "Year", ylab = "delta parameter")
lines(x=SR_56to98$brood_year, y=zscore(SR_56to98$RtovSt), col='blue')

autoplot(dlm_5,  plot.type = "fitted.ytT")
autoplot(dlm_5,  plot.type = "model.resids.ytT")
autoplot(dlm_5,  plot.type = "acf.std.model.resids.ytT")
```

Models 4 and 5 seem to produce similar fits to the data, although they produce different estimates of productivity. The model with summer PDO results in a mean productivity of .04 while the model with winter PDO results in a similar but slightly higher mean productivity of .14. 

# Part 5 - Evaluate models
AICc comparison
```{r}
AIC1 <- dlm_1$AICc
AIC2 <- dlm_2$AICc
AIC3 <- dlm_3$AICc
AIC4 <- dlm_4$AICc
AIC5 <- dlm_5$AICc

print(c(AIC1, AIC2, AIC3, AIC4, AIC5))
```
AICc values suggests that the first and most simple model is the best for this data. 


Parameters for forecasting
```{r}
## get list of Kalman filter output
kf_out <- MARSSkfss(dlm_1)

## forecasts of regr parameters; 2xT matrix
eta <- kf_out$xtt1

#create parameters for forecasting
Z <- array(NA, c(1, TT))
Z[1,] <- rep(1, TT) 

## ts of E(forecasts)
fore_mean <- vector()
for(t in 1:TT) {
  fore_mean[t] <- Z[,t] %*% eta[, t, drop = FALSE]
}

## variance of regr parameters; 1x2xT array
Phi <- kf_out$Vtt1

## obs variance; 1x1 matrix
R_est <- coef(dlm_1, type="matrix")$R

## ts of Var(forecasts)
fore_var <- vector()
for(t in 1:TT) {
  tZ <- matrix(Z[, t], 1, 1) ## transpose of Z
  fore_var[t] <- Z[, t] %*% Phi[, , t] %*% tZ + R_est
}

```

Plot forecast
```{r}
years = SR_56to98$brood_year

par(mar = c(4, 4, 0.1, 0), oma = c(0, 0, 2, 0.5))
ylims=c(min(fore_mean - 2*sqrt(fore_var)), max(fore_mean+2*sqrt(fore_var)))
plot(years, t(dat), type = "p", pch = 16, ylim = ylims,
     col = "blue", xlab = "", ylab = "Logit(s)", xaxt = "n")
lines(years, fore_mean, type = "l", xaxt = "n", ylab = "", lwd = 3) #forecast mean estimate
lines(years, fore_mean+2*sqrt(fore_var)) #upper bound
lines(years, fore_mean-2*sqrt(fore_var)) #lower bound
axis(1, at = seq(1956, 1998, 1))
mtext("Brood year", 1, line = 3)
```
The data falls within the expected parameters of the model, and generally follows the expected mean. Overall, the forecast appears to underestimate the data, particularly when the trend is increasing. Overall, it is an acceptable fit. 

```{r}
## forecast errors
innov <- kf_out$Innov

## Q-Q plot of innovations
qqnorm(t(innov), main = "", pch = 16, col = "blue")
## add y=x line for easier interpretation
qqline(t(innov))
```
The fit of the Q-Q plot could be better, particularly around the middle, but the tails aren't too far off as could often be the case. This means the data may not be normally distributed, but further evaluations are required. 

```{r}
## p-value for t-test of H0: E(innov) = 0
t.test(t(innov), mu = 0)$p.value
```

The p-value is greater than 0.05 so we cannot reject the null hypothesis that E(e-sub-t) = 0. 

```{r}
## plot ACF of innovations
acf(t(innov), lag.max = 10)
```
ACF of the innovations reflects the data with a slight lag at 5 and 10. 

Overall, the simplest model appears to be the best fit for the data though not all model assumptions are met - the model could behave better. Forecasting shows that while the model is a decent fit, there are likely underlying factors that are causing it to fail some of the model evaluations like autocorrelation. There may perhaps be seasonality in the data worth investigating before applying this model to further data. 

# Part 6 - Consider other environmental factors

There appears to be a repeating autocorrelation of lag 5, which could suggest the presence of cycling in the data. The lag at 5 and 10 might suggest am impact of PDO on the population, but the model was not improved by the inclusion of PDO. Seasonality should be investigated further as a possible factor. It may also be important to consider predator-prey interactions or the impact of fishing in the area. The scale of the remaining residual suggests a large-scale covariate, possibly ENSO or NPGO. Sockeye generally mature at 3 or 4 years of age so this does not explain the 5 year lag, however there may be other biological reasons for this shorter lag.   

## Contributions

A preliminary meeting by all was held to discuss and begin the homework. ETS completed parts 1-2. NC completed parts 3-4. MM completed parts 5-6 and drafted report into R markdown. Report was finalized by all. 
