---
title: "Lab 4 - Hidden Markov Models"
author: "Person 1, Person 2, Person 3"
date: May 9, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

* Team 1: Dylan Hubl, Eric French, Zoe Rand
* Team 2: Maria Kuruvilla, Miranda Mudge
* Team 3: Nick Chambers, Madison Shipley, Karl Veggerby
* Team 4: Liz Elmstrom, Terrance Wang, Emma Timmins-Schiffman

Make sure to label your final write-up with "final" in the title so we know which one is the final one.

## Notes from Liz

Please pick a type of indicators, and develop a 2- or
3-state multivariate HMM. A few tips:

Emma -- 
We are going to fit a 2-state multivariate HMM to North and South Copepod biomass and Copepod richness.

Then we fit 3 state and compare

-   Assume all responses are Gaussian.

Liz --

Liz will try deep temp or upper temperature to see if this effects the steps 

-   You're welcome to include covariates (year? Climate variables?) --
    but fitting a simple model without covariates is also totally fine

Summarize the model you've created. Specifically,

Terrance -- 

-   Does it converge?

-   How many states seem to be most supported?

-   What are the transition probabilities?

-   Anything else interesting that you've discovered?

# General tasks

Each group has the same general tasks, but you will adapt them as you work on the data.

Please pick a type of indicators, and develop a 2- or 3-state multivariate HMM. 

# Methods

A few tips:

-   Assume all responses are Gaussian.

-   You're welcome to include covariates (year? Climate variables?) --
    but fitting a simple model without covariates is also totally fine

# Data

As part of the California Current Integrated Ecosystem Report, NOAA
scientists do annual updates of [stoplight charts for ecosystem
indicators](https://www.fisheries.noaa.gov/west-coast/science-data/ocean-conditions-indicators-trends).

We have included the `stoplight.csv` dataset for this week. One of the
columns divides indicators into groups (e.g. Local Physical, Local
Biological, etc). Please pick a type of indicators, and develop a 2- or
3-state multivariate HMM. A few tips:

Describe what plankton groups and covariates you used and indicate any temporal 
subsetting you chose. Also describe any standardization, centering or scaling 
that was used. 

## Load the data

One of the
columns divides indicators into groups (e.g. Local Physical, Local
Biological, etc). 

```{r load_data}
library(depmixS4)
library(ggplot2)
library(tidyverse)

stoplight <- read.csv(here::here("Lab-4", "stoplight.csv"))
```

## Wrangle the data

```{r wrangle_data}
## add some code here
## copepod richness, North/south copepod as response variables

#we will focus on copepod richness, N cop. biomass, S cop. biomass
#rows 9-11
cop.dat<-stoplight[9:11,]
cop.row<-cop.dat[,1]
cop.dat<-cop.dat[,3:26]
rownames(cop.dat)<-cop.row

#look at the data over time
cop.t<-t(cop.dat)
rows <- gsub("X","", rownames(cop.t))
row.names(cop.t)<-rows
cop.df<-data.frame(cop.t)
cop.df$Year<-row.names(cop.df)

plot(x=cop.df$Year, y=cop.df$Copepod_richness, type='l')
lines(x=cop.df$Year, y=cop.df$N_copepod, col='blue')
lines(x=cop.df$Year, y=cop.df$S_copepod, col='magenta')
```
```{r covariates}
library(janitor)

covar <- stoplight %>%
  filter(Ecosystem.Indicators %in% 
    c('Copepod_richness','N_copepod','S_copepod',"Deep_temperature","SST", "PDO_DecMarch"))%>%
  select(!Type)%>% 
  t() %>%
  as_tibble %>% 
  row_to_names(row_number = 1)%>%
  mutate_if(is.character, as.numeric)

covar$Year <- 1998:2021

head(covar)

```

# Results

Summarize the model you've created. Specifically,

-   Does it converge?

-   How many states seem to be most supported?

-   Plot the time series of estimated states. What does this mean?

-   What are the transition probabilities?

-   If included, what are the covariate effects? What do these mean?

-   Anything else interesting that you've discovered?

## Multivariate HMM - 2 state

```{r}
#building the 2 state model
set.seed(1)
mod.2st<-depmix(list(Copepod_richness ~ 1, N_copepod ~ 1, S_copepod ~ 1),
                nstates = 2,
                family=list(gaussian(), gaussian(),gaussian()),
                data=cop.df)
fitmod.2st <- fit(mod.2st)

# Transition matrix
summary(fitmod.2st, which = "transition")
# AIC
AIC(fitmod.2st)

```

Most probable to switch from state 2 back to state 1.

```{r 2 state plots}
#most probable states
prstates2 <- apply(posterior(fitmod.2st) [,c("S1", "S2")], 1, which.max)
plot(prstates2, type='b', xlab='Time', ylab='State')

#estimated data/fits
mu2 <- summary(fitmod.2st)[,1]
pred2 <- data.frame("year"=seq(min(cop.df$Year), max(cop.df$Year)), "fit" = mu2[prstates2])

cop.df$Year<-as.numeric(cop.df$Year)

ggplot(cop.df) + 
  geom_point(aes(x=Year, y=Copepod_richness), col="black", size=1.5) + 
  geom_point(aes(x=Year, y=N_copepod), col="blue", size=1.5) +
  geom_point(aes(x=Year, y=S_copepod), col="magenta", size=1.5) +
  geom_line(data=pred2, aes(year, fit)) + 
  ggtitle("Copepods - raw data and predictions") + 
  ylab("Richness or Biomass") + xlab("Year") + theme_bw()

```

Richness = black points... Does seem to follow some sort of state/regime shift
Others do not... 


## Multivariate HMM - 3 state

```{r 3 state model}

#3 state model
set.seed(314)
mod.3st<-depmix(list(Copepod_richness ~ 1, N_copepod ~ 1, S_copepod ~ 1),
                nstates = 3,
                family=list(gaussian(), gaussian(),gaussian()),
                data=cop.df)
fitmod.3st <- fit(mod.3st)

# Transition matrix
summary(fitmod.3st, which = "transition")
# AIC
AIC(fitmod.3st)

```
Most probable to switch from S1 to S2. 

Adding a state greatly improves the AIC. (But almost always does) 

```{r 3 state plots}
#most probable states, LIZ EDITED THIS
prstates3 <- apply(posterior(fitmod.3st) [,c("S1", "S2", "S3")], 1, which.max)
plot(prstates3, type='b', xlab='Time', ylab='State')

#estimated data
mu3 <- summary(fitmod.3st)[,1]
pred3 <- data.frame("year"=seq(min(cop.df$Year), max(cop.df$Year)), "fit" = mu3[prstates3])

ggplot(cop.df) + 
  geom_point(aes(x=Year, y=Copepod_richness), col="black", size=1.5) + 
  geom_point(aes(x=Year, y=N_copepod), col="blue", size=1.5) +
  geom_point(aes(x=Year, y=S_copepod), col="magenta", size=1.5) +
  geom_line(data=pred3, aes(year, fit)) + 
  ggtitle("Copepods - raw data and predictions") + 
  ylab("Richness or Biomass") + xlab("Year") + theme_bw()

```

## Multivariate HMM with covariates

** Tried PDO, deep temperature, and SST, but not coded here. SST == better model fit. 

```{r covar}

#building the 2 state model
set.seed(1)
mod.2covar <-depmix(list(Copepod_richness ~ 1, N_copepod ~ 1, S_copepod ~ 1),
                nstates = 2,# Could change this to three. Adding states almost always improves AICs
                family=list(gaussian(), gaussian(),gaussian()),
                transition = ~ scale(SST),
                data=covar)

fitmod.2covar <- fit(mod.2covar)
summary(fitmod.2covar, which = "transition")

aics <- c(AIC(fitmod.2st), AIC(fitmod.2covar))
aics

```

SST slightly improves the model fit of the two state model. Don't really know how to interpret the transition matrix... 

SST maybe has a larger effect on the transition of Copepod richness to state 2? So when SST is high, Copepod Richness increases?

```{r covar 3 state}
# 3 state model 
set.seed(1)
mod.3covar <-depmix(list(Copepod_richness ~ 1, N_copepod ~ 1, S_copepod ~ 1),
                 nstates = 3,
                 family=list(gaussian(), gaussian(),gaussian()),
                 transition = ~ scale(SST),
                 data=covar)

fitmod.3covar <- fit(mod.3covar)
summary(fitmod.3covar, which = "transition")

aics <- c(AIC(fitmod.3st), AIC(fitmod.3covar))
aics
#SST slightly improves the 2 state model AIC, no improvement to three state model

```
Adding covariate does not improve model fit to the three state model. 


```{r covariates plot}

#most probable states
prstates_covar <- apply(posterior(fitmod.2covar) [,c("S1", "S2")], 1, which.max)
plot(prstates_covar, type='b', xlab='Time', ylab='State')

#estimated data
mu_covar <- summary(fitmod.2covar)[,1]
pred_covar <- data.frame("year"=seq(min(covar$Year), max(covar$Year)), "fit" = mu_covar[prstates_covar])

ggplot(covar) + 
  geom_point(aes(x=Year, y=Copepod_richness), col="black", size=1.5) + 
  geom_point(aes(x=Year, y=N_copepod), col="blue", size=1.5) +
  geom_point(aes(x=Year, y=S_copepod), col="magenta", size=1.5) +
  geom_line(data=pred2, aes(year, fit), col = "orange")+
  geom_line(data=pred_covar, aes(year, fit)) + 
  ggtitle("Copepods - raw data and predictions") + 
  ylab("Richness or Biomass") + xlab("Year") + theme_bw()

## Fits are visually the same with and without covariates

```

# Discussion


# Team contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

