---
title: "Lab 4 - Hidden Markov Models"
author: "Liz Elmstrom, Terrance Wang, Emma Timmins-Schiffman"
date: May 9, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

Copepods and other zooplankton are responsive to oceanographic conditions that alter their physical environment and food sources. We hypothesize that 4 indicators of the copepod community - richness, N biomass, S biomass, and biological transition - will be significantly impacted by PDO transitions. If copepods are sensitive to PDO, we should see shifts in the copepod community on a ~decadal scale. Covariates related to the PDO should strengthen the model if there is a strong link between the PDO and copepods.
## Notes from Liz

Please pick a type of indicators, and develop a 2- or
3-state multivariate HMM. A few tips:

Emma -- 
We are going to fit a 2-state multivariate HMM to North and South Copepod biomass and Copepod richness.

Then we fit 3 state and compare

-   Assume all responses are Gaussian.

Liz --

Liz will try deep temp or upper temperature to see if this effects the steps 

-   You're welcome to include covariates (year? Climate variables?) --
    but fitting a simple model without covariates is also totally fine

Summarize the model you've created. Specifically,

Terrance -- 

Next steps -- Add in biological transition as another response variable. This time around, zscoring the response variables and re-run the models. 

-   Does it converge?

-   How many states seem to be most supported?

-   What are the transition probabilities?

-   Anything else interesting that you've discovered?

# Data

Copepods and other zooplankton are responsive to oceanographic conditions that alter their physical environment and food sources. We hypothesize that 4 indicators of the copepod community - richness, N biomass, S biomass, and biological transition - will be significantly impacted by PDO transitions. If copepods are sensitive to PDO, we should see shifts in the copepod community on a ~decadal scale. Covariates related to the PDO should strengthen the model if there is a strong link between the PDO and copepods.

Also describe any standardization, centering or scaling 
that was used. 

## Load the data

```{r load_data}
library(depmixS4)
library(ggplot2)
library(tidyverse)

stoplight <- read.csv(here::here("Lab-4", "stoplight.csv"))
```

## Wrangle the data

```{r wrangle_data}
cop.dat<-stoplight[9:12,]
cop.row<-cop.dat[,1]
cop.dat<-cop.dat[,3:26]
rownames(cop.dat)<-cop.row

#look at the data over time
cop.t<-t(cop.dat)
rows <- gsub("X","", rownames(cop.t))
row.names(cop.t)<-rows
cop.df<-data.frame(cop.t)
cop.df$Year<-row.names(cop.df)

# to make this tidyr happy
cop.df.long = cop.df %>% pivot_longer(-Year, names_to="indicator", values_to = "value") %>%
  mutate(Year=as.numeric(Year))

cop.df.long %>% group_by(indicator) %>% mutate(value=scale(value)) %>% 
  ggplot(aes(x=Year,y=value,color=indicator)) + geom_line() + theme_bw() + labs(y="Z score")
```

```{r covariates}
library(janitor)

covar <- stoplight %>%
  filter(Ecosystem.Indicators %in% 
    c('Copepod_richness','N_copepod','S_copepod',"Deep_temperature","SST", "PDO_DecMarch","Biological_transition"))%>%
  select(!Type)%>% 
  t() %>%
  as_tibble %>% 
  row_to_names(row_number = 1)%>%
  mutate_if(is.character, as.numeric)

covar$Year <- 1998:2021

# to make this tidyr happy
cv.df = covar %>% pivot_longer(-Year, names_to="variable", values_to="value") %>%
  # z scores
  group_by(variable) %>% mutate(z_score = scale(value)) %>% group_by(Year) %>%
  filter(variable %in% c("N_copepod","S_copepod","Copepod_richness","Biological_transition"))

# save this as the non z score version
big.df = cv.df %>%
  select(-z_score) %>%
  # repivot for data analysis 
  pivot_wider(names_from="variable", values_from="value")

# save as z score version
big.df.z = cv.df %>%
  select(-value) %>%
  # repivot for data analysis 
  pivot_wider(names_from="variable", values_from="z_score")
```

# Results

## Multivariate HMM - 2 state

```{r}
# loop through a few states and compare AIC
set.seed(123)
AIC_tbl = data.frame(nstate=0, AIC=0)
fit_list = NULL
for (i in 1:3){
mod<-depmix(list(Copepod_richness ~ 1, N_copepod ~ 1, S_copepod ~ 1,Biological_transition ~ 1),
                nstates = i,
                family=list(gaussian(), gaussian(),gaussian(),gaussian()),
                data=big.df)
fitmod <- fit(mod)
fit_list[[i]] = fitmod
# Transition matrix
# summary(fitmod, which = "transition")
# AIC
AIC_tbl[i,"nstate"] = i
AIC_tbl[i,"AIC"] = AIC(fitmod)
}

# do the same but with z scores
set.seed(123)
AIC_tbl_z = data.frame(nstate=0, AIC=0)
fit_list_z = NULL
for (i in 1:3){
mod<-depmix(list(Copepod_richness ~ 1, N_copepod ~ 1, S_copepod ~ 1,Biological_transition ~ 1),
                nstates = i,
                family=list(gaussian(), gaussian(),gaussian(),gaussian()),
                data=big.df.z)
fitmod <- fit(mod)
fit_list_z[[i]] = fitmod
# Transition matrix
# summary(fitmod, which = "transition")
# AIC
AIC_tbl_z[i,"nstate"] = i
AIC_tbl_z[i,"AIC"] = AIC(fitmod)
}

AIC_tbl
AIC_tbl_z
```

2 states seem like the most reasonable. Though a 3rd state lowers the AIC, adding more states usually lowers the AIC. We proceed with a 2 state HMM due to its interpretability (there are 2 states in the PDO fluctuation, our system of interest).

```{r 2 state plots}
# compare HMM with raw data and z score data
# RAW DATA
#most probable states
prstates2 <- apply(posterior(fit_list[[2]]) [,c("S1", "S2")], 1, which.max)
# plot(prstates2, type='b', xlab='Time', ylab='State')

#estimated data/fits
mu2 <- summary(fit_list[[2]])[,1]
pred2 <- data.frame("year"=seq(min(big.df$Year), max(big.df$Year)), "fit" = mu2[prstates2])

# Z SCORE DATA
#most probable states
prstates2_z <- apply(posterior(fit_list_z[[2]]) [,c("S1", "S2")], 1, which.max)
# plot(prstates2, type='b', xlab='Time', ylab='State')

#estimated data/fits
mu2_z <- summary(fit_list_z[[2]])[,1]
pred2_z <- data.frame("year"=seq(min(big.df$Year), max(big.df$Year)), "fit" = mu2_z[prstates2_z])

# Raw data
ggplot() + 
  geom_point(data=cv.df,aes(x=Year,y=value,color=variable)) +
  geom_line(data=pred2, aes(year, fit)) + 
  ggtitle("Copepods - raw data and predictions") + 
  ylab("Indicator Value") + xlab("Year") + theme_bw()

# Z score
ggplot() + 
  geom_point(data=cv.df,aes(x=Year,y=z_score,color=variable),shape=4,size=3) +
  geom_line(data=pred2_z, aes(year, fit), linetype = "dashed") + 
  ggtitle("Copepods - z-score data and predictions") + 
  ylab("Indicator Value") + xlab("Year") + theme_bw()

# Transition Tables
summary(fit_list[[2]], which = "transition")
summary(fit_list_z[[2]], which = "transition")

```

There seems to be 2 state system, where the first state is associated with high copepod richness and S copepod, and low N copepod The second state is associated with low copepod richness and S copepod, and high N copepod. Our analysis shows that z-scoring the indicator data shows the same state fits as the raw data results, albeit at different scales. Z-scores are only useful for plotting purposes. Transition probabilities are identical. Though z-scoring the data is helpful in visualizing the contrast among states, we proceed with raw data when adding covariates. 


## Multivariate HMM with covariates

** Tried PDO, deep temperature, and SST, but not coded here. SST was the better model fit. 

```{r covar}
#building the 2 state model
set.seed(123)
# model sel
cov.terms<-c("SST","Deep_temperature")
# Create a table of indicators as to whethr that predictor is included
mod.ind <- as.matrix(expand.grid(c(FALSE,TRUE),c(FALSE,TRUE)))
mod.tab <- data.frame(mod.ind)
names(mod.tab) <- cov.terms

# Define a column for storing AIC
mod.tab$AIC  <- NA
fit_covar_list = NULL
for (i in 1:nrow(mod.tab)){
  ind_on=as.matrix(mod.tab[i,1:length(cov.terms)])
  
  # Set which covariates should be included in formulas
  if (length(cov.terms[ind_on])==0){
    formula = ~1
  }else{
    formula =  paste("~",paste(cov.terms[ind_on],collapse="+"),sep="")
  }

  mod.2covar <-depmix(list(Copepod_richness ~ 1, N_copepod ~ 1, S_copepod ~ 1,Biological_transition ~ 1),
                nstates = 2,# Could change this to three. Adding states almost always improves AICs
                family=list(gaussian(), gaussian(),gaussian(),gaussian()),
                transition = eval(parse(text=formula)),
                data=covar)
  fit_covar_list[[i]]= fit(mod.2covar)
  
  mod.tab[i,"AIC"] = AIC((fit_covar_list[[i]]))
}
mod.tab
```
Inclusion of SST and Deep Temp as covariates for the transition matrix seems to improve the 2-state model. 


```{r covariates plot}
summary(fit_covar_list[[4]])

#most probable states
prstates_covar <- apply(posterior(fit_covar_list[[4]]) [,c("S1", "S2")], 1, which.max)
plot(prstates_covar, type='b', xlab='Time', ylab='State')

#estimated data
mu_covar <- summary(fit_covar_list[[4]])[,1]
pred_covar <- data.frame("year"=seq(min(covar$Year), max(covar$Year)), "fit" = mu_covar[prstates_covar])

ggplot(covar) + 
  geom_point(data=cv.df,aes(x=Year,y=z_score,color=variable),size=2,alpha=0.7) +
  geom_line(data=pred2, aes(year, scale(fit),color="w/o cov"))+
  geom_line(data=pred_covar, aes(year, scale(fit),color="w cov")) + 
  ggtitle("2 state model - with vs without covariates") + 
  ylab("Z score") + xlab("Year") + theme_bw()

## Fits almost the same with and without covariates
```

Adding covariates only slightly changes the fits of states, but since the AIC is decreased by < 10 it is not worth including the covariates in the final model. The transition from state 1 to 2 only differ by 1 year around 1998 and 2006. 


# Discussion
Our best model was a 2 state model using N copepod, S copepod, copepod richness, and biological transition as indicators with SST and deep temperature as covariates. The 2nd state of high copepod richness (warm PDO), high S copepod biomass (winter dominant), low N copepod biomass (summer dominant), and late biological transition (winter>summer) confirms what we know about copepod ecology in the CA Current. The transition matrix suggests that the systems tends to remain in the 1st state on average. This 1st state tends to be associated with the cool PDO. 

Temperature covariates seem to be minmally important in explaining the state transitions. However, if covariates are included, warmer temperatures are generally associated with increased copepod diversity. 

# Team contributions

All team members contributed to experimental design and planning. ETS did the first draft of the 2- and 3- state HMMs and provided feedback and editing of the other team members' work.

