---
title: "Lab 4 - Hidden Markov Models"
author: "Maria Kuruvilla, Miranda Mudge"
date: May 9, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

# Data

Here we will examine a model fitting 3 different local copepod metrics - copepod richness, northern copepod biomass, and southern copepod biomass - across the entire available time series.
After spending some time looking at the time-series patterns prior to model fitting, we would also like to evaluate the effect of Pacific Decadal Oscillation (PDO) from December to March as a covariate on both the state and transition.
We will evaluate both a 2 and 3 state model with 1 covariate.

## Note

The variables we chose were already de-meaned.
We did not zscore the data.

## Load the data

```{r load_data, include=FALSE}
library(depmixS4)
library(tidyverse)
stoplight <- read.csv(here::here("Lab-4", "stoplight.csv"))
```

## Wrangle the data

```{r wrangle_data}

rownames(stoplight) <- stoplight[,1]

subset_data <- stoplight[c("Copepod_richness","N_copepod","S_copepod","PDO_DecMarch"),3:26]

#column names should be years
colnames(subset_data) <- seq(1998,2021)

par(mfrow = c(2,1), mai = c(0.5, 1, 0.1, 0.1), omi = c(0, 0, 0, 0))
plot(seq(1998,2021),subset_data["Copepod_richness",], type = "l", xlab = "years", ylab ="Copepod richness")
plot(seq(1998,2021),subset_data["N_copepod",], type = "l", xlab = "years", ylab ="N_copepod")
plot(seq(1998,2021),subset_data["S_copepod",], type = "l", xlab = "years", ylab ="S_copepod")
plot(seq(1998,2021),subset_data["PDO_DecMarch",], type = "l", xlab = "years", ylab ="PDO_DecMarch")

```

# Univariate test

We will start by looking at copepod richness.

```{r}
data <- data.frame(t(subset_data))
fitmod <- depmix(Copepod_richness ~1,
                 nstates = 2,
                 transition = ~ 1,
                 family = gaussian(),
                 data = data)
```

```{r}
set.seed(123)
fit1 <- fit(fitmod)
pars <- getpars(fit1)
matrix(pars[3:6], 2, 2, byrow = TRUE)
AIC1 <- AIC(fitmod)

```

Here we look at the effect of 1 covariate on copepod richness.

```{r}
fitmod_cov <- depmix(Copepod_richness ~1 + PDO_DecMarch,
                 nstates = 2,
                 transition = ~ 1,
                 family = gaussian(),
                 data = data)
```

```{r}
set.seed(123)
fit_cov <- fit(fitmod_cov)
pars_cov <- getpars(fit_cov)
matrix(pars_cov[3:6], 2, 2, byrow = TRUE)
AIC_cov <- AIC(fitmod_cov)
```

# Reframe multivariate dataset

Wrangle data to multivariate subset.

```{r}
subset_data_l <- subset_data
subset_data_l$ID <- rownames(subset_data_l)

subset_data_long <- subset_data_l %>%
  pivot_longer(!ID, names_to = "variable",
               values_to = "value")


subset_multi <-pivot_wider(subset_data_long, names_from = ID, values_from = value, names_repair = "check_unique")
```

# HMM on 2 states

Basic 2 state model

```{r}
set.seed(123)
mod2_1 = depmix(list(Copepod_richness ~ 1, N_copepod~1, S_copepod~1), 
             nstates = 2, 
             family = list(gaussian(),gaussian(),gaussian()),
             data=subset_multi)
fitmod2_1 = fit(mod2_1)
AIC2_1 = AIC(fitmod2_1)
```

```{r}
prstates = apply(posterior(fitmod2_1)[,c("S1","S2")], 1, which.max)
plot(prstates, type="b", xlab="Time", ylab="State")

```

```{r}
plot(subset_multi$variable, scale(apply(posterior(fitmod2_1) [,c("S1", "S2")], 1, which.max)),
     ylab="probability",
frame=FALSE, ylim = c(-1,4), type = "l")
points(subset_multi$variable,scale(subset_multi$Copepod_richness))

```

HMM with 2 states and covariate

```{r}
set.seed(123)
mod2_2 = depmix(list(Copepod_richness ~ PDO_DecMarch, N_copepod~ PDO_DecMarch, 
                     S_copepod~PDO_DecMarch), 
             nstates = 2, 
             family = list(gaussian(),gaussian(),gaussian()),
             data=subset_multi)
fitmod2_2 = fit(mod2_2)
AIC2_2 = AIC(fitmod2_2)
```

Covariate in the transition probabilty

```{r}
set.seed(123)
mod2_3 = depmix(list(Copepod_richness ~ 1, N_copepod ~ 1, 
                     S_copepod ~ 1), 
             nstates = 2, 
             transition = ~PDO_DecMarch,
             family = list(gaussian(),gaussian(),gaussian()),
             data=subset_multi)
fitmod2_3 = fit(mod2_3)
AIC2_3 = AIC(fitmod2_3)

```

# HMM on 3 states

Basic 3 state model

```{r}
set.seed(123)
mod3_1 = depmix(list(Copepod_richness ~ 1, N_copepod~1, S_copepod~1), 
             nstates = 3, 
             family = list(gaussian(),gaussian(),gaussian()),
             data=subset_multi)
fitmod3_1 = fit(mod3_1)
AIC3_1 <- AIC(fitmod3_1)
```

## Effect of covariate on transition parameters

3 state model with covariate effects on transition parameters

```{r}
set.seed(123)
mod3_2 = depmix(list(Copepod_richness ~ 1, N_copepod~1, S_copepod~1), 
             nstates = 3, 
             family = list(gaussian(),gaussian(),gaussian()),
             transition = ~scale(PDO_DecMarch),
             data=subset_multi)
fitmod3_2 = fit(mod3_2)

summary(fitmod3_2, which = "transition")
summary(fitmod3_2, which = "response")
AIC3_2 <- AIC(fitmod3_2)
```

3 state model with covariate effects on state parameters

```{r}
set.seed(123)
mod3_3 = depmix(list(Copepod_richness ~ 1 +PDO_DecMarch, 
                     N_copepod~1 +PDO_DecMarch,
                     S_copepod~1+PDO_DecMarch), 
             nstates = 3, 
             family = list(gaussian(),gaussian(),gaussian()),
             data=subset_multi)
fitmod3_3 = fit(mod3_3)

summary(fitmod3_3, which = "transition")
summary(fitmod3_3, which = "response")
AIC3_3 <- AIC(fitmod3_3)
```

3 state model with covariate effects on state and transition parameters

```{r}
set.seed(123)
mod3_4 = depmix(list(Copepod_richness ~ 1 +PDO_DecMarch, 
                     N_copepod~1 +PDO_DecMarch,
                     S_copepod~1+PDO_DecMarch), 
             nstates = 3, 
             family = list(gaussian(),gaussian(),gaussian()),
             transition = ~scale(PDO_DecMarch),
             data=subset_multi)
fitmod3_4 = fit(mod3_4)

summary(fitmod3_4, which = "transition")
summary(fitmod3_4, which = "response")
AIC3_4 <- AIC(fitmod3_4)
```

```{r}
AIC2_1
AIC2_2
AIC2_3
AIC3_1
AIC3_2
AIC3_3 # best model 
AIC3_4
```

```{r}
print(fitmod3_3)
```

Plotting the 3 state model

```{r}
prstates = apply(posterior(fitmod3_3)[,c("S1","S2", "S3")], 1, which.max)
plot(prstates, type="b", xlab="Time", ylab="State")
```



Based on the AIC values, the best model for assessing our subset of copepod data is the multivariate 3 state HMM model with covariate effects on the state parameters.

# Results / Discussion

Does it converge?

-   Yes, all our models converge.

How many states seem to be most supported?

-   An HMM model with 3 states is the most supported by AIC comparison but the fit looks better with a 2 state model.

Plot the time series of estimated states.

```{r warning=FALSE}
prstates = apply(posterior(fitmod3_3)[,c("S1","S2", "S3")], 1, which.max)
plot(subset_multi$variable,scale(apply(posterior(fitmod3_3)[,c("S1","S2", "S3")], 1, which.max)), xlab="Time", ylab="State", type = "b")
points(subset_multi$variable,scale(subset_multi$Copepod_richness), pch = 19)
```


What does this mean?
- The 3 state model shows that shows a smooth transition between states 1 - 3.
While the transition probabilities below show that it is possible for the model to jump from state 3 to state 1, the opposite is not true (probability of moving from state 1 to state 3 is zero).
This is reflected when plotting the likely states.
It is more likely to switch smoothly between the states over time, which makes sense when considering the influence of PDO - which indicates an overall warm or cool phase of ocean activity.
It is more likely that temperature fluctuations have smooth transitions than drastic jumps over a large time scale, and this is reflected in the 3 state model.
It also explains why the 2 state model with the same covariate effects also fits the data well - eliminating the middle state from a time series with smooth transitions shouldn't drastically effect the model.




What are the transition probabilities?
- Our transition matrix for the 3 state model: 



\begin{bmatrix}
0.66 &  0.33 & 0 \\
0.5 &  0 & 0.5 \\
0.4 &  0.4 & 0.2 \\
\end{bmatrix}



If included, what are the covariate effects?
What do these mean?
- We evaluated the effect of the climate covariate Pacific Decadal Oscillation (PDO) from December to March and found that PDO allowed the HMM model to have a better fit to the data when influencing the state process.

Anything else interesting that you've discovered?
- In both the 2 and 3 state HMM models, the covariate effects were best explained as state effects.

# Team contributions

Miranda and Maria both decided on the variables and covariates to test.
Maria fit the 2 state model and Miranda fit the 3 state model.
Miranda drafted the contributions into a final report.
Maria wrote the discussion and finalized the report.
