---
title: "Lab 4 - Hidden Markov Models"
author: "Maria Kuruvilla, Miranda Mudge"
date: May 9, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

# Data

Here we will examine a model fitting 3 different local copepod metrics - copepod richness, northern copepod biomass, and southern copepod biomass - across the entire available time series. After spending some time looking at the time-series patterns prior to model fitting, we would also like to evaluate the effect of Pacific Decadal Oscillation (PDO) from December to March as a covariate on both the state and transition. We will evaluate both a 2 and 3 state model with 1 covariate. 

## Note ?? 
Also describe any standardization, centering or scaling 
that was used. 


## Load the data

```{r load_data}
library(depmixS4)
stoplight <- read.csv(here::here("Lab-4", "stoplight.csv"))
```

## Wrangle the data

```{r wrangle_data}

rownames(stoplight) <- stoplight[,1]

subset_data <- stoplight[c("Copepod_richness","N_copepod","S_copepod","PDO_DecMarch"),3:26]

#column names should be years
colnames(subset_data) <- seq(1998,2021)

par(mfrow = c(2,1), mai = c(0.5, 0.7, 0.1, 0.1), omi = c(0, 0, 0, 0))
plot(seq(1998,2021),subset_data["Copepod_richness",], type = "l", xlab = "years", ylab ="Copepod richness")
plot(seq(1998,2021),subset_data["N_copepod",], type = "l", xlab = "years", ylab ="N_copepod")
plot(seq(1998,2021),subset_data["S_copepod",], type = "l", xlab = "years", ylab ="S_copepod")
plot(seq(1998,2021),subset_data["PDO_DecMarch",], type = "l", xlab = "years", ylab ="PDO_DecMarch")

```
# Univariate test

We will start by looking at copepod richness. 
```{r}
data <- data.frame(t(subset_data))
fitmod <- depmix(Copepod_richness ~1,
                 nstates = 2,
                 transition = ~ 1,
                 family = gaussian(),
                 data = data)
```

```{r}
set.seed(123)
fit1 <- fit(fitmod)
pars <- getpars(fit1)
matrix(pars[3:6], 2, 2, byrow = TRUE)
AIC1 <- AIC(fitmod)

```

Here we look at the effect of 1 covariate on copepod richness. 
```{r}
fitmod_cov <- depmix(Copepod_richness ~1 + PDO_DecMarch,
                 nstates = 2,
                 transition = ~ 1,
                 family = gaussian(),
                 data = data)
```

```{r}
set.seed(123)
fit_cov <- fit(fitmod_cov)
pars_cov <- getpars(fit_cov)
matrix(pars_cov[3:6], 2, 2, byrow = TRUE)
AIC_cov <- AIC(fitmod_cov)
```

# Reframe multivariate dataset

Wrangle data to multivariate subset. 
```{r}
subset_data_l <- subset_data
subset_data_l$ID <- rownames(subset_data_l)

subset_data_long <- subset_data_l %>%
  pivot_longer(!ID, names_to = "variable",
               values_to = "value")


subset_multi <-pivot_wider(subset_data_long, names_from = ID, values_from = value, names_repair = "check_unique")
```

# HMM on 2 states
Basic 2 state model
```{r}
set.seed(123)
mod2_1 = depmix(list(Copepod_richness ~ 1, N_copepod~1, S_copepod~1), 
             nstates = 2, 
             family = list(gaussian(),gaussian(),gaussian()),
             data=subset_multi)
fitmod2_1 = fit(mod2_1)
AIC2_1 = AIC(fitmod2_1)
```

```{r}
prstates = apply(posterior(fitmod2_1)[,c("S1","S2")], 1, which.max)
prstates2 <- apply(posterior(fit_list[[2]]) [,c("S1", "S2")], 1, which.max)

plot(prstates, type="b", xlab="Time", ylab="State")

```


```{r}
plot(ts(posterior(fitmod2_1, type="smoothing")[,1], start=1998, end = 2021),
     ylab="probability",
main="Posterior probability of state 1",
frame=FALSE, ylim = c(-1,8))
points(subset_multi$variable,subset_multi$Copepod_richness)

plot(ts(posterior(fitmod2_1, type="smoothing")[,2], start=c(1998,2021)),
     ylab="probability",
main="Posterior probability of state 1",
frame=FALSE)

```

HMM with 2 states and covariate
```{r}
set.seed(123)
mod2_2 = depmix(list(Copepod_richness ~ PDO_DecMarch, N_copepod~ PDO_DecMarch, 
                     S_copepod~PDO_DecMarch), 
             nstates = 2, 
             family = list(gaussian(),gaussian(),gaussian()),
             data=subset_multi)
fitmod2_2 = fit(mod2_2)
AIC2_2 = AIC(fitmod2_2)
```

Covariate in the transition probabilty
```{r}
set.seed(123)
mod2_3 = depmix(list(Copepod_richness ~ 1, N_copepod ~ 1, 
                     S_copepod ~ 1), 
             nstates = 2, 
             transition = ~PDO_DecMarch,
             family = list(gaussian(),gaussian(),gaussian()),
             data=subset_multi)
fitmod2_3 = fit(mod2_3)
AIC2_3 = AIC(fitmod2_3)
```
# HMM on 3 states

Basic 3 state model
```{r}
set.seed(123)
mod3_1 = depmix(list(Copepod_richness ~ 1, N_copepod~1, S_copepod~1), 
             nstates = 3, 
             family = list(gaussian(),gaussian(),gaussian()),
             data=subset_multi)
fitmod3_1 = fit(mod3_1)
AIC3_1 <- AIC(fitmod3_1)
```

## Effect of covariate on transition parameters
3 state model with covariate effects on transition parameters
```{r}
set.seed(123)
mod3_2 = depmix(list(Copepod_richness ~ 1, N_copepod~1, S_copepod~1), 
             nstates = 3, 
             family = list(gaussian(),gaussian(),gaussian()),
             transition = ~scale(PDO_DecMarch),
             data=subset_multi)
fitmod3_2 = fit(mod3_2)

summary(fitmod3_2, which = "transition")
summary(fitmod3_2, which = "response")
AIC3_2 <- AIC(fitmod3_2)
```

3 state model with covariate effects on state parameters
```{r}
set.seed(123)
mod3_3 = depmix(list(Copepod_richness ~ 1 +PDO_DecMarch, 
                     N_copepod~1 +PDO_DecMarch,
                     S_copepod~1+PDO_DecMarch), 
             nstates = 3, 
             family = list(gaussian(),gaussian(),gaussian()),
             data=subset_multi)
fitmod3_3 = fit(mod3_3)

summary(fitmod3_3, which = "transition")
summary(fitmod3_3, which = "response")
AIC3_3 <- AIC(fitmod3_3)
```

3 state model with covariate effects on state and transition parameters
```{r}
set.seed(123)
mod3_4 = depmix(list(Copepod_richness ~ 1 +PDO_DecMarch, 
                     N_copepod~1 +PDO_DecMarch,
                     S_copepod~1+PDO_DecMarch), 
             nstates = 3, 
             family = list(gaussian(),gaussian(),gaussian()),
             transition = ~scale(PDO_DecMarch),
             data=subset_multi)
fitmod3_4 = fit(mod3_4)

summary(fitmod3_4, which = "transition")
summary(fitmod3_4, which = "response")
AIC3_4 <- AIC(fitmod3_4)
```

```{r}
AIC2_1
AIC2_2
AIC2_3
AIC3_1
AIC3_2
AIC3_3 # best model 
AIC3_4
```

```{r}
print(fitmod3_3)
```

Based on the AIC values, the best model for assessing our subset of copepod data is the multivariate 3 state HMM model with covariate effects on the state parameters. 

# Results

-   Does it converge?
Our model does converge. 

How many states seem to be most supported?
- An HMM model with 3 states is the most supported. 

Plot the time series of estimated states. What does this mean?
- 
What are the transition probabilities?
- Our transition matrix: 
        toS1  toS2 toS3
fromS1 0.666 0.334  0.0
fromS2 0.500 0.000  0.5
fromS3 0.400 0.400  0.2

If included, what are the covariate effects? What do these mean?
- 

Anything else interesting that you've discovered?
- In both the 2 and 3 state HMM models, the covariate effects were best explained as state effects. 

Plot BIC
```{r}
plot(1:4,c(BIC(fitmod3_1),BIC(fitmod3_2),BIC(fitmod3_3),BIC(fitmod3_4)),ty="b")
```

# Discussion

Model with covariate effects on state parameters yields lowest AIC, suggesting this is the optimal 3 state model. 

```{r}
convert_probs = function(fit, coef_value = 0) {
  # extract coefficients
  p1 <- fit@transition[[1]]@parameters$coefficients
  p2 <- fit@transition[[2]]@parameters$coefficients
  
  # assemble transition matrix in mlogit space
  m <- matrix(0, 2, 2)
  m[1,] <- p1[1,] + p1[2,] * coef_value
  m[2,] <- p2[1,] + p2[2,] * coef_value
  
  # exponentiate
  m <- exp(m)
  
  # normalize
  m[1,] <- m[1,] / sum(m[1,])
  m[2,] <- m[2,] / sum(m[2,])
  return(m)
}

# So to look at the effects of the matrix when covariate 'corr' is 0,
convert_probs(fit, coef_value = 0)

# and compare these to
convert_probs(fit, coef_value = 1)


```


# code to plot 3 state model with covariate effects on state parameters
The first column of posterior(fit) will give you the viterbi states, which
represents the most probable sequence of states. I think you can use this
-- along with the means / intercepts for each of the time series to create
predicted values for each time series. Then I'd maybe bind them all into a
single dataframe, and use ggplot to show them all (using facetwrap() with
facets = variables?)
```{r}
# plot posterior state sequence for the 3-state model
state1 <- plot(ts(posterior(fitmod3_3, type="smoothing")[,1], # [] state #
        start=c(1998)))

```

##### code from class ##############
To start from the models we talked about in class, we start with the intercept only model / no covariates
```{r}
data(speed)
mod <- depmix(list(rt~1),
              data=speed,
              nstates=2,
              family=list(gaussian()))
fit <- fit(mod)

# look at this to see the estimated intercepts for each of the 2 states
summary(fit)

fit@response
fit@transition
```

Using the same dataset, we'll now add in covariates on the intercept:
```{r}
mod <- depmix(list(rt~corr),
              data=speed,
              nstates=2,
              family=list(gaussian()))
fit <- fit(mod)

# In this case, fit@transition is the same, but
fit@response
# gives 2 intercepts / slopes / sds. In other words, there are 2 regression models being fit -- either in state 1 or state 2
```

Now, we can add in covariates onto the transition probabilities. This is where things get trickier.
```{r}
mod <- depmix(list(rt~1),
              transition=~corr,
              data=speed,
              nstates=2,
              family=list(gaussian()))
fit <- fit(mod)

fit@response
# contains 2 intercepts, 2 sds as expected

attributes(fit)$transition
# gives the intercept and slope for the transition probabilities for each model in logit space.
```
Why are some set to 0 arbitrarily? These are the reference cases that we'll see below -- probabilities are normalized, so these aren't identifiable.

To look at the effect of covariates, if often helps to look at several levels of the covariate. I put together this little function for this example (2 states, 1 covariate on transition probabilities). Note that the steps are (1) create transition matrix in logit space, and (2) back transform
```{r}
convert_probs = function(fit, coef_value = 0) {
  # extract coefficients
  p1 <- fit@transition[[1]]@parameters$coefficients
  p2 <- fit@transition[[2]]@parameters$coefficients
  
  # assemble transition matrix in mlogit space
  m <- matrix(0, 2, 2)
  m[1,] <- p1[1,] + p1[2,] * coef_value
  m[2,] <- p2[1,] + p2[2,] * coef_value
  
  # exponentiate
  m <- exp(m)
  
  # normalize
  m[1,] <- m[1,] / sum(m[1,])
  m[2,] <- m[2,] / sum(m[2,])
  return(m)
}

# So to look at the effects of the matrix when covariate 'corr' is 0,
convert_probs(fit, coef_value = 0)

# and compare these to
convert_probs(fit, coef_value = 1)


```


##### Address later #################################### 

## Example code to potentially try
This should eliminate need for set.seed by providing the best framework for the model
```{r}
best = 1.0e10
best_model = NA
for(i in 1:iter) {
  # fit the model
  fitmod = fit(mod)
  
  # check to see if this is the best solution?
  if(AIC(fitmod) < best) {
    best_model = fitmod
    best = AIC(fitmod)
  }
}

best(fitmod3_1)
```


# from notes
```{r}
#mu = summary(fitmod3_1)[,1]

#pred = data.frame("year"=seq(min(subset_multi$variable), 
#  max(subset_multi$variable)),
#  "fit" = mu[prstates])
#pred
```

# SO thread 
```{r}
init_mod <- depmix(response, data, nstates) ## no solid value
mod <- setpars(init_mod, getpars(fm))
forwardbackward(mod)
```

# Results

Summarize the model you've created. Specifically,

-   Does it converge?

-   How many states seem to be most supported?

-   Plot the time series of estimated states. What does this mean?

-   What are the transition probabilities?

-   If included, what are the covariate effects? What do these mean?

-   Anything else interesting that you've discovered?

# Discussion


# Team contributions



