---
title: "Lab 3 - Common trends in plankton data"
subtitle: "Dynamic Factor Analysis (DFA)"
author: "Eric, Madison, Maria"
date: April 20, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

------------------------------------------------------------------------

# Introduction

This lab focused on dynamic factor analysis, exploring plankton blooms in Lake Washington.

# Data

```{r load_data}
## load MARSS for data and analyses
library(MARSS)

## load the raw data (there are 3 datasets contained here)
data(lakeWAplankton, package = "MARSS")

## we want `lakeWAplanktonTrans`, which has been transformed
## so the 0's are replaced with NA's and the data z-scored
all_dat <- lakeWAplanktonTrans
```

# Methods

## Plankton Selection

**Which plankton taxa did you choose and how did you choose them?**

We selected two primary producers (Diatoms & Other Algae), two grazers (Daphnia & Cyclops), and one predator (Epischura). We wanted to consider potential interaction between species (primary producers, grazers, predators), and also select species that were data-rich for the majority of the time series.  

*Diatoms* (Primary Producer) dominated in early spring time (Phytoplankton) 

*Other Algae* (Primary Producer)  --\> Various algal species that grazers eat. Complete data set.

*Daphnia* (Grazer) --\> Eat diatoms bloom after diatoms (zooplankton)

*Cyclops* (Grazer) --\> Grazer taxa, eat phytoplankton 

*Epischura* (Predator) --\> Calanus Copepod, tend to peak after diatoms


**What time period did you examine and why?**

We are choosing years 1978 to 1985 since Daphnia data is more consistent after 1978 and we are able to capture an interesting trend in Total Phosphorus.

## Wrangle the data

```{r wrangle_data}
## add some code here

yr_frst <- 1978
yr_last <- 1985
init_dat <- all_dat[all_dat[, "Year"] >= yr_frst & 
                      all_dat[, "Year"] <= yr_last,]

# put the 5 response variables in the yt matrix, can be replace whenever
# This represents multiple trophic levels
# Primary prod: Diatoms & Other Algae
# Grazers: Daphnia & Cyclops
# Predators: Epischura
plank_taxa <- c("Diatoms","Other.algae","Daphnia","Cyclops","Epischura")
dat <- init_dat[,plank_taxa] %>%
  t()
## mean of each taxon
y_bar <- apply(dat, 1, mean, na.rm = TRUE)
## subtract the means
plank_dat <- dat - y_bar
## assign new column names
spp <- rownames(plank_dat)
rownames(plank_dat) <- spp

# create a covariates matrix
covnames <- c("Temp","TP","pH") #pH has missing data points so it will only work in certain date ranges
covar <- init_dat[,covnames] %>%
  t()
print(plank_dat[,1:5])

print(covar[,1:5])


```

```{r plot_data}


clr <- c("brown", "blue", "darkgreen", "darkred", "purple")

## initialize a counter
cnt <- 1

## set up plotting space & make plots
par(mfrow = c(nrow(plank_dat), 1), mai = c(0.5, 0.7, 0.1, 0.1), omi = c(0, 0, 0, 0))
mm = ncol(plank_dat)

for(i in spp){
  plot(plank_dat[i,],  bty = "L", xaxt = "n", pch = 16,
       xlab = "",
       ylab = "Abundance index", col=clr[cnt], type="b")
  axis(1, 12 * (0:mm) + 1, yr_frst + 0:mm)
  title(i)
  cnt <- cnt + 1
}


## set plot colors
clr <- c("brown", "blue", "darkgreen","purple")

## initialize a counter
cnt <- 1

## set up plotting space & make plots
par(mfrow = c(nrow(covar), 1), mai = c(0.5, 0.7, 0.1, 0.1), omi = c(0, 0, 0, 0))

for(i in row.names(covar)){
  plot(covar[i,],  bty = "L", xaxt = "n", pch = 16,
       xlab = "",
       ylab = "Value", col=clr[cnt], type="b")
  axis(1, 12 * (0:mm) + 1, yr_frst + 0:mm)
  title(i)
  cnt <- cnt + 1
}


```

## Initial Model Set-up

We examined models that considered three underlying state options: two, three, and four states. We also tested three observation error matrices, diagonal and equal, diagonal and unequal, and equal variance and covariance. We originally were going to include unconstrained but omitted it due to convergence issues. We wanted to explore assumptions for observation error given our uncertainty surrounding collection and counting methods for the data set. Finally, we considered all combinations of three possible covariates, temperature (Temp), total phosphorus (TP), and pH.

```{r model_params}
model.states <- 2:4
names(model.states) <- c("two_states","three_states","four_states")

# Also testing different observation error matrices
R.models <- c("diagonal and unequal","equalvarcov", "diagonal and equal")

# Also the covariates must be z-scored if the response variables are demeaned
d.models <- list( 
  # d1 = "zero", # No covariates
  d2 = zscore(t(init_dat[,"Temp"])), # Temperature
  d3 = zscore(t(init_dat[,"TP"])), # Total Phosphorus
  d4 = zscore(t(init_dat[,"pH"])), # ph
  d5 = zscore(t(init_dat[,c("Temp","TP")])),
  d6 = zscore(t(init_dat[,c("TP", "pH")])),
  d7 = zscore(t(init_dat[,c("Temp", "pH")])),
  d8 = zscore(t(init_dat[,c("Temp","TP", "pH")])) # Temperature and Total Phosphorus
)
names(d.models) <- c("Temp", "TP", "pH", "Temp and TP", "TP and pH", "Temp and pH", "Temp,TP, and pH")


# Setting fixed portion of mod list, some of these we can mess around with but we will already be testing 48 models and I didn't want to eat up a bunch of processing time
mod.list = list(
  A = "zero", # this is set to zero because the data has been demeaned
  Q = "identity", # Q is set to the default value, this can also be "diagonal and equal" or "diagonal and unequal"
  x0 = "zero" # x0 can be also be set to "unconstrained" or "unequal" it might be useful to try these later
)

#DFA form for mars model Q--reduces option to streamline runtime

# all other params are left to their default settings

```

We then set up for model runs for 63 model runs, looking at various combination of states, covariates (d) and observation error (R).

```{r MARSS_models, cache = TRUE}
# Added in another for loop to run through cycling options, be warned, this means there are 80 total models. It takes a long time to run
start.time <- Sys.time()


out.tab <- NULL
fits <- list()
for(i in model.states){
  for (j in 1:length(d.models)){
    for(R.model in R.models){
      fit.model = c(list(m=i, R=R.model), mod.list)
      
      #MARSS will throw an error if covariates are not a matrix, This will not pass the covariate arg if it does not exist
      #      if(j==1) {
      #      fit <- MARSS(plank_dat, model=fit.model, 
      #                form = "dfa", z.score = FALSE, silent = TRUE,
      #                control=list(maxit=2000, allow.degen=TRUE))
      #      }
      #      else {
      fit <- MARSS(plank_dat, model=fit.model, 
                   form = "dfa", z.score = FALSE, silent = TRUE,
                   control=list(maxit=800, allow.degen = TRUE), 
                   covariates = d.models[[j]])
      #      }
      
      out=data.frame(States=names(model.states)[i-1],d=names(d.models)[j],R=R.model, # 
                     logLik=fit$logLik, AICc=fit$AICc, num.param=fit$num.params,
                     num.iter=fit$numIter, converged=!fit$convergence,
                     stringsAsFactors = FALSE)
      out.tab=rbind(out.tab,out)
      fits=c(fits,list(fit))
      
    }
  }
}
end.time <- Sys.time()

elapsed.time <- round((end.time - start.time), 3)
print(elapsed.time)
min.AICc <- order(out.tab$AICc)
out.tab.1 <- out.tab[min.AICc, ]
out.tab.1
```

# Initial Results

The time for all models to fit was approximately 40 min, and the model that has the least AICc is the model with two states, diagonal and unequal errors and all 3 covariates. Many of the models did not converge, which could have been due to data being inconsistent with our model assumptions, insufficient data (we selected a data rich time period, so I'm not sure about this one), or not enough number of iterations for fitting. (MARSS Manual, Chapter 10).

### Matrix for best fit model 

What forms of models did you fit (ie, write them out in matrix form)?

\text{Best Model}:
Observation Model
$$
\begin{bmatrix}
y_1\\
y_2\\
y_3\\
y_4\\
y_5\\
\end{bmatrix}_t=
\begin{bmatrix}
z_{1,1} & z_{1,2}\\
z_{2,1} & z_{2,2}\\
z_{3,1} & z_{3,2}\\
z_{4,1} & z_{4,2}\\
z_{5,1} & z_{5,2}\\
\end{bmatrix}*
\begin{bmatrix}
x_1\\
x_2\\
\end{bmatrix}_t+
\begin{bmatrix}
d_{1,1} & d_{1,2} & d_{1,3}\\
d_{2,1} & d_{2,2} & d_{2,3}\\
d_{3,1} & d_{3,2} & d_{3,3}\\
d_{4,1} & d_{4,2} & d_{4,3}\\
d_{5,1} & d_{5,2} & d_{5,3}\\
\end{bmatrix}*
\begin{bmatrix}
Temp \\
TP \\
pH\\
\end{bmatrix}_t +
\begin{bmatrix}
v_1\\
v_2\\
v_3\\
v_4\\
v_5
\end{bmatrix}_t

$$
Process Model
$$
\begin{bmatrix}
x_1\\
x_2\\
\end{bmatrix}_t=
\begin{bmatrix}
1 & 0\\
0 & 1\\
\end{bmatrix}*
\begin{bmatrix}
x_1\\
x_2\\
\end{bmatrix}_{t-1}+
\begin{bmatrix}
w_1\\
w_2\\
\end{bmatrix}_t

$$

Process Error
$$

\text{Where }
\begin{bmatrix}
w_1\\
w_2\\
\end{bmatrix}
 \sim MVN
\begin{pmatrix}
\begin{bmatrix}
0\\
0\\
\end{bmatrix},
\begin{bmatrix}
1 & 0\\
0 & 1\\
\end{bmatrix}
\end{pmatrix}
$$
Observation Error 
$$
\text{Where }
\begin{bmatrix}
v_1\\
v_2\\
v_3\\
v_4\\
v_5\\
\end{bmatrix}
 \sim MVN
\begin{pmatrix}
\begin{bmatrix}
0\\
0\\
0\\
0\\
0\\
\end{bmatrix},
\begin{bmatrix}
r_1 & 0& 0& 0& 0\\
0 & r_2& 0& 0& 0\\
0 & 0& r_3& 0& 0\\
0 & 0& 0& r_4& 0\\
0 & 0& 0& 0& r_5\\
\end{bmatrix}
\end{pmatrix}
$$
## Diagnostics 

```{r factor_loadings}
# Rotating the model, plotting the states and loadings
best_fit <- fits[[19]]

# Get Z Estimations and rotate the data
## get the estimated ZZ
Z_est <- coef(best_fit, type = "matrix")$Z

## get the inverse of the rotation matrix
H_inv <- varimax(Z_est)$rotmat

## rotate factor loadings
Z_rot = Z_est %*% H_inv   

## rotate processes
proc_rot = solve(H_inv) %*% best_fit$states

N_ts <- nrow(plank_dat)
m = ncol(Z_est)
## plot labels
ylbl <- plank_taxa
w_ts <- seq(mm)

## set up plot area
layout(matrix(1:4, m, 2), widths = c(3,2))
par(mai = c(0.3, 0.3, 0.3, 0.1), omi = c(0, 0, 0, 0))

## plot the processes
for(i in 1:m) {
  ylm <- c(-1, 1) * max(abs(proc_rot[i,]))
  ## set up plot area
  plot(w_ts,proc_rot[i,], type = "n", bty = "L",
       ylim = ylm, xlab = "", ylab = "", xaxt = "n")
  ## draw zero-line
  abline(h = 0, col = "gray")
  ## plot trend line
  lines(w_ts, proc_rot[i,], lwd = 2)
  lines(w_ts, proc_rot[i,], lwd = 2)
  ## add panel labels
  mtext(paste("State",i), side = 3, line = 0.5)
  axis(1, 12 * (0:mm) + 1, yr_frst + 0:mm)
}
## plot the loadings
minZ <- 0
ylm <- c(-1, 1) * max(abs(Z_rot))
for(i in 1:m) {
  plot(x = c(1:N_ts)[abs(Z_rot[,i])>minZ],
       y = as.vector(Z_rot[abs(Z_rot[,i])>minZ,i]),
       type = "h",
       lwd = 2, xlab = "", ylab = "", xaxt = "n", ylim = ylm,
       xlim = c(0.5, N_ts + 0.5), col = clr)
  for(j in 1:N_ts) {
    if(Z_rot[j,i] > minZ) {text(j, -0.03, ylbl[j], srt = 90, adj = 1, cex = 1.2, col = clr[j])}
    if(Z_rot[j,i] < -minZ) {text(j, 0.03, ylbl[j], srt = 90, adj = 0, cex = 1.2, col = clr[j])}
    abline(h = 0, lwd = 1.5, col = "gray")
  } 
  mtext(paste("Factor loadings on state", i), side = 3, line = 0.5)
}

## plot CCF's
ccf(proc_rot[1,],proc_rot[2,], lag.max = 15, main="")
title(main = "State 1 vs 2")
```
```{r Continued Diagnostics}

autoplot(best_fit) #pick the residual one, and the qq plots 

```
It looks like Epischura is following the negative of state two whereas most of the other species are following either just state 1 or a mixture of state one and state two.

The diagnostic plots show relativly good fits, though other algae and diatoms aren't capturing the full range of data. There may be some structure to the residuals for diatoms and maybe cyclops with few outliers, but all species and states show normality in the QQ plots. 

THIS COULD BE FLESHED OUT A LITTLE MORE


### Initial Model Fits 
```{r}
get_DFA_fits <- function(MLEobj, dd = NULL, alpha = 0.05) {
    ## empty list for results
    fits <- list()
    ## extra stuff for var() calcs
    Ey <- MARSS:::MARSShatyt(MLEobj)
    ## model params
    ZZ <- coef(MLEobj, type = "matrix")$Z
    ## number of obs ts
    nn <- dim(Ey$ytT)[1]
    ## number of time steps
    TT <- dim(Ey$ytT)[2]
    ## get the inverse of the rotation matrix
    H_inv <- varimax(ZZ)$rotmat
    ## check for covars
    if (!is.null(dd)) {
        DD <- coef(MLEobj, type = "matrix")$D
        ## model expectation
        fits$ex <- ZZ %*% H_inv %*% MLEobj$states + DD %*% dd
    } else {
        ## model expectation
        fits$ex <- ZZ %*% H_inv %*% MLEobj$states
    }
    ## Var in model fits
    VtT <- MARSSkfss(MLEobj)$VtT
    VV <- NULL
    for (tt in 1:TT) {
        RZVZ <- coef(MLEobj, type = "matrix")$R - ZZ %*% VtT[, 
            , tt] %*% t(ZZ)
        SS <- Ey$yxtT[, , tt] - Ey$ytT[, tt, drop = FALSE] %*% 
            t(MLEobj$states[, tt, drop = FALSE])
        VV <- cbind(VV, diag(RZVZ + SS %*% t(ZZ) + ZZ %*% t(SS)))
    }
    SE <- sqrt(VV)
    ## upper & lower (1-alpha)% CI
    fits$up <- qnorm(1 - alpha/2) * SE + fits$ex
    fits$lo <- qnorm(alpha/2) * SE + fits$ex
    return(fits)
}


clr <- c("brown", "blue", "darkgreen", "darkred", "purple")
N_ts <- dim(plank_dat)[1]
w_ts <- seq(dim(plank_dat)[2])
## get model fits & CI's
mod_fit <- get_DFA_fits(best_fit)
## plot the fits
ylbl <- plank_taxa
par(mfrow = c(N_ts, 1), mai = c(0.5, 0.7, 0.1, 0.1), omi = c(0, 
    0, 0, 0))
for (i in 1:N_ts) {
    up <- mod_fit$up[i, ]
    mn <- mod_fit$ex[i, ]
    lo <- mod_fit$lo[i, ]
    plot(w_ts, mn, xlab = "", ylab = ylbl[i], xaxt = "n", type = "n", 
        cex.lab = 1.2, ylim = c(min(lo), max(up)))
    axis(1, 12 * (0:dim(plank_dat)[2]) + 1, yr_frst + 0:dim(plank_dat)[2])
    points(w_ts, plank_dat[i, ], pch = 16, col = clr[i])
    lines(w_ts, up, col = "darkgray")
    lines(w_ts, mn, col = "black", lwd = 2)
    lines(w_ts, lo, col = "darkgray")
}
```
These fits seem to be missing a lot of the interannual 

# Seasonal Model Exploration

**Describe the effects of environmental or dummy variables on (possibly seasonal) patterns in the data.**

## Seasonal Model Set-up
We decided to look at three seasonal patterns, an annual pattern (12 months), a six month pattern, and a 3 month pattern (Winter, Spring, Summer, Fall).

```{r Seasonal exploration, cache = TRUE}

TT <- dim(plank_dat)[2]

fits_season <- list()

out.tab.season <- NULL

model.states <- 1
names(model.states) <- c("two_states")

R.models <- c("diagonal and unequal")

# 12 month period 
  cos.12 <- cos(2 * pi * seq(TT)/12)
  sin.12 <- sin(2 * pi * seq(TT)/12)
  c.Four_12 <- rbind(cos.12, sin.12)
  
# 6 month period 
  cos.6 <- cos(2 * pi * seq(TT)/6)
  sin.6 <- sin(2 * pi * seq(TT)/6)
  c.Four_6 <- rbind(cos.6, sin.6)
  
# 3 month period 
  cos.3 <- cos(2 * pi * seq(TT)/3)
  sin.3 <- sin(2 * pi * seq(TT)/3)
  c.Four_3 <- rbind(cos.3, sin.3)

  d10 = zscore(t(init_dat[,c("Temp","TP", "pH")]))
  d12 = rbind(c.Four_12, d10)
  d6 = rbind(c.Four_6, d10)
  d3 = rbind(c.Four_3, d10)

  d.models <- list(d10, d12, d6, d3)
  names(d.models) <- c("Temp,TP, and pH", "12 month", "6 month", "3 month")

  mod.list = list(
    A = "zero",
    Q = "identity",
    x0 = "zero"
  )

  start.time <- Sys.time()
    for (j in 1:length(d.models)){
      for(R.model in R.models){
        fit.model = c(list(m=i, R=R.model), mod.list)
        fit <- MARSS(plank_dat, model=fit.model, 
                     form = "dfa", z.score = FALSE, silent = TRUE,
                     control=list(maxit=800, allow.degen=TRUE), 
                     covariates = d.models[[j]])
        out=data.frame(States=names(model.states)[i-1],d=names(d.models)[j],R=R.model, # 
                       logLik=fit$logLik, AICc=fit$AICc, num.param=fit$num.params,
                       num.iter=fit$numIter, converged=!fit$convergence,
                       stringsAsFactors = FALSE)
        out.tab.season=rbind(out.tab.season,out)
        fits_season=c(fits,list(fit))
      }
    }
 
  end.time <- Sys.time()

  elapsed.time <- round((end.time - start.time), 3)
  print(paste("Elapsed time for period", p, ":", elapsed.time))


min.AICc <- order(out.tab.season$AICc)
out.tab.2 <- out.tab.season[min.AICc, ]
out.tab.2

```

# Seasonal Results 
The best 
## Matrix for Best Seasonal Model 

$$
\begin{bmatrix}
y_1\\
y_2\\
y_3\\
y_4\\
y_5\\
\end{bmatrix}_t=
\begin{bmatrix}
z_{1,1} & z_{1,2}\\
z_{2,1} & z_{2,2}\\
z_{3,1} & z_{3,2}\\
z_{4,1} & z_{4,2}\\
z_{5,1} & z_{5,2}\\
\end{bmatrix}*
\begin{bmatrix}
x_1\\
x_2\\
\end{bmatrix}_t+
\begin{bmatrix}
d_{1,1} & d_{1,2} & d_{1,3} & d_{1,4} & d_{1,5}\\
d_{2,1} & d_{2,2} & d_{2,3} & d_{2,4} & d_{2,5}\\
d_{3,1} & d_{3,2} & d_{3,3} & d_{3,4} & d_{3,5}\\
d_{4,1} & d_{4,2} & d_{4,3} & d_{4,4} & d_{4,5}\\
d_{5,1} & d_{5,2} & d_{5,3} & d_{5,4} & d_{5,5}\\
\end{bmatrix}*
\begin{bmatrix}
Temp \\
TP \\
pH\\
cos(2pi*TT/6)\\
sin(2pi*TT/6\\
\end{bmatrix}_t +
\begin{bmatrix}
v_1\\
v_2\\
v_3\\
v_4\\
v_5
\end{bmatrix}_t

$$
Process Error
$$

\text{Where }
\begin{bmatrix}
w_1\\
w_2\\
\end{bmatrix}
 \sim MVN
\begin{pmatrix}
\begin{bmatrix}
0\\
0\\
\end{bmatrix},
\begin{bmatrix}
1 & 0\\
0 & 1\\
\end{bmatrix}
\end{pmatrix}
$$
Observation Error 
$$
\text{Where }
\begin{bmatrix}
v_1\\
v_2\\
v_3\\
v_4\\
v_5\\
\end{bmatrix}
 \sim MVN
\begin{pmatrix}
\begin{bmatrix}
0\\
0\\
0\\
0\\
0\\
\end{bmatrix},
\begin{bmatrix}
r_1 & 0& 0& 0& 0\\
0 & r_2& 0& 0& 0\\
0 & 0& r_3& 0& 0\\
0 & 0& 0& r_4& 0\\
0 & 0& 0& 0& r_5\\
\end{bmatrix}
\end{pmatrix}
$$
## Plotting the Seasonal Model
```{r factor_loadings including seasonality}
# Rotating the model, plotting the states and loadings
best_fit_s <- fits_season[[3]]

# Get Z Estimations and rotate the data
## get the estimated ZZ
Z_est_s <- coef(best_fit_s, type = "matrix")$Z

## get the inverse of the rotation matrix
H_inv_s <- varimax(Z_est_s)$rotmat

## rotate factor loadings
Z_rot_s = Z_est_s %*% H_inv_s   

## rotate processes
proc_rot_s = solve(H_inv_s) %*% best_fit_S$states

N_ts <- nrow(plank_dat)
m = ncol(Z_est)
## plot labels
ylbl <- plank_taxa
w_ts <- seq(mm)

## set up plot area
layout(matrix(1:4, m, 2), widths = c(3,2))
par(mai = c(0.3, 0.3, 0.3, 0.1), omi = c(0, 0, 0, 0))

## plot the processes
for(i in 1:m) {
  ylm <- c(-1, 1) * max(abs(proc_rot_s[i,]))
  ## set up plot area
  plot(w_ts,proc_rot_s[i,], type = "n", bty = "L",
       ylim = ylm, xlab = "", ylab = "", xaxt = "n")
  ## draw zero-line
  abline(h = 0, col = "gray")
  ## plot trend line
  lines(w_ts, proc_rot_s[i,], lwd = 2)
  lines(w_ts, proc_rot_s[i,], lwd = 2)
  ## add panel labels
  mtext(paste("State",i), side = 3, line = 0.5)
  axis(1, 12 * (0:mm) + 1, yr_frst + 0:mm)
}
## plot the loadings
minZ <- 0
ylm <- c(-1, 1) * max(abs(Z_rot_s))
for(i in 1:m) {
  plot(x = c(1:N_ts)[abs(Z_rot_s[,i])>minZ],
       y = as.vector(Z_rot_s[abs(Z_rot_s[,i])>minZ,i]),
       type = "h",
       lwd = 2, xlab = "", ylab = "", xaxt = "n", ylim = ylm,
       xlim = c(0.5, N_ts + 0.5), col = clr)
  for(j in 1:N_ts) {
    if(Z_rot_s[j,i] > minZ) {text(j, -0.03, ylbl[j], srt = 90, adj = 1, cex = 1.2, col = clr[j])}
    if(Z_rot_s[j,i] < -minZ) {text(j, 0.03, ylbl[j], srt = 90, adj = 0, cex = 1.2, col = clr[j])}
    abline(h = 0, lwd = 1.5, col = "gray")
  } 
  mtext(paste("Factor loadings on state with seasonality", i), side = 3, line = 0.5)
}

## plot CCF's
ccf(proc_rot_s[1,],proc_rot_s[2,], lag.max = 15, main="")
title(main = "State 1 vs 2 with seasonality")
```
The factor loading seen in the model that is including seasonality seems to explain the dynamics of the speices better than the model that did not include seasonality, however, the sign of factor loading for Epishura, our predator species, changed to become positive in the seasonal model. 

Factor loadings: 
* Pattern for primary producers and grazers is better explained by state one 
* Pattern for predator species is better explained by state two

CCF plot indicates that there is still some seasonal mechanism occurring that our model is not capturing. 

## Additional Diagnostic 
```{r Continued Diagnostics, cache = TRUE}

autoplot(best_fit_s)  

```
This model is fitting the species better overall compared to the non-seaosnal model. Residuals and assumptions of normality look good overall. 


### Model Fits 
```{r}

clr <- c("brown", "blue", "darkgreen", "darkred", "purple")
N_ts <- dim(plank_dat)[1]
w_ts <- seq(dim(plank_dat)[2])
## get model fits & CI's
mod_fit_s <- get_DFA_fits(best_fit_s)
## plot the fits
ylbl <- plank_taxa
par(mfrow = c(N_ts, 1), mai = c(0.5, 0.7, 0.1, 0.1), omi = c(0, 
    0, 0, 0))
for (i in 1:N_ts) {
    up <- mod_fit_s$up[i, ]
    mn <- mod_fit_s$ex[i, ]
    lo <- mod_fit_s$lo[i, ]
    plot(w_ts, mn, xlab = "", ylab = ylbl[i], xaxt = "n", type = "n", 
        cex.lab = 1.2, ylim = c(min(lo), max(up)))
    axis(1, 12 * (0:dim(plank_dat)[2]) + 1, yr_frst + 0:dim(plank_dat)[2])
    points(w_ts, plank_dat[i, ], pch = 16, col = clr[i])
    lines(w_ts, up, col = "darkgray")
    lines(w_ts, mn, col = "black", lwd = 2)
    lines(w_ts, lo, col = "darkgray")
}
```
Still not fitting other algae and daphina particularly well, but following the trends and capturing most of the data within the confidence intervals.  

# Discussion

In this exercise our goal was to capturing the dynamics of predator and prey species in Lake Washington in the late 70's and early 80's. 

Seasonal model (six month period) fit better and the factor loading results showed more of a deviance between state one and state two for primary producers and grazers, and predators.

All three environmental covariates were included in the best fit model, showing that they are explaning levels of variability for the selected species in the selected time period. 

Epishura seem to have an offset from the other species. We're fitting two independent states to five species, and while one state captured the dynamics of grazers and primary producers relatively well, epishura was not captured at all in this state, indicating independence between these two groups of plankton in the time period analyzed. 

All three environmental covariates were included in the best fit model, showing that they are explaning levels of variability for the selected species in the selected time period. 


# Team contributions

Madi set up plotting code for the entire time series for all species to select species of interest in data rich periods. 

Maria and Eric wrote the majority of the model set up code and created functions used for plotting. 

Madi took the Maria's code and edited to include seasonal considerations, added commentary for diagnostics, and helped to set up the Rmarkdown structure. 

All team members helped with interpretation of results and writing of the report. 

