---
title: "Lab 3 - Common trends in plankton data"
subtitle: "Dynamic Factor Analysis (DFA)"
author: "Person 1, Person 2, Person 3"
date: April 20, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

------------------------------------------------------------------------

# Data

Describe what plankton groups and covariates you used and indicate any temporal subsetting you chose.

## Load the data

```{r load_data}
## load MARSS for data and analyses
library(MARSS)

## load the raw data (there are 3 datasets contained here)
data(lakeWAplankton, package = "MARSS")

## we want `lakeWAplanktonTrans`, which has been transformed
## so the 0's are replaced with NA's and the data z-scored
all_dat <- lakeWAplanktonTrans
```

## Wrangle the data

-   You can choose any taxa to include in your analyses, but make sure there are at least 5 groups. You can mix phytoplankton and zooplankton, if you'd like, but justify any choices you make.

-   You will want to subset the data to a time period with adequate observations for the taxa of interest. Your time window should include at lest 5 years (60 months).

```{r wrangle_data}
## add some code here
```

## General Notes

labmate emails:

Madi Heller-Shipley --[shipmadison\@gmail.com](mailto:shipmadison@gmail.com){.email} Eric French -- [ericvfrench\@gmail.com](mailto:ericvfrench@gmail.com){.email} Maria Mariakuruvilla -- [mariakur\@uw.edu](mailto:mariakur@uw.edu){.email}

*Diatoms* dominated in early spring time (Phytoplankton) *Daphnia* eat diatoms --\> bloom after diatoms (zooplankton)

*bluegreen* (Cyano-bacteria) is toxic --\> dominace correlated to TP (not as nitrogen limited)

*Epischura* --\> calanus copepod peaks after diatoms

*Neomysis* --\> opossum shrimp (predator) Data limited

# First Steps

1)  Plot all species all years and look for data-based relationships -connect what we see in the data to literature and the biology of Lake Washington

2)  Pick our species and connect them!!!

3)  Connect species to avaialable environmental data

4)  Play with models, move to general tasks

# General tasks

Each group has the same general tasks, but you will adapt them as you work on the data.

1)  Find the most parsimonious model among a set that examines the effects of environmental covariates and/or an indicator of seasonality, varying numbers of trends, and different forms of variance-covariance matrices for the observation errors.

2)  Plot trends and individual loadings for the model identified in task (1) above.

3)  Plot model fits and uncertainty intervals for the model identified in task (1) above.

4)  Describe the effects of environmental or dummy variables on (possibly seasonal) patterns in the data.

# Model fitting from lab 2

MARSS assumptions
```{r}

# Here we evaluated the data support for the following hypotheses about Lower Columbia salmon river trends
# Each Z model is a hypothesis

## Simple run timing groupings
# summer = kalama_river_summer
# winter = remaining population

## Simple spatial groupings
# coweeman = coweeman and sf toutle
# ef_lewis = east fork lewis
# kalama = kalama summer and winter
# cowlitz = lower cowlitz, tilton, and upper cowlitz
# washougal = washougal

## Correlation spatial groupings
# coweeman = coweema, ef_lewis, sf toutle, washougal
# kalama = kalama summer and winter
# cowlitz = lower cowlitz, tilton, and upper cowlitz

Z.models <- list(
  H1 = matrix(1,9,1), #one hidden population state
  H2 = factor(c("coweeman","ef_lewis", "kalama_sum",
                'kalama_win', "low_cowlitz", "sf_toutle",
                "tilton", "up_cowlitz", "washougal")), #states are defined by individual population
  H3 = factor(c("winter","winter", "summer",
                'winter', "winter", "winter",
                "winter", "winter", "winter")),# states defined by running timing grouping (n = 2)
  H4 = factor(c("coweeman","ef_lewis", "kalama",
                'kalama', "cowlitz", "coweeman",
                "cowlitz", "cowlitz", "washougal")),# states defined by spatial grouping (n = 5)
  H5 = factor(c("coweeman","coweeman", "kalama",
                'kalama', "cowlitz", "coweeman",
                "cowlitz", "cowlitz", "coweeman"))# states defined by correlation (n = 3)
)

names(Z.models) <- c("one_population","indiv_population","run_timing_groups",'spatial_groups',"corr_groups")

# Also testing different process error varcovar matrices
Q.models2 <- c("equalvarcov","unconstrained")

# Process cycling terms
# Each is on a 10-year cycle, will test orders 0-4
# see Evaluating Cycling Section

ct.models <- list(
  c0 = "zero", # No cycling
  
  c1 = t(fourier(ts(1:ncol(dat), frequency = 10), K=1)), # seasonality follows a 1st order fourier series with the same # rows as y
 
  c2 = t(fourier(ts(1:ncol(dat), frequency = 10), K=2)), # seasonality follows a 2nd order fourier series
  
  c3 = t(fourier(ts(1:ncol(dat), frequency = 10), K=3)), # seasonality follows a 3rd order fourier series

  c4 = t(fourier(ts(1:ncol(dat), frequency = 10), K=3)) # seasonality follows a 3rd order fourier series
  )
names(ct.models) <- c("no_cycling", "1st_order", "2nd_order", "3rd_order", "4th_order")

# Bias terms
u2 <- c("unequal", 'equal')

# Setting fixed portion of mod list
mod.list = list(
  A = "scaling",
  R = "diagonal and equal")

```

## MARSS Model Selection

Run MARSS models

```{r}
# Added in another for loop to run through cycling options, be warned, this means there are 80 total models. It takes a long time to run
out.tab <- NULL
fits <- list()
for(i in 1:length(Z.models)){
  for (j in 1:length(ct.models)){
    for(Q.model in Q.models2){
      for(U.model in u2){
      fit.model = c(list(Z=Z.models[[i]], c=ct.models[[j]], Q=Q.model, U=U.model), mod.list)
      fit = MARSS(dat, model=fit.model,
                silent=TRUE, control=list(maxit=5000))
      out=data.frame(H=names(Z.models)[i],ct=names(ct.models)[j],Q=Q.model,U=U.model,
                   logLik=fit$logLik, AICc=fit$AICc, num.param=fit$num.params,
                   m=length(unique(Z.models[[i]])),
                   num.iter=fit$numIter, converged=!fit$convergence,
                   stringsAsFactors = FALSE)
      out.tab=rbind(out.tab,out)
      fits=c(fits,list(fit))
      } 
    }
  }
}

min.AICc <- order(out.tab$AICc)
out.tab.1 <- out.tab[min.AICc, ]
out.tab.1 <- cbind(out.tab.1, delta.AICc = out.tab.1$AICc - out.tab.1$AICc[1])
out.tab.1 <- cbind(out.tab.1, rel.like = exp(-1 * out.tab.1$delta.AICc/2))
out.tab.1 <- cbind(out.tab.1, AIC.weight = out.tab.1$rel.like/sum(out.tab.1$rel.like))
out.tab.1


```

# Methods

Please address the following in your methods:

-   Which plankton taxa did you choose and how did you choose them?

-   What time period(s) did you examine and why?

-   What environmental or dummy variables did you include and why?

-   What forms of models did you fit (ie, write them out in matrix form)?

-   What sort of model diagnostics did you use to examine model assumptions?

# Results

# Discussion

# Team contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."
