---
title: Lab 1 Forecasting with ARIMA models
author: Veggerby, Wang
date: April 10, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r include=FALSE}
library(tidyverse)
library(here)
library(tseries)
library(knitr)
options(dplyr.summarise.inform = FALSE)
```

# Lab 1: ARIMA models

Team member names: Terrance Wang (SAFS), Karl Veggerby (SAFS)


# Data

Ruggerone & Irvine Data. We will focus on sockeye salmon and all their 15 regional stocks in the North Pacific. 

# Question your team will address

What is the autoregressive structure for sockeye salmon? Is there a general autoregressive structure to describe all sockeye salmon regional stocks? Are there similarities and differences within and among North American West Coast, Alaska, and East Asian sockeye stocks? 


# Method you will use

ARIMA for individual regional stock time series trends. Compare ARIMA structure and parameters. 

### Initial plan

We will fit ARIMA models for each of the sockeye regional stocks with the Box Jenkins method, which includes selecting the model form, estimating the parameters, and diagnosing the model. We will then compare the parameters of the individual ARIMA model structure among regional stocks and larger regional aggregations visually. 

### What you actually did

Example, "We were able to do our plan fairly quickly, so after writing the basic code, we divided up all 12 regions in Ruggerone & Irvine and each team member did 4 regions. We compared the accuracy of forecasts for different time periods using 20-years for training and 5-year forecasts each time. We compared the RMSE, MAE, and MAPE accuracy measures."

# Diagnostics and preliminary exploration
### Massage the data
```{r}
# read in data
rug_path = file.path(here(), 'Lab-1','Data_Images','ruggerone_data.rds')
ruggerone_data = readRDS(rug_path)

# get sockeye data per region
sockeye_data=ruggerone_data %>% 
  mutate(lnreturns=log(returns)) %>%
  filter(species=="sockeye" & !region %in% c("korea","japan")) # do not include NA or all 0 data

# summarise data by area
subdata <- ruggerone_data %>% 
  mutate(
    area = case_match(
      region, 
      c("japan", "korea", "m_i", "e_kam", "w_kam") ~ "East_Asia",
      c("wak", "s_pen", "kod", "ci", "pws", "seak") ~ "Alaska",
      c("nbc", "sbc", "wa", "wc") ~ "WC",
      .default = region
    )) %>%
  group_by(area, species, year) %>%
  summarize(lntotal = log(sum(returns, na.rm=TRUE)))

    ```
### Plot the data
```{r}
sockeye_data %>% 
  ggplot(aes(x=year, y=lnreturns)) + 
    geom_line() + 
    ggtitle("log abundance by region")+
    facet_wrap(~region,scales = "free") + theme_bw()

subdata %>% filter(species=="sockeye")%>%
  ggplot(aes(x=year, y=lntotal)) + 
    geom_line() + 
    ggtitle("log abundance by area")+
    facet_wrap(~area)+ theme_bw()

```

At the regional level, systematic changes in the mean are present in many of the stocks (e.g., nbc, w_kam, kod). There are a few increasing means and some random-walk-like ones. There is a strong linear trend in the ci, kod, and s_pen regions. Changes in the variance may also be present in a few stocks (e.g., wak, kod). 4-year period cycles are possible in several regions (e.g., wak, sbc). 

At the area-wide level, Alaska, East Asia, and NA West Coast all exhibit systematic changes. There are some similarities among areas. Alaska and NA West Coast share periodic variations. Alaska and East Asia both have increasing regional stocks. There also differences among areas, such as the decreasing trend and increasing variance in the West Coast regional stocks.

From these plots, we interpret that there is likely non-stationarity at both the regional and area-wide scale, and proceed with our analysis.  

### Use ACF and PACF
```{r}
library(forecast)
# run ACF and PACF for each region, massage data into longer format for ggplotting
sockeye_acf_pacf = sockeye_data %>% 
  group_by(region) %>% 
  summarise(list_acf=list(acf(lnreturns, plot=FALSE)),
            list_pacf=list(pacf(lnreturns, plot=FALSE))) %>%
  mutate(acf_lnreturns=purrr::map(list_acf, ~as.numeric(.x$acf)),
         # add 1 to first value of PCF, one of the weird things Mark mentioned
         pacf_lnreturns=purrr::map(list_pacf, ~as.numeric(c(0,.x$acf)))) %>% 
  select(-c(list_acf,list_pacf)) %>% 
  unnest(c(acf_lnreturns,pacf_lnreturns)) %>% 
  group_by(region) %>% 
  mutate(lag=row_number() - 1) %>%
  pivot_longer(cols=c(acf_lnreturns,pacf_lnreturns),
               values_to = "cf",
               names_to = "cf_type")

sockeye_ci = sockeye_data %>% 
  group_by(region) %>% 
  summarise(ci = qnorm((1 + 0.95)/2)/sqrt(n()))

ggplot(sockeye_acf_pacf,aes(x=lag, y=cf, fill=cf_type)) +
  geom_bar(stat="identity", position = 'dodge',width=1) +
  geom_hline(yintercept = 0) +
  geom_hline(data = sockeye_ci, aes(yintercept = -ci), color="blue", linetype="dotted") +
  geom_hline(data = sockeye_ci, aes(yintercept = ci), color="blue", linetype="dotted") +
  labs(x="Lag", y="ACF") +
  facet_wrap(~region) + theme_bw()
```
A handful of regions (e.g., ci, seak) show ACF with a long tail and PACFs that cut off at 1-3 orders, which may indicate some autoregressive structure. Other regions (e.g., pws, m_i) have a sharp cut off with ACF and tailing PACFs, which may suggest a moving average structure. Some regions (s_pen,w_kam,kod) show long tails for both ACF and PACF, suggesting ARMA structure. A few regions show periodic trends (negative AR parameters) from their oscillating correlations factors. 

### Test for stationarity

We ran stationarity tests at the area-wide level for simplicity. 

```{r, warning=FALSE, message=FALSE}
# stationarity tests
subdata_sockeye<-subdata %>% filter(species=="sockeye")

# can't get select to work here
alaska_sockeye<-subdata_sockeye %>% 
  filter(area=="Alaska") %>%
  select(lntotal)

# dickey fuller test for stationarity alaska
alaska_sockeye_vec<-alaska_sockeye$lntotal

ak_adf = adf.test(alaska_sockeye_vec)

# do first difference to see if we can get to stationarity
alaska_sockeye_vec_1diff<-diff(alaska_sockeye_vec)

ak_adf_d1 = adf.test(alaska_sockeye_vec_1diff)
# doing a first difference successful for Alaska


# dickey fuller test for stationarity east asia
asia_sockeye<-subdata_sockeye %>% 
  filter(area=="East_Asia") %>%
  select(lntotal)

asia_sockeye_vec<-asia_sockeye$lntotal

ea_adf = adf.test(asia_sockeye_vec)

asia_sockeye_vec_1diff<-diff(asia_sockeye_vec)
ea_adf_d1 = adf.test(asia_sockeye_vec_1diff)

# dickey fuller test for stationarity west coast
wc_sockeye<-subdata_sockeye %>% 
  filter(area=="WC") %>%
  select(lntotal)

wc_sockeye_vec<-wc_sockeye$lntotal

wc_adf = adf.test(wc_sockeye_vec)

wc_sockeye_vec_1diff<-diff(wc_sockeye_vec)
wc_adf_d1 = adf.test(wc_sockeye_vec_1diff)

# confirm with forcast 
ak_nd = forecast::ndiffs(alaska_sockeye_vec)
ea_nd = forecast::ndiffs(asia_sockeye_vec)
wc_nd = forecast::ndiffs(wc_sockeye_vec)

adf_tbl = data.frame(Region=c("AK","E_Asia","US_WC"), 
           DF_Test_diff0_p = c(ak_adf$p.value,ea_adf$p.value,wc_adf$p.value), 
           DF_Test_diff1_p = c(ak_adf_d1$p.value,ea_adf_d1$p.value,wc_adf_d1$p.value), 
           N_pred_diffs = c(ak_nd,ea_nd,wc_nd))

kable(adf_tbl)

```

The Dickey-Fuller tests for each of the 3 areas do not reject the null hypothesis that the data is stationary. Differencing all 3 areas by 1 order makes the sockeye abundance stationary, suggesting random walk characteristics. The unit root tests of estimated order of differencing confirms the D-F tests for Alaska and East Asia, but not NA West Coast. The NA West Coast does have a non-moving mean, which partly explains the difference between D-F test and unit root test. 

Altogether, the abundance plots, ACF and PCF plots, and stationarity tests demonstrate that the abundances of the regional stocks can be explained by some combination of auto-regressive, differencing, and moving average components. For this reason, we proceed forward with fitting ARIMA models to each of the regional stocks. 

# Results

We fit ARIMA models to each individual regional stock and report the fits and diagnostics in the table below. 
```{r,results='hide',fig.show='hide'}
# make a data frame to fill in values for each stock
stock_arima = data.frame(matrix(ncol = 4))
names = c("region","p","d","q")
colnames(stock_arima) = names
sock_stocks = unique(sockeye_data$region)

# loop through each stock and auto arima and save outputs
for (i in 1:length(sock_stocks)){
# subset data to each region
stock=sock_stocks[i]
stock_data = sockeye_data%>% filter(region==stock)
datts = ts(stock_data$lnreturns, start=stock_data$year[1])

# arima tests 
mod = auto.arima(datts)
mod$coef
arima_order = arimaorder(mod)

# look at model resids
resid = checkresiduals(mod,plot=F)

# save all this data
stock_arima[i,c("p","d","q")] = arima_order
stock_arima[i,c(names(mod$coef))] = mod$coef
stock_arima[i,c("region")] = stock
stock_arima[i,c("resid_p")] = resid$p.value

}

stock_arima_area = stock_arima %>% 
  # fill in NAs with 0s
  mutate_if(is.numeric, ~replace_na(., 0)) %>% 
  # add area
  mutate(
    area = case_match(
      region, 
      c("japan", "korea", "m_i", "e_kam", "w_kam") ~ "East_Asia",
      c("wak", "s_pen", "kod", "ci", "pws", "seak") ~ "Alaska",
      c("nbc", "sbc", "wa", "wc") ~ "WC",
      .default = region)) 
```
```{r}
kable(stock_arima_area%>% select(area,region, p, d, q,ar1,ar2,ma1,ma2,resid_p)%>%arrange(area),
      caption = "ARIMA fitting results and residual diagnostics for all sockeye regions")
```

The best fitting ARIMA models show a wide diversity of structure. Similarities and differences within and among areas will be discussed in more detail in the next section. Residuals are generally stationary, meaning acceptable ARIMA fits, with a few exceptions (e.g., kod, sbc). Below, the example of w_kam shows that residuals show signs of stationarity like a stable mean, normal distribution of values, and little autocorrelation. Though there is some temporal pattern in the variance. 

```{r,message = FALSE,fig.cap = "Example of residual diagnostics of one of the regions"}
resid = checkresiduals(mod,plot=T)
```

```{r,fig.cap = c("ARIMA structure by area","AR parameters by area","MA parameters by area")}
# make into long format for ggplotting
stock_gg_long = stock_arima_area %>% 
  pivot_longer(-c(region,area), names_to = "parameter", values_to = "param_est")

# General ARIMA structure
stock_gg_long %>% filter(parameter %in% c("p","d","q")) %>%
  ggplot(aes(x=param_est,fill=area)) + geom_bar() +
  facet_wrap(vars(parameter),ncol=1) + theme_bw()

# Autoregressive plot
stock_gg_long %>% filter(parameter %in% c("ar1","ar2")) %>%
  ggplot(aes(x=param_est,y=area,color=area)) + geom_boxplot(outlier.shape = NA) +
  geom_jitter()+
  facet_wrap(vars(parameter),ncol=1, scales = "free")+ theme_bw()

# Moving average plot
stock_gg_long %>% filter(parameter %in% c("ma1","ma2")) %>%
  ggplot(aes(x=param_est,y=area,color=area)) + geom_boxplot(outlier.shape = NA) +
  geom_jitter()+
  facet_wrap(vars(parameter),ncol=1, scales = "free")+ theme_bw()

```

The ARIMA structure for all sockeye stocks (ignoring area) do not show a clear pattern in AR, differencing, and MA. A slight majority of stocks support first order AR, differencing, and MA. When we look closer at the parameter values, there is a small tendency for sockeye to have positive AR1 values and negative MA1 values. This suggests sockeye populations are loosely related to the previous time step and deviate around the general trajectory with a 2 year period. 

Alaska shows a consistent ARIMA pattern for all of its regions. The best fit ARIMA models for AK regions all predict first difference. AK strongly supports a first order of the moving average part. Additionally, all AK regions have negative MA1 values. There is some substructure of ARIMA properties within AK. AK shows no area-wide consensus on the order of the autoregressive part, but  Looking more closely at the autoregressive parameter values of Alaska regions, we see only positive AR1 values for the regions with at least 1 order of AR (wak,ci,pws). Wak, ci, and pws populations tend to be more related to the previous year than other AK regions are. 

West Coast regions do not have a consensus on ARIMA structure, with an exception of 0 differencing. AR1 and MA1 parameters range from negative to positive. This inconclusiveness is partly explained by the low sample size (n=3). East Asia regions are also inconsistent on the predicted order number of the moving average and autoregressive components. With only 3 East Asia regions with data, it is difficult to construe any pattern.

# Discussion
Overall, the sockeye stocks population trajectories are clearly not stationary. All regions benefit from ARIMA models to understand their underlying temporal dynamics. There is no unifying ARIMA structure that can reliably describe all sockeye populations. This is unsurprising because of the wide spectrum of biological and environmental forces that impact sockeye across its species range. Scattered across the North Pacific are 3 sockeye ecotypes (kokanee, lake, sea/river) that often occur together in the same watershed. 


# Description of each team member's contributions

Terrance did most of it, Karl did stationarity section.
