---
title: Lab 2 MARSS Models
author: Liz Elmstrom (SAFS), Terrance Wang, Eric French
date: April 20, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

# General Questions

Each group has the same general tasks, but you will adapt them as you work on the data.

1. Create estimates of spawner abundance for all missing years and provide estimates of the decline from the historical abundance.

2. Evaluate support for the major population groups. Are the populations in the groups more correlated than outside the groups?

3. Evaluate the evidence of cycling in the data. *We will talk about how to do this on the Tuesday after lab.*

Step 1: Estimating data that doesn't exist-- Terrance
Step 2A: Testing temporal Z matrices (different run times vs. one pop)-- Terrance
Step 2B: Testing spatial Z matrices (different spatial groupings vs. one pop)-- Liz
Step 3: Adding in seasonality as a covariate (C/c matrix) -- Eric

Describe what data set you will work with and any subsetting you decided on. For example, you may have decided to look only at a section of the ESU that your team was assigned.

# Methods

Address the following in your methods

* Describe your assumptions about the x and how the data time series are related to x.

   - How are the x and y (data) related? 1 x for 1 y or will you assume 1 x for all y or 1 x for each major population group? How will you choose? 
   
We tested 4 different population groupings: one population, individual populations, spatial groupings, and groupings based on correlations
   
   - What will you assume about the U for the x's?
   
We tested the support for both equal and unequal population growth rates. 
   
   - What will you assume about the Q matrix?
   
We assumed that these populations were correlated through time and tested model support for unconstrained and equal variance covariance matrices. 
   
* Write out your assumptions as different models **in matrix form**, fit each and then compare these with AIC or AICc.
   
Our models are written in matrix form in the "Model Assumptions" section of the results. 

* Do your estimates differ depending on the assumptions you make about the structure of the data, i.e. you assumptions about the x's, Q, and U.

# Data exploration

Load the data.
```{r}
library(tidyverse)
library(dplyr)
library(janitor)
load(here::here("Lab-2", "Data_Images", "columbia-river.rda"))

#Evolutionary Significant Units
esu <- unique(columbia.river$esu_dps)
esu
```
```{r}

df <- columbia.river %>% subset(esu_dps %in% "Steelhead (Lower Columbia River DPS)")

cat("colnames: ", colnames(df), "\n")

unique(df$esapopname)
unique(df$commonpopname)

df$esapopname2 <- stringr::str_replace(df$esapopname, "Steelhead [(]Lower Columbia River DPS[)] ", "")
unique(df$esapopname2)

ggplot(df, aes(x=spawningyear, y=log(value), color=majorpopgroup)) + 
  geom_point(size=2, na.rm = TRUE) + 
  theme(strip.text.x = element_text(size = 3)) +
  theme(axis.text.x = element_text(size = 5, angle = 90)) +
  facet_wrap(~esapopname2) +
  ggtitle('Lower Columbia Steelhead Populations')+
  theme(strip.text.x = element_text(size = 10))

```

Wrangle the data.
``` {r}

esuname <- esu[3]
dat <- columbia.river %>% 
  subset(esu_dps == esuname) %>% # get only this ESU
  mutate(log.spawner = log(value)) %>% # create a column called log.spawner
  select(esapopname, spawningyear, log.spawner) %>% # get just the columns that I need
  pivot_wider(names_from = "esapopname", values_from = "log.spawner") %>% 
  column_to_rownames(var = "spawningyear") %>% # make the years rownames
  as.matrix() %>% # turn into a matrix with year down the rows
  t() # make time across the columns
dat[is.na(dat)] <- NA

## Fixing row names
tmp <- rownames(dat)
tmp <- stringr::str_replace(tmp, "Steelhead [(]Lower Columbia River DPS[)]", "")
tmp <- make_clean_names(tmp)
rownames(dat) <- tmp

dat <- dat[1:9,]# remove upper gorge
dat <- dat[ order(row.names(dat)), ]## sort

```

# Results

## The most flexible model 

Fitting the most flexible model to help determine spatial Z hypotheses: Individual populations with an unconstrained Q matrix and unequal U
```{R}

mod.list1 <- list(
  U = "unequal",
  R = "diagonal and equal",
  Q = "unconstrained"
)
fit1 <- MARSS(dat, model=mod.list1, control=list(maxit=1000))
```


```{r}
autoplot(fit1, plot.type="fitted.ytT")

library(corrplot)
Q <- coef(fit1, type="matrix")$Q
corrmat <- diag(1/sqrt(diag(Q))) %*% Q %*% diag(1/sqrt(diag(Q)))
corrplot(corrmat)

```
Some of the populations are correlated. This implies it would be reasonable to test for equal variance covariance, along with some other population structures. 

We opted to test this in a for loop to compare multiple spatial structures, u options, and two options for Q.

## Model Assumptions

Below we setting up population grouping hypotheses (Z matrices), correlation structures, drift or "bias" terms, and the fixed model list.

```{r}

# Here we evaluated the data support for the following hypotheses about Lower Columbia salmon river trends
# Each Z model is a hypothesis

## Simple spatial groupings
# coweeman = coweeman and sf toutle
# ef_lewis = east fork lewis
# kalama = kalama summer and winter
# cowlitz = lower cowlitz, tilton, and upper cowlitz
# washougal = washougal

## Correlation spatial groupings
# coweeman = coweema, ef_lewis, sf toutle, washougal
# kalama = kalama summer and winter
# cowlitz = lower cowlitz, tilton, and upper cowlitz

Z.models <- list(
  H1 = matrix(1,9,1), #one hidden population state
  H2 = factor(c("coweeman","ef_lewis", "kalama_sum",
                'kalama_win', "low_cowlitz", "sf_toutle",
                "tilton", "up_cowlitz", "washougal")), #states are defined by individual population
  H3 = factor(c("coweeman","ef_lewis", "kalama",
                'kalama', "cowlitz", "coweeman",
                "cowlitz", "cowlitz", "washougal")),# states defined by spatial grouping (n = 5)
  H4 = factor(c("coweeman","coweeman", "kalama",
                'kalama', "cowlitz", "coweeman",
                "cowlitz", "cowlitz", "coweeman"))# states defined by correlation (n = 3)
)

names(Z.models) <- c("one_population","indiv_population",'spatial_groups',"corr_groups")

# Also testing different process error varcovar matrices
Q.models2 <- c("equalvarcov","unconstrained")

# Bias terms
u2 <- c("unequal", 'equal')

# Setting fixed portion of mod list 
mod.list = list(
  A = "scaling",
  R = "diagonal and equal")

```

## MARSS Model Selection

Run MARSS models
```{r}


out.tab <- NULL
fits <- list()
for(i in 1:length(Z.models)){
  
  for(Q.model in Q.models2){
    for(U.model in u2){
    fit.model = c(list(Z=Z.models[[i]], Q=Q.model, U=U.model), mod.list)
    fit = MARSS(dat, model=fit.model,
                silent=TRUE, control=list(maxit=5000))
    out=data.frame(H=names(Z.models)[i], Q=Q.model,U=U.model,
                   logLik=fit$logLik, AICc=fit$AICc, num.param=fit$num.params,
                   m=length(unique(Z.models[[i]])),
                   num.iter=fit$numIter, converged=!fit$convergence,
                   stringsAsFactors = FALSE)
    out.tab=rbind(out.tab,out)
    fits=c(fits,list(fit))
    } 
  }
}

min.AICc <- order(out.tab$AICc)
out.tab.1 <- out.tab[min.AICc, ]
out.tab.1 <- cbind(out.tab.1, delta.AICc = out.tab.1$AICc - out.tab.1$AICc[1])
out.tab.1 <- cbind(out.tab.1, rel.like = exp(-1 * out.tab.1$delta.AICc/2))
out.tab.1 <- cbind(out.tab.1, AIC.weight = out.tab.1$rel.like/sum(out.tab.1$rel.like))
out.tab.1


```

The best model (as supported by AICc) includes 9 states (individual populations) that share equal variance and covariance (Q) with the same drift term (U)

## Inspecting best model

```{r}

best_mod <- fits[[6]]
best_mod

autoplot(best_mod, plot.type="fitted.ytT")


```

Inspecting model residuals
```{r}

autoplot(best_mod, plot.type="residuals")
```

## Comparing best model to the most flexible model 

* Do your estimates differ depending on the assumptions you make about the structure of the data, i.e. you assumptions about the x's, Q, and U.

Definitely!

Not sure why autoplot keeps spitting out two plots??

```{r}

autoplot(fit1, plot.type="fitted.ytT")
autoplot(best_mod, plot.type="fitted.ytT")

```

# Discussion

# Description of each team member's contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis of the regions. Team member 4 researched approaches for measuring accuracy of forecasts in [Hyndman & Athanasopoulos[OTexts.com/fpp2] and team member 2 added code for that to the methods. Team member 4 also researched tests for stationarity and worked with team member 2 to code that up. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

