---
title: Lab 2 MARSS Models
author: Liz Elmstrom (SAFS), Terrance Wang, Eric French
date: April 20, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

# General Questions

Each group has the same general tasks, but you will adapt them as you work on the data.

1.  Create estimates of spawner abundance for all missing years and provide estimates of the decline from the historical abundance.

2.  Evaluate support for the major population groups. Are the populations in the groups more correlated than outside the groups?

3.  Evaluate the evidence of cycling in the data. *We will talk about how to do this on the Tuesday after lab.*

Step 1: Estimating data that doesn't exist-- Terrance 
Step 2A: Testing temporal Z matrices (different run times vs. one pop)-- Terrance 
Step 2B: Testing spatial Z matrices (different spatial groupings vs. one pop)-- Liz 
Step 3: Adding in seasonality as a covariate (C/c matrix) -- Eric
Step 4: Evaluating best model out of pool of candidates -- Liz

Describe what data set you will work with and any subsetting you decided on. For example, you may have decided to look only at a section of the ESU that your team was assigned.

# Methods

Address the following in your methods

We tested 5 different population groups hypotheses: one population, individual populations, run timing groups, spatial groups, and correlation groups. The first hypothesis 

-   What will you assume about the U for the x's?

We tested the support for both equal and unequal population growth rates.

-   What will you assume about the Q matrix?

We assumed that these populations were correlated through time and tested model support for unconstrained and equal variance covariance matrices.

-   Write out your assumptions as different models **in matrix form**, fit each and then compare these with AIC or AICc.

Our models are written in matrix form in the "Model Assumptions" section of the results.

-   Do your estimates differ depending on the assumptions you make about the structure of the data, i.e. you assumptions about the x's, Q, and U.

# Data exploration

Load the data.

```{r}
library(tidyverse)
library(dplyr)
library(janitor)
library(MARSS)
library(forecast)
load(here::here("Lab-2", "Data_Images", "columbia-river.rda"))

#Evolutionary Significant Units
esu <- unique(columbia.river$esu_dps)
esu
```

```{r}

df <- columbia.river %>% subset(esu_dps %in% "Steelhead (Lower Columbia River DPS)")

cat("colnames: ", colnames(df), "\n")

unique(df$esapopname)
unique(df$commonpopname)

df$esapopname2 <- stringr::str_replace(df$esapopname, "Steelhead [(]Lower Columbia River DPS[)] ", "")
unique(df$esapopname2)

ggplot(df, aes(x=spawningyear, y=log(value), color=majorpopgroup)) + 
  geom_point(size=2, na.rm = TRUE) + 
  theme(strip.text.x = element_text(size = 3)) +
  theme(axis.text.x = element_text(size = 5, angle = 90)) +
  facet_wrap(~esapopname2) +
  ggtitle('Lower Columbia Steelhead Populations')+
  theme(strip.text.x = element_text(size = 10))

```

Wrangle the data.

```{r}

esuname <- esu[3]
dat <- columbia.river %>% 
  subset(esu_dps == esuname) %>% # get only this ESU
  mutate(log.spawner = log(value)) %>% # create a column called log.spawner
  select(esapopname, spawningyear, log.spawner) %>% # get just the columns that I need
  pivot_wider(names_from = "esapopname", values_from = "log.spawner") %>% 
  column_to_rownames(var = "spawningyear") %>% # make the years rownames
  as.matrix() %>% # turn into a matrix with year down the rows
  t() # make time across the columns
dat[is.na(dat)] <- NA

## Fixing row names
tmp <- rownames(dat)
tmp <- stringr::str_replace(tmp, "Steelhead [(]Lower Columbia River DPS[)]", "")
tmp <- make_clean_names(tmp)
rownames(dat) <- tmp

dat <- dat[1:9,]# remove upper gorge
dat <- dat[ order(row.names(dat)), ]## sort

```

# Results

## The simplest model

Fitting the most simple model to help determine need to group populations: One big population with different observations (populations). Id

```{R}
mod.list.0 <- list(B = matrix(1), U = matrix("u"), Q = matrix("q"), 
    Z = matrix(1, 9, 1), A = "scaling", R = "diagonal and unequal", 
    x0 = matrix("mu"), tinitx = 0)
fit_simple <- MARSS(dat, model=mod.list.0, control=list(maxit=1000))
autoplot(fit_simple, plot.type="fitted.ytT")
```

We see that the simple model is not the best at fitting the population estimates for many of the populations (e.g., Upper Cowlitz winter, East Fork Lewis winter). Many of the residuals do not appear to be stationary.

## The most flexible model

Fitting the most flexible model to help determine spatial Z hypotheses: Individual populations with an unconstrained Q matrix and unequal U

```{R}

mod.list1 <- list(
  U = "unequal",
  R = "diagonal and equal",
  Q = "unconstrained"
)
fit1 <- MARSS(dat, model=mod.list1, control=list(maxit=1000))
```

```{r}
autoplot(fit1, plot.type="fitted.ytT")

library(corrplot)
Q <- coef(fit1, type="matrix")$Q
corrmat <- diag(1/sqrt(diag(Q))) %*% Q %*% diag(1/sqrt(diag(Q)))
corrplot(corrmat)

```

Some of the populations are correlated. This implies it would be reasonable to test for equal variance covariance, along with some other population structures.

We opted to test this in a for loop to compare multiple spatial structures, u options, and two options for Q.

## Model Assumptions

Below we setting up population grouping hypotheses (Z matrices), correlation structures, drift or "bias" terms, and the fixed model list.

```{r}

# Here we evaluated the data support for the following hypotheses about Lower Columbia salmon river trends
# Each Z model is a hypothesis

## Simple run timing groupings
# summer = kalama_river_summer
# winter = remaining population

## Simple spatial groupings
# coweeman = coweeman and sf toutle
# ef_lewis = east fork lewis
# kalama = kalama summer and winter
# cowlitz = lower cowlitz, tilton, and upper cowlitz
# washougal = washougal

## Correlation spatial groupings
# coweeman = coweema, ef_lewis, sf toutle, washougal
# kalama = kalama summer and winter
# cowlitz = lower cowlitz, tilton, and upper cowlitz

Z.models <- list(
  H1 = matrix(1,9,1), #one hidden population state
  H2 = factor(c("coweeman","ef_lewis", "kalama_sum",
                'kalama_win', "low_cowlitz", "sf_toutle",
                "tilton", "up_cowlitz", "washougal")), #states are defined by individual population
  H3 = factor(c("winter","winter", "summer",
                'winter', "winter", "winter",
                "winter", "winter", "winter")),# states defined by running timing grouping (n = 2)
  H4 = factor(c("coweeman","ef_lewis", "kalama",
                'kalama', "cowlitz", "coweeman",
                "cowlitz", "cowlitz", "washougal")),# states defined by spatial grouping (n = 5)
  H5 = factor(c("coweeman","coweeman", "kalama",
                'kalama', "cowlitz", "coweeman",
                "cowlitz", "cowlitz", "coweeman"))# states defined by correlation (n = 3)
)

names(Z.models) <- c("one_population","indiv_population","run_timing_groups",'spatial_groups',"corr_groups")

# Also testing different process error varcovar matrices
Q.models2 <- c("equalvarcov","unconstrained")

# Process cycling terms
# Each is on a 10-year cycle, will test orders 0-4
# see Evaluating Cycling Section

ct.models <- list(
  c0 = "zero", # No cycling
  
  c1 = t(fourier(ts(1:ncol(dat), frequency = 10), K=1)), # seasonality follows a 1st order fourier series with the same # rows as y
 
  c2 = t(fourier(ts(1:ncol(dat), frequency = 10), K=2)), # seasonality follows a 2nd order fourier series
  
  c3 = t(fourier(ts(1:ncol(dat), frequency = 10), K=3)), # seasonality follows a 3rd order fourier series

  c4 = t(fourier(ts(1:ncol(dat), frequency = 10), K=3)) # seasonality follows a 3rd order fourier series
  )
names(ct.models) <- c("no_cycling", "1st_order", "2nd_order", "3rd_order", "4th_order")

# Bias terms
u2 <- c("unequal", 'equal')

# Setting fixed portion of mod list
mod.list = list(
  A = "scaling",
  R = "diagonal and equal")

```

## MARSS Model Selection

Run MARSS models

```{r}
# Added in another for loop to run through cycling options, be warned, this means there are 80 total models. It takes a long time to run
out.tab <- NULL
fits <- list()
for(i in 1:length(Z.models)){
  for (j in 1:length(ct.models)){
    for(Q.model in Q.models2){
      for(U.model in u2){
      fit.model = c(list(Z=Z.models[[i]], c=ct.models[[j]], Q=Q.model, U=U.model), mod.list)
      fit = MARSS(dat, model=fit.model,
                silent=TRUE, control=list(maxit=5000))
      out=data.frame(H=names(Z.models)[i],ct=names(ct.models)[j],Q=Q.model,U=U.model,
                   logLik=fit$logLik, AICc=fit$AICc, num.param=fit$num.params,
                   m=length(unique(Z.models[[i]])),
                   num.iter=fit$numIter, converged=!fit$convergence,
                   stringsAsFactors = FALSE)
      out.tab=rbind(out.tab,out)
      fits=c(fits,list(fit))
      } 
    }
  }
}

min.AICc <- order(out.tab$AICc)
out.tab.1 <- out.tab[min.AICc, ]
out.tab.1 <- cbind(out.tab.1, delta.AICc = out.tab.1$AICc - out.tab.1$AICc[1])
out.tab.1 <- cbind(out.tab.1, rel.like = exp(-1 * out.tab.1$delta.AICc/2))
out.tab.1 <- cbind(out.tab.1, AIC.weight = out.tab.1$rel.like/sum(out.tab.1$rel.like))
out.tab.1


```

Including cycling in the test, the best fit still appears to be Z split into spatial groups, no cyclic behavior, unconstrained Q, and equal drift - Eric



## Inspecting second best model (spatial group, 1st order fourier, unconstrained Q, equal U)

```{r}

best_mod <- fits[[68]]
best_mod

autoplot(best_mod, plot.type="fitted.ytT")
```

Inspecting model residuals

```{r}

# autoplot(best_mod, plot.type="residuals")

par(mfrow = c(3, 3))
resids <- MARSSresiduals(best_mod, type = "tt1")
for (i in 1:9) {
    plot(resids$model.residuals[i, ], ylab = "model residuals", 
        xlab = "")
    abline(h = 0)
    title(rownames(dat)[i])
}
```

Even with added cyclic covariates in the process, there still appears to be some cyclic behavior in the residuals


Evaluating Cycling
Judging by the residuals a 10 year cycle would be a good first guess. Now what order should the series be?
```{r}

#attempting to get this function working
x <- ts(1:ncol(dat), frequency = 10) # following the example in lecture 7, but with a cycle of 10 years
plot.ts(x)
covariate <- fourier(x, K=5) 
plot.ts(covariate)
# got it working, now how does this fit into a MARSS model, maybe in the for loop?
# 3rd order will probably be enough, will test to 4th order to be safe


```

## Comparing best model to the most flexible model

-   Do your estimates differ depending on the assumptions you make about the structure of the data, i.e. you assumptions about the x's, Q, and U.

Definitely!

Not sure why autoplot keeps spitting out two plots??

```{r}

autoplot(fit_simple, plot.type ="fitted.ytT")
autoplot(fit1, plot.type="fitted.ytT")
autoplot(best_mod, plot.type="fitted.ytT")

```

# Discussion

# Description of each team member's contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis of the regions. Team member 4 researched approaches for measuring accuracy of forecasts in [Hyndman & Athanasopoulos[OTexts.com/fpp2] and team member 2 added code for that to the methods. Team member 4 also researched tests for stationarity and worked with team member 2 to code that up. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."
