---
title: "Lab 5 - Time-varying productivity of salmon"
subtitle: "Dynamic Linear Models (DLMs)"
author: "Zoe Rand, Terrance Wang"
date: May 11, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

------------------------------------------------------------------------

# Data

Describe the salmon data and any environmental covariates you decide to examine.

## Load the data

```{r dlm-load-atsa, eval=TRUE}
## library(devtools)
## Windows users will likely need to set this
## Sys.setenv("R_REMOTES_NO_ERRORS_FROM_WARNINGS" = "true")
#devtools::install_github("nwfsc-timeseries/atsalibrary")

## get data
data(KvichakSockeye, package="atsalibrary")
SR_data <- KvichakSockeye
```

The data are a dataframe with columns for brood year (`brood_year`), number of spawners (`spawners`), number of recruits (`recruits`) and PDO at year $t-2$ in summer (`pdo_summer_t2`) and in winter (`pdo_winter_t2`).

```{r dlm-data-head}
## head of data file
head(SR_data)
```

## Wrangle the data

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
```

```{r wrangle_data}
#removing years with no spawner data
#creating log (recruits/spawners)
SR_data<-SR_data %>% filter(!is.na(spawners)) %>%mutate(logSR = log(recruits/spawners))
head(SR_data)
ggplot(SR_data) + geom_line(aes(x = brood_year, y = logSR))
```

# General tasks

Use the information and data in the previous section to answer the following questions. Note that if any model is not converging, then you will need to increase the `maxit` parameter in the `control` argument/list that gets passed to `MARSS()`. For example, you might try `control=list(maxit=2000)`.

1.  Begin by fitting a reduced form of Ricker's model that includes only a time-varying level ($\alpha_t$) and observation error ($v_t$). That is,

```{=tex}
\begin{align*}
\text{log}(R_t) &= \alpha_t + \text{log}(S_t) + v_t \\
\text{log}(R_t/S_t) &= \alpha_t + v_t
\end{align*}
```
This model assumes no density-dependent survival in that the number of recruits is an ascending function of spawners. Plot the ts of $\alpha_t$ and note the AICc for this model. Also plot appropriate model diagnostics.

2.  Fit the full Ricker model with a time-varying intercept and a static effect of spawners. For this model, obtain the time series of $\alpha_t$, which is an estimate of the stock productivity in the absence of density-dependent effects. How do these estimates of productivity compare to those from the previous question? Plot the ts of $\alpha_t$ and note the AICc for this model. Also plot appropriate model diagnostics. ($Hint$: If you don't want a parameter to vary with time, what does that say about its process variance?)

3.  Fit the expanded model that includes the summer PDO index as the covariate (`pdo_summer_t2`). What is the mean level of productivity? Plot the ts of $\delta_t$ and note the AICc for this model. Also plot appropriate model diagnostics.

4.  Fit the expanded model that includes the winter PDO index as the covariate (`pdo_winter_t2`). What is the mean level of productivity? Plot the ts of $\delta_t$ and note the AICc for this model. Also plot appropriate model diagnostics.

5.  Based on AICc, which of the models above is the most parsimonious? Is it well behaved (i.e., are the model assumptions met)? Plot the model forecasts for the best model. Is this a good forecast model?

6.  Consider other environmental factors that could influence the time-varying intrinsic productivity of these salmon. For example, are there other large-scale indices of ocean conditions (e.g., sea-surface temperature, El Niño Southern Oscillation)? You could also consider the possible influence of other salmon stocks such as those we examined in Lab 1.

# Methods

Please address the following in your methods:

-   What environmental or dummy variables did you include and why?

-   What forms of models did you fit (ie, write them out in matrix form)?

-   What sort of model diagnostics did you use to examine model assumptions?

## 1. Reduced Ricker Model:

### Equations:

$$
\text{log}(R_t/S_t) = \alpha_t + v_t
$$

$$
\alpha_t = \alpha_{t-1} + w_t
$$

This is a MARSS model where $Z = 1$, a random walk with observation error:

Set $y_t = log(R_t/S_t)$

$$
y_t = Zx_t + v_t
$$

$$
x_t = x_{t-1} + w_t
$$

### MARSS:

```{r, message = FALSE, warning = FALSE}
library(MARSS)
```

```{r}
mod_list1<-list(B = "identity", Z = "identity", U = "zero", A = "zero", Q = matrix("q"), R = matrix("r"))
```

```{r}
#fit model with MARSS:
fit1<-MARSS(SR_data$logSR, mod_list1)
```

```{r}
#model diagnostics
autoplot(fit1, plot.type="fitted.ytT")
```

## 2. Full Ricker Model:

### Equations:

$$
\text{log}(R_t/S_t) = \alpha_t - bS_t +  v_t
$$

$$
\alpha_t = \alpha_{t-1} + w_t
$$

### MARSS:

$$
y_t = Zx_t + v_t
$$

$$
v_t \sim N(0, r)
$$

$$
x_t = Bx_{t-1} + w_t
$$

$$
w_t \sim N(0, \begin{bmatrix} q & 0 \\ 0 & 0 \end{bmatrix})
$$

where $B = \begin{bmatrix}1 & 0\\0 & 1\end{bmatrix}$, $x_t = \begin{bmatrix} \alpha_t \\ b \end{bmatrix}$, $Z = \begin{bmatrix} 1 & S_t \end{bmatrix}$

```{r}
#setting up model
Bmat <- matrix(c(1,0,0,1), 2, 2, byrow = TRUE)
TT<-nrow(SR_data)
ZZ <- array(NA, c(1, 2, TT))  
ZZ[1,1,] <- rep(1, TT) #intercept
ZZ[1,2,]<-SR_data$spawners #St
QQ<-matrix(list(0), 2, 2)
QQ[1,1]<-"q"
mod_list2<-list(B = Bmat, Z = ZZ, U = "zero", A = "zero", Q = QQ, R = matrix("r"))

inits_list <- list(x0 = matrix(c(0, 0), nrow = 2))
```

```{r}
fit2<-MARSS(SR_data$logSR, mod_list2, inits = inits_list)
```

```{r}
#model diagnostics
autoplot(fit2, plot.type="fitted.ytT")
```

## 3. Full Ricker Model with Summer PDO

### Equations:

$$
\text{log}(R_t/S_t) = \alpha_t + \delta_t X_t - bS_t +  v_t
$$

$$
\alpha_t = \alpha_{t-1} + w_t
$$

$$
\delta_t = \delta_{t-1} + w_t
$$

### MARSS:

$$
y_t = Zx_t + v_t
$$

$$
v_t \sim N(0, r)
$$

$$
x_t = Bx_{t-1} + w_t
$$

$$
w_t \sim N(0, \begin{bmatrix} q_\alpha & 0 & 0\\ 0 & q_\delta & 0 \\ 0 & 0 & 0 \end{bmatrix})
$$

where $B = \begin{bmatrix}1 & 0 & 0 \\0 & 1 & 0 \\ 0&0&1\end{bmatrix}$, $x_t = \begin{bmatrix} \alpha_t \\ \delta_t \\ b \end{bmatrix}$, $Z = \begin{bmatrix} 1 & X_t & S_t \end{bmatrix}$

```{r}
#setting up model
Bmat <- matrix(c(1,0,0,0,1,0,0,0,1), 3, 3, byrow = TRUE)
TT<-nrow(SR_data)
ZZ <- array(NA, c(1, 3, TT))  
ZZ[1,1,] <- rep(1, TT) #intercept
ZZ[1,2,]<-(SR_data$pdo_summer_t2) #summer PDO
ZZ[1,3,]<-(SR_data$spawners) #St
QQ<-matrix(list(0), 3, 3)
QQ[1,1]<-"q_alpha"
QQ[2,2]<-"q_delta"
mod_list3<-list(B = Bmat, Z = ZZ, U = "zero", A = "zero", Q = QQ, R = matrix("r"))

inits_list <- list(x0 = matrix(c(0, 0, 0), nrow = 3))
```

```{r}
fit3<-MARSS(SR_data$logSR, mod_list3, inits = inits_list,control = list(maxit=1000,allow.degen=F))
```

## 4. Full Ricker Model with Winter PDO

The equations for #4 are identical with #3, with the exception of Winter PDO as $X_t$

```{r}
#setting up model
Bmat <- matrix(c(1,0,0,0,1,0,0,0,1), 3, 3, byrow = TRUE)
TT<-nrow(SR_data)
ZZ <- array(NA, c(1, 3, TT))  
ZZ[1,1,] <- rep(1, TT) #intercept
ZZ[1,2,]<-(SR_data$pdo_winter_t2) #summer PDO
ZZ[1,3,]<-(SR_data$spawners) #St
QQ<-matrix(list(0), 3, 3)
QQ[1,1]<-"q_alpha"
QQ[2,2]<-"q_delta"
mod_list3<-list(B = Bmat, Z = ZZ, U = "zero", A = "zero", Q = QQ, R = matrix("r"))

inits_list <- list(x0 = matrix(c(0, 0, 0), nrow = 3))

fit4<-MARSS(SR_data$logSR, mod_list3, inits = inits_list,control = list(maxit=1000,allow.degen=F))
```

# Results

## 1. Reduced Ricker Model

Plot of model fit (time series of $\alpha_t$)

```{r}
ggplot(data = SR_data) + geom_line(aes(x = brood_year, y = logSR), color = "black") + geom_ribbon(aes(x = brood_year, ymin = t(fit1$states - 2*fit1$states.se), ymax = t(fit1$states + 2*fit1$states.se)), linetype = "dashed", color = "blue", fill = NA) + geom_line(aes(x = brood_year, y = t(fit1$states)), color = 'blue') + theme_classic()
```

## 2. Full Ricker Model

Time series of regression parameters

```{r}
#intercept
int_mn<-fit2$states[1,]
int_se<-fit2$states.se[1,]
ggplot() + geom_ribbon(aes(x = SR_data$brood_year, ymin = int_mn-2*int_se, ymax = int_mn + 2*int_se), linetype = "dashed", fill = NA, color = "blue") + geom_line(aes(x = SR_data$brood_year, y = int_mn), color = "blue") + labs(x = "Brood Year", y = "alpha_t")
```

```{r}
#slope results
slp_mn<-fit2$states[2,1]
slp_se<-fit2$states.se[2,1]
print(paste("Mean: ", slp_mn))
print(paste("SE: ", slp_se))
```

## 3. Full Ricker Model with Summer PDO

```{r}
#intercept
int_mn<-fit3$states[1,]
int_se<-fit3$states.se[1,]
ggplot() + geom_ribbon(aes(x = SR_data$brood_year, ymin = int_mn-2*int_se, ymax = int_mn + 2*int_se), linetype = "dashed", fill = NA, color = "blue") + geom_line(aes(x = SR_data$brood_year, y = int_mn), color = "blue") + labs(x = "Brood Year", y = "alpha_t")
```

```{r}
#summer pdo slope results
slp_mn<-fit3$states[2,]
slp_se<-fit3$states.se[2,]
ggplot() + geom_ribbon(aes(x = SR_data$brood_year, ymin = slp_mn-2*slp_se, ymax = slp_mn + 2*slp_se), linetype = "dashed", fill = NA, color = "blue") + geom_line(aes(x = SR_data$brood_year, y = slp_mn), color = "blue") + labs(x = "Brood Year", y = "delta_t (SummerPDO)")

```

```{r}
# spawner slope results
slp_mn<-fit3$states[3,1]
slp_se<-fit3$states.se[3,1]
print(paste("Mean: ", slp_mn))
print(paste("SE: ", slp_se))
```

## 4. Full Ricker Model with Winter PDO

```{r}
#intercept
int_mn<-fit4$states[1,]
int_se<-fit4$states.se[1,]
ggplot() + geom_ribbon(aes(x = SR_data$brood_year, ymin = int_mn-2*int_se, ymax = int_mn + 2*int_se), linetype = "dashed", fill = NA, color = "blue") + geom_line(aes(x = SR_data$brood_year, y = int_mn), color = "blue") + labs(x = "Brood Year", y = "alpha_t")

#winter pdo slope results
slp_mn<-fit4$states[2,]
slp_se<-fit4$states.se[2,]
ggplot() + geom_ribbon(aes(x = SR_data$brood_year, ymin = slp_mn-2*slp_se, ymax = slp_mn + 2*slp_se), linetype = "dashed", fill = NA, color = "blue") + geom_line(aes(x = SR_data$brood_year, y = slp_mn), color = "blue") + labs(x = "Brood Year", y = "delta_t (WinterPDO)")

# spawner slope results
slp_mn<-fit4$states[3,1]
slp_se<-fit4$states.se[3,1]
print(paste("Mean: ", slp_mn))
print(paste("SE: ", slp_se))
```

## Model Comparison

```{r}
AIC_tab<-tibble("Model" = c("Reduced Ricker", "Full Ricker", "Full w/ Summer PDO", "Full w/ Winter PDO"), "AICc" = c(fit1$AICc, fit2$AICc, fit3$AICc, fit4$AICc))
AIC_tab<- AIC_tab %>% mutate(DeltaAIC = AICc-min(AICc))
AIC_tab
```

Reduced ricker is the better model, which makes sense since the esimate for the b parameter is essentially 0, so both models are basically the same except the full ricker has an extra parameter.

# Discussion

# Team contributions

ZR did the data wrangling and wrote the methods, code, and results for the first two models (reduced and full Ricker). TW wrote the methods, code, and results for the models with covariates. Both team members contributed to the discussion.
