---
title: "Lab-1-team-1_Maria2.Rmd"
author: "Maria Kuruvilla"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(tidyverse)
library(tseries)
library(forecast)
library(zoo)


options(dplyr.summarise.inform = FALSE)
```


## Wood River


## Read the data
```{r}
bb_data <- readRDS(here::here("Lab-1", "Data_Images", 
                              "bristol_bay_data_plus_covariates.rds"))
head(bb_data)

subdata <- bb_data %>%
  group_by(system, ret_yr) %>%
  summarize(lntotal = log(sum(ret, na.rm=TRUE)))%>%
  filter(system == "Wood")
head(subdata)

```



## Plot the data

Plot the data and discuss any obvious problems with stationarity from your visual test.
```{r, echo = FALSE}
plot1 <- ggplot(data = subdata, aes(x=ret_yr, y=lntotal)) + 
  geom_line() + 
  ggtitle("log abundance by river") +
  facet_wrap(~system)
plot1

```


It looks like there is a trend in Wood river log returns and maybe some negative autocorrelation.


## Use ACF and PACF

Use the ACF and PACF plots to explore the time series. What did you learn? Also try decomposition.

```{r}


data_ts_Wood <- ts(subdata$lntotal[subdata$system == "Wood"],
                      start=subdata$ret_yr[subdata$system == "Wood"][1])

train_Wood <- window(data_ts_Wood, 1963, 2005)
test_Wood <- window(data_ts_Wood, 2006, 2020)


acf(data_ts_Wood)

pacf(data_ts_Wood)

```


The acf plots for Wood river is slowly decaying and the pacf plot shows significance at lab 1. There might be AR1 structure in the data.

```{r message=FALSE}
mod_Wood <- auto.arima(data_ts_Wood)
mod_Wood
```




```{r message=FALSE}


adf.test(data_ts_Wood, k = 0)



kpss.test(data_ts_Wood, null =c("Trend"))
#note that this gives different results depending on what null is

#wood is stationary around trend



```

Wood river data is probably stationary around a trend.

```{r}
ndiffs(data_ts_Wood, test='kpss')

ndiffs(data_ts_Wood, test='adf')
```


## Differencing the data
```{r}

Wood_diff <- diff(data_ts_Wood)

plot.ts(Wood_diff)

```

This looks pretty stationary.


## Stationarity on differenced data



```{r}
adf.test(Wood_diff, k = 0)



kpss.test(Wood_diff, null =c("Level"))

```

These results say that Wood river data is also probably stationary.



```{r}

train_Wood <- window(data_ts_Wood, 1963, 2015)
test_Wood <- window(data_ts_Wood, 2015, 2020)
mod_Wood_train <- auto.arima(train_Wood)
mod_Wood_train
### Notes from Eli, testing with data from the Wood. (auto.arima chose a 0,1,1 model)

# Chosing our own arima structure. This is how we would write it if we just wanted a single model with a specified structure. 
#fit = Arima(train_Wood, order = c(2,1,0))# just a random example of how its written, didn't chose the order for any reason

# For loop to test multiple models and find their accuracy
mod.list <- list(c(0,1,1), c(1,1,1), c(2,1,1), c(1,0,1))## List of mods to test. First option is what auto.arima chose
fits <- list()# list to store the model fits in 
err.list <- list(0) # list for accuracy data
for(i in 1:4){
  fits[[i]] <- Arima(train_Wood, order = mod.list[[i]])# loops through list, fits model, and stores in 'fits'
  err.list[[i]] <- accuracy(forecast(fits[[i]], h =5), test_Wood) # puts accuracy data into list
}

fits
err.list

```


## Plot forecasts

```{r}
# Testing plot with the last ARIMA model (2,0,1)


# fr.test <- forecast(fits[[3]], h=30)
# autoplot(fr.test) + geom_point(aes(x=x, y=y), data=fortify(test_Wood))+
#   geom_point(aes(x=ret_yr, y=lntotal), data = subdata, col='red')

# testing all models

plots <- list()
for(i in 1:4){
  fr.test <- forecast(fits[[i]], h=20)
  plots[[i]] <- autoplot(fr.test) + geom_point(aes(x=x, y=y), data=fortify(test_Wood))+
  geom_point(aes(x=ret_yr, y=lntotal), data = subdata, col='red')
  
plots[[i]]
}

plots[[1]]
plots[[2]]
plots[[3]]
plots[[4]]
```


