---
output:
  pdf_document: default
  html_document: default
---
# Lab 1: ARIMA models

Team member names: Nick Chambers (SAFS), Liz Elmstrom (SAFS), Maria Kuruvilla (QERM)


# Data

We will use Bristol Bay Data. We want to use sockeye salmon returns for each river for each year.

# Question your team will address

Our team decided to compare the accuracy of forecasts using best fit ARIMA models for Sockeye salmon in different rivers in Bristol Bay data. Our question is whether forecast accuracy is different for different rivers.





# Method you will use

* fit ARIMA models (note you'll want to log the abundance data). You can fit other models in addition ot ARIMA if you want.
* do diagnostics for ARIMA models
* make forecasts
* test how good your forecasts or compare forecasts (many options here)

## Initial plan

Describe what you plan to do to address your question. Note this example is with Ruggerone & Irvine data but your team will use the Bristol Bay data.

Example, "We will fit ARIMA models to 1960-1980 data on pink salmon in 2 regions in Alaska and 2 regions in E Asia. We will make 1981-1985  forecasts (5 year) and compare the accuracy of the forecasts to the actual values. We will measure accuracy with mean squared error. We will use the forecast package to fit the ARIMA models."

## What you actually did

Example, "We were able to do our plan fairly quickly, so after writing the basic code, we divided up all 12 regions in Ruggerone & Irvine and each team member did 4 regions. We compared the accuracy of forecasts for different time periods using 20-years for training and 5-year forecasts each time. We compared the RMSE, MAE, and MAPE accuracy measures."

# Diagnostics and preliminary exploration
## Read the data
```{r}
bb_data <- readRDS(here::here("Lab-1", "Data_Images", 
                              "bristol_bay_data_plus_covariates.rds"))
head(bb_data)

```
## Checking out the unique data groups
```{r echo=FALSE}
cat("colnames: ", colnames(bb_data), "\n")
cat("system (river): ", unique(bb_data$system), "\n")
cat("age groups: ", unique(bb_data$age_group), "\n")

```
## Retrieving a subset of the data

```{r}
library(tidyverse)
subdata <- bb_data %>%
  group_by(system, ret_yr) %>%
  summarize(lntotal = log(sum(ret, na.rm=TRUE)))%>%
  filter(system == "Ugashik" || system == "Wood")
head(subdata)
```


## Plot the data

Plot the data and discuss any obvious problems with stationarity from your visual test.
```{r}
plot1 <- ggplot(data = subdata, aes(x=ret_yr, y=lntotal)) + 
  geom_line() + 
  ggtitle("log abundance by river") +
  facet_wrap(~system)
plot1

```

It looks like there is a trend in Wood river log returns and maybe some negative autocorrelation. Variance in Ugashik seems to be higher until 1980 and lower afterwards.


## Use ACF and PACF

Use the ACF and PACF plots to explore the time series. What did you learn? Also try decomposition.

```{r}

data_ts_Ugashik <- ts(subdata$lntotal[subdata$system == "Ugashik"],
                      start=subdata$ret_yr[subdata$system == "Ugashik"][1])
data_ts_Wood <- ts(subdata$lntotal[subdata$system == "Wood"],
                      start=subdata$ret_yr[subdata$system == "Wood"][1])
train_Ugashik <- window(data_ts_Ugashik, 1963, 2005)
test_Ugashik <- window(data_ts_Ugashik, 2006, 2020)

train_Wood <- window(data_ts_Wood, 1963, 2005)
test_Wood <- window(data_ts_Wood, 2006, 2020)

acf(data_ts_Ugashik)

pacf(data_ts_Ugashik)

acf(data_ts_Wood)

pacf(data_ts_Wood)

```

The acf plots for both rivers are both slowly decaying and the pacf plots for both rivers show significants at various lags. In the Ugashik river lags 1, 4 and 7 are significant whereas in the Wood river, only lag 1 is significant. There might be AR1 structure in the ata from both rivers with some potentially higher order AR terms in Ugashik.

## Test for stationarity

Run tests and discuss any stationarity issues and how these were addressed.

```{r message=FALSE}
library(forecast)
mod_Ugashik <- auto.arima(data_ts_Ugashik)
mod_Ugashik

mod_Wood <- auto.arima(data_ts_Wood)
mod_Wood
```

auto.arima suggests that ARIMA(0,1,0) fits Ugashik river data the best and ARIMA(1,1,1) fits the Wood river data the best.

```{r message=FALSE}
library(tseries)
adf.test(data_ts_Ugashik, k = 0)



kpss.test(data_ts_Ugashik, null =c("Level"))
##Tests give different results
## Ugashik data is probably NON stationary


adf.test(data_ts_Wood, k = 0)



kpss.test(data_ts_Wood, null =c("Trend"))
#note that this gives different results depending on what null is

#wood is stationary around trend



```

Wood river data is probably stationary around a trend, but tests give differnet results for Ugashik data. Ugashik data is probably non stationary.

```{r}
ndiffs(data_ts_Ugashik, test='kpss')

ndiffs(data_ts_Ugashik, test='adf')


```
```{r}
ndiffs(data_ts_Wood, test='kpss')

ndiffs(data_ts_Wood, test='adf')
```
## Differencing the data
```{r}
Ugashik_diff <- diff(data_ts_Ugashik)
Wood_diff <- diff(data_ts_Wood)

plot.ts(Ugashik_diff)
plot.ts(Wood_diff)

```
This looks pretty stationary.

## Stationarity on differenced data

```{r}
adf.test(Ugashik_diff, k = 0)



kpss.test(Ugashik_diff, null =c("Level"))

```

These results say Ugashik data is probably stationary.


```{r}
adf.test(Wood_diff, k = 0)



kpss.test(Wood_diff, null =c("Level"))

```
These results say that Wood river data is also probably stationary.

## Using auto.arima
```{r message=FALSE}
library(forecast)
mod_Ugashik <- auto.arima(data_ts_Ugashik)
mod_Ugashik

mod_Wood <- auto.arima(data_ts_Wood)
mod_Wood
```
```{r}
library(zoo)
fr <- forecast(mod_Ugashik, h=30)
autoplot(fr) + geom_point(aes(x=x, y=y), data=fortify(test_Ugashik))
checkresiduals(fr)
```

```{r}
library(zoo)
fr_Wood <- forecast(mod_Wood, h=30)
autoplot(fr_Wood) + geom_point(aes(x=x, y=y), data=fortify(test_Wood))
checkresiduals(fr_Wood)
```


```{r}
library(zoo)
train_Wood <- window(data_ts_Wood, 1963, 1990)
test_Wood <- window(data_ts_Wood, 1990, 2020)
mod_Wood_train <- auto.arima(train_Wood)
mod_Wood_train
fr_Wood_train <- forecast(mod_Wood_train, h=30)
autoplot(fr_Wood_train) + geom_point(aes(x=x, y=y), data=fortify(test_Wood))
checkresiduals(fr_Wood_train)
```


# Results

# Discussion

# Description of each team member's contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis of the regions. Team member 4 researched approaches for measuring accuracy of forecasts in [Hyndman & Athanasopoulos[OTexts.com/fpp2] and team member 2 added code for that to the methods. Team member 4 also researched tests for stationarity and worked with team member 2 to code that up. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

