---
editor_options: 
  markdown: 
    wrap: 72
---

# Lab 1: ARIMA models

Team member names: Eric French (Civil), Dylan Hubl (ESRM), Miranda Mudge
(Molecular & Cell Bio)

# Data

Bristol Bay Data.

```{r}
library(tidyverse)
library(ggplot2)
library(forecast)
library(zoo)
```

Read in the data

```{r}
bb_data <- readRDS(here::here("Lab-1", "Data_Images", "bristol_bay_data_plus_covariates.rds"))
```

Explore the data

```{r}
bb_data_reduced <- bb_data[,c('ret_yr', 'ret', 'system', 'age_group', 
                              'forecast.adfw', "forecast.fri", 
                              'env_pdo', 'env_sst', 'env_slp', 'env_upstr')]
#str(BB_data_reduced)

```

We will need to remove NA value to avoid dip in 1.3 data...

# Data of interest for our question

```{r}
region = "Ugashik"

bb_data %>% 
  filter(system==region) %>% 
  filter(age_group== 1.3 | age_group==2.2) %>% 
  ggplot(aes(x=ret_yr, y=log(ret))) + 
    geom_line() + 
    ggtitle("Log abundance by age group in Ugashik") +
    facet_wrap(~age_group)
```

# Interested in log totals grouped by different times spent in fresh water vs ocean water for 4 year old fish in Ugashik region.

Get one time series (ret_yr==2.2 or 1.3) and split into train and test.
Each with 10 years. Ran models for both age groups using same code below
by changing age variable below.

```{r}
age = 1.3

dat <- bb_data %>%
  filter(system==region) %>%
  filter(age_group == age) %>%
  mutate(lnreturns = log(ret), #returns same values as code above
         year = ret_yr) %>%
  select(year, lnreturns)

datts <- ts(dat$lnreturns, start=dat$year[1])

train <- window(datts, 1963, 2010) # range 1963 - 2010
test <- window(datts, 2011, 2020) # range 2011 - 2020
```

## Visualize the time series

```{r}
plot.ts(datts, ylab = "Log abundance") #plot time series for ret_yr
```

Data for 2 yr freshwater-2 yr ocean water fish show variance around a
mean of 5, suggesting differencing will be necessary to achieve
stationarity.

## Data exploration through differencing

```{r}

datts_diff <- diff(datts, differences = 1)
plot(datts_diff, ylab = "Log abundance first differenced")

```

First differencing the data centers the variance around zero - indicates
an AR(1) model would be appropriate.

## ACF

Tails off slowly -\> indicates AR(1) model Multiple lags significant
with potential seasonality...going to ignore for this assignment by
recommendation of Eli

```{r}
acf(datts)
```

## PACF

Cuts off after lag 1 -\> indicates AR(1) model

```{r}
pacf(datts)
```

## Decomposition

When attempting decomposition, error: time series has no or less than 2
periods. Suggests the time series has no seasonality, or not enough
seasonality to be measured.

```{r}
datts_decomp <- decompose(datts)
```

# Model selection

Select yrs 1963-2010 to train model starting with ret_yr

```{r}
dat_model <- auto.arima(train)
dat_model
```

### Results for ret_yr 2.2

Series: train ARIMA(1,0,0) with non-zero mean

Coefficients: ar1 mean 0.4545 5.9958 s.e. 0.1268 0.3373

sigma\^2 = 1.753: log likelihood = -80.67 AIC=167.34 AICc=167.88
BIC=172.95

ARIMA model results from train data for ret_yr 2.2 fish in Bristol Bay
consistent with results of ACF and PACF tests suggesting AR(1). First
differencing the time series object returned the mean to zero, but
auto.arima suggests 0 differencing and that data has non-zero mean.

### Results for ret_yr 1.3

Series: train ARIMA(0,1,1)

Coefficients: ma1 -0.7305 s.e. 0.1400

sigma\^2 = 18.34: log likelihood = -134.93 AIC=273.86 AICc=274.13
BIC=277.56

ARIMA model results from train data for ret_yr 1.3 fish in Bristol Bay
very different from ret_yr 2.2 despite both being 4 yr old fish. Ret_yr
1.3 inconsistent with results of ACF and PACF tests suggesting AR(1).
First differencing the time series object returned the mean to zero, and
auto.arima agrees first differencing the data is the best approach for
model.

## Fit the model to train data

```{r}
fit <- forecast::auto.arima(train)
```

## Test the accuracy of the model using forecast

```{r}
fr <- accuracy(forecast(fit, h=10), test)
autoplot(fr) + geom_point(aes(x=x, y=y), data=fortify(test))
```

### Results for 2.2 fish in Ugashik

```         
                   ME     RMSE       MAE       MPE     MAPE      MASE        ACF1 Theil's U
```

Training set -0.001068187 1.295973 1.0706311 -6.326379 21.04040
0.8937811 -0.01828833 NA Test set -0.314622455 1.051713 0.7971697
-9.454614 16.49798 0.6654909 0.31998321 0.9446332

### Results for 1.3 fish in Ugashik

```         
                 ME      RMSE       MAE       MPE      MAPE      MASE        ACF1 Theil's U
```

Training set 0.1861463 4.1924404 2.0115438 7.247855 40.584446 0.9675341
0.007029462 NA Test set -0.2848240 0.6145452 0.3903876 -4.698664
6.086718 0.1877728 0.098332425 0.810348

```{r}

```
