---
editor_options: 
  markdown: 
    wrap: 72
---

# Lab 1: ARIMA models

Team member names: Eric French (Civil), Dylan Hubl (ESRM), Miranda Mudge
(Molecular & Cell Bio)

# Data

Bristol Bay Data.

```{r}
library(tidyverse)
library(ggplot2)
library(forecast)
library(zoo)
```

# Read in the data

```{r}
bb_data <- readRDS(here::here("Lab-1", "Data_Images", "bristol_bay_data_plus_covariates.rds"))
```

# Data of interest for our question

```{r}
region = "Ugashik"

bb_data %>% 
  filter(system==region) %>% 
  filter(age_group== 1.3 | age_group==2.2) %>% 
  ggplot(aes(x=ret_yr, y=log(ret))) + 
    geom_line() + 
    ggtitle("Log abundance by age group in Ugashik") +
    facet_wrap(~age_group)
```

We may need to remove NA value to avoid dip in 1.3 data...

# Interested in log totals grouped by different times spent in fresh water vs ocean water for 4 year old fish in Ugashik region.

Get one time series (ret_yr==2.2 or 1.3) and split into train and test.
Each with 10 years. Ran models for both age groups using same code below
by changing age variable below.

## Subset and plot time series

1.3 = 1 year freshwater, 3 years ocean 2.2 = 2 year freshwater, 3 years
ocean

```{r}
age1.3 = 1.3
age2.2 = 2.2

dat_1.3 <- bb_data %>%
  filter(system==region) %>%
  filter(age_group == age1.3) %>%
  mutate(lnreturns = log(ret), #returns same values as code above
         year = ret_yr) %>%
  select(year, lnreturns) 

dat_2.2 <- bb_data %>%
  filter(system==region) %>%
  filter(age_group == age2.2) %>%
  mutate(lnreturns = log(ret), #returns same values as code above
         year = ret_yr) %>%
  select(year, lnreturns)


datts_1.3 <- ts(dat_1.3$lnreturns, start=dat_1.3$year[1]) # time series 1.3 fish 
datts_1.3_out <- window(datts_1.3, start=1980, end = 2020) # time series 1.3 fish 
datts_2.2 <- ts(dat_2.2$lnreturns, start=dat_2.2$year[1]) # time series 2.2 fish

plot.ts(datts_1.3, ylab = "Log abundance", main = "Time series of fish: 1 year freshwater, 3 years ocean") 
plot.ts(datts_1.3_out, ylab = "Log abundance", main = "Time series of fish: 1 year freshwater, 3 years ocean") 
plot.ts(datts_2.2, ylab = "Log abundance", main = "Time series of fish: 2 year freshwater, 2 years ocean") 
```

Note that after looking at 1.3 data, went back to experiment with how
removing outlier value would affect data and overall model for 1.3 fish.
After removing negative value, the time series for 1 year fresh water -
3 year ocean water fish looks very different.

Data for 1 yr freshwater-3 yr ocean water fish show variance around a
mean of 5. Removal of outlier shows higher variance prior to 1980 with a
rise in log abundance after. Data for 2 yr freshwater-2 yr ocean water
fish show variance around a mean of 5. This suggests differencing may be
necessary to achieve stationarity.

## Data exploration through differencing

```{r}
datts_diff_1.3 <- diff(datts_1.3, differences = 1)
datts_diff_1.3_out <- diff(datts_1.3_out, differences = 1)
datts_diff_2.2 <- diff(datts_2.2, differences = 1)

plot(datts_diff_1.3, ylab = "Log abundance", main = "Time series, first differenced: 1 year freshwater, 3 years ocean")
plot(datts_diff_1.3_out, ylab = "Log abundance", main = "Time series, first differenced: 1 year freshwater, 3 years ocean, outlier removed")
plot(datts_diff_2.2, ylab = "Log abundance", main = "Time series, first differenced: 2 year freshwater, 2 years ocean")

```

1 year fresh water - 3 year ocean water: First differencing the data
centers the variance around zero. Originally non-zero mean.

2 year fresh water - 2 year ocean water:

## ACF

1 year fresh water - 3 year ocean water: Tails off slowly which
indicates AR(1) model Multiple lags significant with potential
seasonality...going to ignore for this assignment by recommendation of
Eli

```{r}
acf(datts_1.3)
acf(datts_1.3_out, na.action = na.pass) # allows na value
acf(datts_2.2)
```

## PACF

1 year fresh water - 3 year ocean water:Cuts off after lag 1 which
indicates AR(1) model

```{r}
pacf(datts_1.3)
pacf(datts_1.3_out,na.action = na.pass)
pacf(datts_2.2)
```

## Decomposition

When attempting decomposition, error: time series has no or less than 2
periods. Suggests the time series has no seasonality, or not
enoughseasonality to be measured.

```{r}
datts_decomp <- decompose(datts_2.2)
```

## Testing for stationarity

## 1.3 fish

1 year fresh water - 3 year ocean water: - Dickey-Fuller = -3.1174, Lag
order = 3, p-value = 0.1236, alternative hypothesis: stationary -- The
null hypothesis for both this test is that the data are non-stationary,
we want to reject but p-value too high. Test suggests alternative
hypothesis: stationary

-   KPSS Level = 0.54744, Truncation lag parameter = 3, p-value =
    0.03098 -- The p-value is less than 0.05. The null hypothesis of
    stationarity around a level is rejected.

-   KPSS Trend = 0.075997, Truncation lag parameter = 3, p-value = 0.1
    -- The null hypothesis of stationarity around a trend is not
    rejected.This confirms the above test that the data is stationary
    around a trend.

```{r}
tseries::adf.test(datts_1.3) 
tseries::kpss.test(datts_1.3, null = "Level") 
tseries::kpss.test(datts_1.3, null = "Trend") 
```

## 1.3 fish, outlier removed

-   Dickey-Fuller = -3.4337, Lag order = 3, p-value = 0.06582 -- The
    null hypothesis for both this test is that the data are
    non-stationary, we want to reject, that is confirmed here suggesting
    data are stationary.

-   KPSS Level = 0.26391, Truncation lag parameter = 3, p-value = 0.1 --
    The p-value is greater than 0.05. The null hypothesis of
    stationarity around a level is not rejected, data stationary

-   KPSS Trend = 0.065518, Truncation lag parameter = 3, p-value = 0.1
    -- The null hypothesis of stationarity around a trend is not
    rejected. This confirms the above test that the data is stationary
    around a trend.

```{r}
tseries::adf.test(datts_1.3_out) 
tseries::kpss.test(datts_1.3_out, null = "Level") 
tseries::kpss.test(datts_1.3_out, null = "Trend") 
```

## 2.2 fish

```{r}
tseries::adf.test(datts_2.2) 
tseries::kpss.test(datts_2.2, null = "Level") 
tseries::kpss.test(datts_2.2, null = "Trend") 
```

# Model selection

Select yrs 1963-2010 to train model starting with ret_yr

```{r}
# Make dataframes for testing models later
train_1.3 <- window(datts_1.3, 1963, 2010) # range 1963 - 2010
test_1.3 <- window(datts_1.3, 2011, 2020) # range 2011 - 2020

train_1.3_out <- window(datts_1.3_out, 1980, 2010) # range 1963 - 2010
test_1.3_out <- window(datts_1.3_out, 2011, 2020) # range 2011 - 2020

train_2.2 <- window(datts_2.2, 1963, 2010) # range 1963 - 2010
test_2.2 <- window(datts_2.2, 2011, 2020) # range 2011 - 2020
```

# Train the models

```{r}
model_1.3 <- auto.arima(train_1.3)
model_1.3

model_1.3_out <- auto.arima(train_1.3_out)
model_1.3_out

model_2.2 <- auto.arima(train_2.2)
model_2.2
```

### Printed results for ret_yr 1.3

Series: train ARIMA(0,1,1)

Coefficients: ma1 -0.7305 s.e. 0.1400

sigma\^2 = 18.34: log likelihood = -134.93 AIC=273.86 AICc=274.13
BIC=277.56

Series: train_1.3 ARIMA(2,1,2)

Coefficients: ar1 ar2 ma1 ma2 0.6167 -0.9676 -0.8030 0.8451 s.e. 0.0500
0.0341 0.1359 0.1212

sigma\^2 = 1.161: log likelihood = -68.56 AIC=147.11 AICc=148.58
BIC=156.36

ARIMA model results from train data for ret_yr 1.3 fish in Bristol Bay
very different from ret_yr 2.2 despite both being 4 yr old fish. Ret_yr
1.3 inconsistent with results of ACF and PACF tests suggesting AR(1).
First differencing the time series object returned the mean to zero, and
auto.arima agrees first differencing the data is the best approach for
model.

Unfortunately, this is totally different from the stationarity tests
which suggest that that the data is stationary and the AICc is huge
which suggests this is not the best model for this data. Below is the
better model.

### Printed results for ret_yr 1.3 outlier removed

Series: train_1.3_out ARIMA(2,0,0) with non-zero mean

Coefficients: ar1 ar2 mean 0.3818 -0.5662 6.9256 s.e. 0.1506 0.1451
0.0890

sigma\^2 = 0.365: log likelihood = -27.2 AIC=62.41 AICc=63.94 BIC=68.14

By focusing on years without the outlier value, the model changed to
ARIMA (2,0,0) and the AICc is much lower, suggesting this model is the
better fit for the data. For the 1 year freshwater - 3 year ocean fish
we will proceed with this model.

### Printed results for ret_yr 2.2

Series: train ARIMA(1,0,0) with non-zero mean

Coefficients: ar1 mean 0.4545 5.9958 s.e. 0.1268 0.3373

sigma\^2 = 1.753: log likelihood = -80.67 AIC=167.34 AICc=167.88
BIC=172.95

ARIMA model results from train data for ret_yr 2.2 fish in Bristol Bay
consistent with results of ACF and PACF tests suggesting AR(1). First
differencing the time series object returned the mean to zero, but
auto.arima suggests 0 differencing and that data has non-zero mean.

## Test the accuracy of the model using forecast

Decided to test the model for 1 year freshwater, 3 year ocean fish using
data trained on 1980-2010 because the AICc was much better.

```{r}
fit_1.3 <- auto.arima(train_1.3_out) #fit the model
fc_1.3 <- forecast(fit_1.3, h=10)

fr_1.3 <- accuracy(forecast(fit_1.3, h=10), test_1.3_out) #test accuracy

plot.ts(datts_1.3) # check fit
lines(fitted(fit_1.3), col="red") #model seems to fit well

forecast::checkresiduals(fit_1.3) # looks better than 1.3 trained on 1960-2010

autoplot(fc_1.3) + 
  geom_point(aes(x=x, y=y), data=fortify(test_1.3_out)) 
```

### Results for 1.3 fish in Ugashik

The model itself seems to fit the data well based on the visual fit and
the residuals.

The model forecast centers around the mean of the train data, predicting
that 2010 will drop which fits the first few actual data points well,
but doesn't do as good of a job of predicting that the data from
2015-2020 look closer to the more recent years than further out (prior
to 2005). Data are included in the confidence interval though which is
great.

```{r}
fit_2.2 <- auto.arima(train_2.2) #fit the model
fc_2.2 <- forecast(fit_2.2, h=10)

fr_2.2 <- accuracy(forecast(fit_2.2, h=10), test_2.2) #test accuracy

plot.ts(datts_2.2) # check fit
lines(fitted(fit_2.2), col="red") #model seems to fit well, follows overall trend as it changes

forecast::checkresiduals(fit_2.2) 

autoplot(fc_2.2) + 
  geom_point(aes(x=x, y=y), data=fortify(test_2.2)) 
```

### Results for 2.2 fish in Ugashik

The model fits the timeseries train data pretty well, matching the
overall change in curve. The residuals show the ACF plot has 1
significant lag reflecting the ARIMA (1,0,0) model, the histogram is
fairly centered as expected, and model fails to reject Ljung-Box test.

Model forecast doesn't look great. The forecast flatlines while the real
data from 2011-2020 have much higher variance. This reflects the high
AICc from the model. But the data are contained within the predicted
confidence interval which is good.
