---
title: "Lab 4 - Hidden Markov Models"
author: "Dylan Hub, Eric French, Zoe Rand"
date: May 9, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

-   Team 1: Dylan Hubl, Eric French, Zoe Rand
-   Team 2: Maria Kuruvilla, Miranda Mudge
-   Team 3: Nick Chambers, Madison Shipley, Karl Veggerby
-   Team 4: Liz Elmstrom, Terrance Wang, Emma Timmins-Schiffman

Make sure to label your final write-up with "final" in the title so we
know which one is the final one.

```{r}
library(tidyverse)
library(depmixS4)
library(knitr)
```

# Data

As part of the California Current Integrated Ecosystem Report, NOAA
scientists do annual updates of [stoplight charts for ecosystem
indicators](https://www.fisheries.noaa.gov/west-coast/science-data/ocean-conditions-indicators-trends).

We have included the `stoplight.csv` dataset for this week. One of the
columns divides indicators into groups (e.g. Local Physical, Local
Biological, etc). Please pick a type of indicators, and develop a 2- or
3-state multivariate HMM. A few tips:

Describe what plankton groups and covariates you used and indicate any
temporal subsetting you chose. Also describe any standardization,
centering or scaling that was used.

## Load the data

One of the columns divides indicators into groups (e.g. Local Physical,
Local Biological, etc).

```{r load_data}
stoplight <- read.csv(here::here("Lab-4", "stoplight.csv"))
```

## Wrangle the data

```{r wrangle_data}
## add some code here
# we will look at Copepod_richness, N_copepod, S_copepod

df <- rbind.data.frame(stoplight[stoplight$Ecosystem.Indicators == "Copepod_richness",],stoplight[stoplight$Ecosystem.Indicators == "N_copepod",],stoplight[stoplight$Ecosystem.Indicators == "S_copepod",])
```

```{r plot_data}
newdf<-df%>%pivot_longer(-c(Ecosystem.Indicators, Type), names_to = "Year") %>%
  mutate(Year = as.numeric(gsub("X","",Year)))

ggplot(newdf) + 
  geom_line(aes(x = Year, y = value, color = Ecosystem.Indicators)) + 
  theme_classic()
```

We see that Copepod Richness and the Southern Copepod biomass measures
are in phase with one another across the plot.

```{r wrangle_data_mods}
#arranging data for model fits
df2<-df %>% dplyr::select(-Type) %>%
  gather(Year, value, -Ecosystem.Indicators) %>% 
  spread(Ecosystem.Indicators, value) 

df2$Year<-as.numeric(gsub("X","",df2$Year))
head(df2)
```

## 2-State HMM

```{r 2_state_HMM}
set.seed(123)
#set up 2-state model
mod<-depmix(list(Copepod_richness ~1, N_copepod ~1, S_copepod ~1), 
           nstates = 2, 
           #assume gaussian errors for all
           family = list(gaussian(), gaussian(), gaussian()), 
           data = df2)
#fit the model
fitmod<-fit(mod)
#model results
summary(fitmod)
```

The model results change slightly depending on the seed, so making sure
it found the best model:

```{r}
iter <-100 #number of times to run the model
seeds<-sample(100:1000, size = iter, replace = F) #get 100 seed values
best <- 1e10 #this is large so first AIC is definitely smaller
best_model <- NA
for(i in 1:iter){
  #set the seed from the ones we picked (makes sure its new each time)
  set.seed(seeds[i]) 
  #fit the model
  fitmod <- fit(mod)
  #is this model better than the last?
  if(AIC(fitmod)< best){
    best_model <- fitmod
    best <- AIC(fitmod)
  }
}

summary(best_model)
```

The transition probabilities for the best model are in the table below:

```{r}
FromS1 <- c(0.555, 0.445)
FromS2 <- c(0.217, 0.783)
tab <- rbind.data.frame(FromS1,FromS2)
colnames(tab) <- c("To S1", "To S2")
row.names(tab) <- c("From S1", "From S2")
tab
```

Plot of states and data from the 2-State HMM

```{r}
prstates<-apply(posterior(fitmod)[,c("S1", "S2")],1, which.max)
mu<-summary(best_model)[,1]
mu2<-summary(best_model)[,3]
mu3<-summary(best_model)[,5]
pred <- tibble("Year" = df2$Year, "Copepod_richness" = mu[prstates], "N_copepod" = mu2[prstates], "S_copepod" = mu3[prstates])
pred2<-pred %>% pivot_longer(-Year, names_to = "Ecosystem.Indicators", values_to = "Fit")
ggplot() + geom_point(data = newdf, aes(x = Year, y = value, color = Ecosystem.Indicators)) + 
  geom_line(data = pred2, aes(x = Year, y = Fit, group = Ecosystem.Indicators, color = Ecosystem.Indicators)) + theme_classic()


```

Our best model does converge for the two state model.

We see that the predicted fit lines follow the general trend of the
data. Southern Copepod Biomass is in phase with the Copepod Richness
measure. The observations for Northern and Southern Copepod Biomass have
much less variance than the Copepod Richness measurement and thus the
fitted lines are much closer to the observed points.

## 3-State HMM

```{r 3_state_HMM}
set.seed(123)
#set up 3-state model
mod3<-depmix(list(Copepod_richness ~1, N_copepod ~1, S_copepod ~1), 
           nstates = 3, 
           #assume gaussian errors for all
           family = list(gaussian(), gaussian(), gaussian()), 
           data = df2)
#fit the model
fitmod3<-fit(mod3)
#model results
summary(fitmod3)
```

Again, the model results change slightly depending on the seed, so
making sure it found the best model:

```{r}
iter <-100 #number of times to run the model
seeds<-sample(100:1000, size = iter, replace = F) #get 100 seed values
best <- 1e10 #this is large so first AIC is definitely smaller
best_model <- NA
for(i in 1:iter){
  #set the seed from the ones we picked (makes sure its new each time)
  set.seed(seeds[i]) 
  #fit the model
  fitmod3 <- fit(mod3)
  #is this model better than the last?
  if(AIC(fitmod3)< best){
    best_model <- fitmod3
    best <- AIC(fitmod3)
  }
}

summary(best_model)
```

The transition probabilities for the 3-State model are in the table
below

```{r}
FromS1 <- c(0.400, 0.200, 0.400)
FromS2 <- c(0.000, 0.700, 0.300)
FromS3 <- c(0.250, 0.375, 0.375)
tab2 <- rbind.data.frame(FromS1,FromS2,FromS3)
colnames(tab2) <- c("To S1", "To S2", "To S3")
row.names(tab2) <- c("From S1", "From S2", "From S3")
tab2
```

Plot of states and data from the 3-State HMM

```{r}
prstates<-apply(posterior(fitmod3)[,c("S1", "S2","S3")],1, which.max)
mu<-summary(best_model)[,1]
mu2<-summary(best_model)[,3]
mu3<-summary(best_model)[,5]
pred <- tibble("Year" = df2$Year, "Copepod_richness" = mu[prstates], "N_copepod" = mu2[prstates], "S_copepod" = mu3[prstates])
pred2<-pred %>% pivot_longer(-Year, names_to = "Ecosystem.Indicators", values_to = "Fit")
ggplot() + geom_point(data = newdf, aes(x = Year, y = value, color = Ecosystem.Indicators)) + 
  geom_line(data = pred2, aes(x = Year, y = Fit, group = Ecosystem.Indicators, color = Ecosystem.Indicators)) + theme_classic()

# The 3-State HMM seems to fit the data worse than the 2-State HMM
```

The best model does converge for the 3 state model.

The predicted values do not appear to fit the data well in the three
state model. The Northern and Southern Copepod predicted values seem to
switch states away from their respective observations in a few places.
And the Copepod Richness predicted values do not drop down to the lowest
state in a few sections where it seems like it would be appropriate to
do so.

One thing to note is that the transition probabilities state that there
is a 0.00 chance of switching from state 2 to state 1. State 2 being the
lowest level for each of the fitted lines and State 1 being the highest.
We do see that the fitted values do make this transition so this is
apparently just a very small number that is left at 0.00 due to rounding
in the table.

# Results

```{r}
`3-State` <- AIC(fitmod3)
`2-State` <- AIC(fitmod)
d <- c(`3-State`,`2-State`)-min(`3-State`,`2-State`)

e <- cbind.data.frame(rbind(`3-State`,`2-State`),d)
colnames(e) <- c("AIC","delta AIC")
row.names(e) <- c("3-State","2-State")
e
```

The three state model has the lower AIC value despite appearing to not
fit the data as well.

# Discussion

When looking at the three Copepod measurements, both of our models
converge. We did not use any covariates in order to keep our models
simple. The AIC score indicates that the 3-state model is better
supported than the 2-state. However, visually it appears that the two
state model fits the data better. Additionally, when thinking of a
biological reason for selecting a 2 vs 3 state model, it seems to make
more sense to go with a two state model as we see on the NOAA website
that copepods richness gradually increases as the summer starts and then
decreases at the end of the summer. There is a very brief plateau in May
but the 3 state model doesn't seem to follow that general pattern,
instead it seems to make large random jumps. At one point Copepod
Richness drops to the lowest state but there is no observed data there.
In contrast, the 2 state model seems to follow the general pattern of
the observed data quite well.

We see that Southern Copepod Biomass is in phase with the Copepod
Richness measurement in both models. This is because there are more
species representing the Southern Copepod group than the Northern so
when the Southern Copepod biomass is high the total richness is higher
as well due to the increased number of species present. It also makes
sense that the Northern and Southern groups are out of phase as two
different water currents which are opposed to one another carry the
copepods from each group.

All three sets of observations are tied to the same transition
probabilities and therefore the model must balance each set of
observation data to estimate the probabilities. At each data point the
model gives decides based on the probabilities whether the fitted line
should change states. Sometimes the model makes a switch that does not
seem appropriate based on those probabilities. This is why we see that
some of the transition switches in the 3 state model don't seem to make
sense.

# Team contributions

Dylan and Zoe did the data wrangling. Zoe coded the 2 state model. Eric
adapted this code for the 3 state model. Dylan created tables, results,
and led discussion section for group. Entire group contributed thoughts
and comments to the discussion section.
