---
title: "Lab 3 - Common trends in plankton data"
subtitle: "Dynamic Factor Analysis (DFA)"
author: "Person 1, Person 2, Person 3"
date: April 20, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

***

# General tasks

Each group has the same general tasks, but you will adapt them as you work on the data.

1) Find the most parsimonious model among a set that examines the effects of environmental covariates and/or an indicator of seasonality, (Dylan did this)

Liz (varying numbers of trends, and different forms of variance-covariance matrices for the observation errors, compare AIC)

Another set of model selection that varies the different covariates. (Dylan, compare AIC)

2) Plot trends and individual loadings for the model identified in task (1) above. (Code from Lab book, Dylan has written, just need to apply to best model)

3) Plot model fits and uncertainty intervals for the model identified in task (1) above. (Model automatically does this? Already have the code, just need to apply it to the best model)

4) Describe the effects of environmental or dummy variables on (possibly seasonal) patterns in the data. (D matrix, mod.fit$D, make this into a table or figure, Liz might have?)


# Methods

Please address the following in your methods:

* Which plankton taxa did you choose and how did you choose them?
   
* What time period(s) did you examine and why?

* What environmental or dummy variables did you include and why?

* What forms of models did you fit (ie, write them out in matrix form)?

* What sort of model diagnostics did you use to examine model assumptions?

# Data

Describe what plankton groups and covariates you used and indicate any temporal subsetting you chose. 

Notes from Dylan... 

I took a look at the data for lab 3 and I think we should make a couple adjustments to our original plan to avoid having large gaps in our data. I think we should start our dataset in 1967 and end in 1986. This will avoid having any NA values in the covariates which would require some extra work for us to deal with. I'm not as committed to the extra ten years added on I just thought maybe a longer time series would give the models more data to use to get better confidence interval estimates. 

We had originally selected Cryptomonas, diatoms, greens, daphnia, and cyclops as our focus. I think we should drop Cryptomonas and Daphnia and replace them with Unicells and non-colonial-rotifers. Crypto and daphnia both have a lot of missing values early in the dataset. Unicell and rotifers still meet the edible and grazer categories we were using.

## Load the data

```{r load_data}
## load MARSS for data and analyses
library(MARSS)
library(dplyr)
library(tidyverse)

## load the raw data (there are 3 datasets contained here)
data(lakeWAplankton, package = "MARSS")

## we want `lakeWAplanktonTrans`, which has been transformed
## so the 0's are replaced with NA's and the data z-scored
all_dat <- lakeWAplanktonTrans
```

## Explore the data

* You can choose any taxa to include in your analyses, but make sure there are at least 5 groups. You can mix phytoplankton and zooplankton, if you'd like, but justify any choices you make. 

* You will want to subset the data to a time period with adequate observations for the taxa of interest. Your time window should include at lest 5 years (60 months).

```{r explore_data}
## add some code here

cat("colnames: ", colnames(all_dat), "\n")

#Phytoplankton
phyto_long <- as_tibble(all_dat) %>%
  pivot_longer(values_to = 'value', cols = Cryptomonas:Other.algae)

ggplot(phyto_long, aes(x = Year, y= value), color=name) +
  geom_point(size=2, pch= 21, na.rm = TRUE) +
  theme_bw()+
  facet_wrap(~name)+ ggtitle('Phytoplankton raw data')

#Zooplankton
zoop_long <- as_tibble(all_dat) %>%
  pivot_longer(values_to = 'value', cols = Conochilus:Non.colonial.rotifers)

ggplot(zoop_long, aes(x = Year, y= value), color=name) +
  geom_point(size=2, pch= 21, na.rm = TRUE) +
  theme_bw()+
  facet_wrap(~name)+ ggtitle('Zooplankton raw data')

#Covariates
covar_long <- as_tibble(all_dat) %>%
  pivot_longer(values_to = 'value', cols = Temp:pH)

ggplot(covar_long, aes(x = Year, y= value), color=name) +
  geom_point(size=2, pch= 21, na.rm = TRUE) +
  theme_bw()+
  facet_wrap(~name)+ ggtitle('Covariates raw data')

```

## Wrangle the data

We want diatoms, greens, unicells, cyclops and non-colonial rotifers. 
``` {r}
# ## Selecting data
# cols_keep <- c('Year', 'Month' ,'Temp', 'TP' ,'pH'  ,'Diatoms', 'Greens'  ,'Unicells'  ,'Cyclops' ,'Non.colonial.rotifers')
# 
# our_dat <- as_tibble(all_dat) %>%
#   select(all_of(cols_keep))%>%
#   filter(Year > 1966 & Year < 1987)%>%
#   as.matrix()

#we only want 5 of them
crit <- all_dat[c(61:300),c(1:5,7,8,10,13,20)]

head(crit)

plankton_ts <- t(crit[,6:10])
colnames(plankton_ts) <- crit[,1]

#get the number of time series (y)
n_ts <- nrow(plankton_ts)
#get length of time series
TT <- ncol(plankton_ts)

#find the mean of each time series and remove it
y_bar <- apply(plankton_ts, 1, mean, na.rm = TRUE)
plankton_av <- plankton_ts - y_bar

plankton_z <- zscore(plankton_ts)

```

# DFA Model Selection

## Step 1- Fit global model with all covariates and number of trends

We are keeping R to diagonal and equal based on the sample design. The default DFA setting within MARSS sets our Z matrix, A to scaling and Q to identity. 

``` {r}
#set up observation model parameters

#covariates
#create a matrix for each of the covariates
covar_ts <- t(crit[,3:5])
colnames(covar_ts) <- crit[,1]

covar_z <- zscore(covar_ts) ## If we zscore the response, do we also zscore covariates?

#make a season matrix
cos_t <- cos(2 * pi * seq(TT)/12)
sin_t <- sin(2 * pi * seq(TT)/12)
season <- rbind(cos_t,sin_t)
#dim(season)

d <- rbind(covar_ts,season)
# In this case, we are not altering the upper D matrix, so every covariate will have its own effect

#tell MARSS what values to use to start with
mm <- 3
init_list <- list(x0 = matrix(rep(0, mm), mm, 1))

#Setting latent trends to 3 and R to diagonal and equal
mod_list = list(m = 3, R = "diagonal and equal", A = "zero") # If we zscore, A goes to zero (Also prob the same for demean? Not sure.)

# Model iterations
cont_list <- list(maxit = 3000, allow.degen = TRUE)

## Fit global modal, using the zscored plankton data, right now covariates are not Z-scored
dfa_global <- MARSS(plankton_z, model = mod_list, control = cont_list, inits = init_list, form = "dfa",
                    z.score = FALSE, covariates = d)

```

## Step 2- Testing number of trends

Fitting full DFA model of covariates and testing for different latent trends. 

```{r}

## Liz's original for loop
mod_list = list(R = "diagonal and equal", A = "zero") # All other parameters are the default

m_list <- 1:3 # Latent trends to loop through
  
out.tab <- NULL
fits <- list()
for(i in 1:length(m_list)){
      fit.model = c(list(m=m_list[[i]]), mod_list) ## model list to loop through
      
      fit = MARSS(plankton_z, model = fit.model, control = cont_list, 
                  form = "dfa", z.score = FALSE, covariates = d)
      
      out=data.frame(
                   m=m_list[[i]],logLik=fit$logLik, AICc=fit$AICc, num.param=fit$num.params,
                   num.iter=fit$numIter, converged=!fit$convergence,
                   stringsAsFactors = FALSE)
      
      out.tab=rbind(out.tab,out)
      fits=c(fits,list(fit))
} 

min.AICc <- order(out.tab$AICc)
out.tab.1 <- out.tab[min.AICc, ]
out.tab.1 <- cbind(out.tab.1, delta.AICc = out.tab.1$AICc - out.tab.1$AICc[1])
out.tab.1
```

Step 3 of model selection shows that we have three (at least) latent trends. So far our best model is still the global model. 

We can alter the for loop to test for different R structures.... But not sure that other structures make sense.

```{r}
## for loop testing m and R adapted from the user manual

m_list <- 1:3 # Latent trends to loop through
R_list <- list("diagonal and equal", 'diagonal and unequal')# just tested two options for R, still took a long time

model.data <- data.frame(stringsAsFactors = FALSE)
for (R in R_list) {
  for (m in 1:length(m_list)) {
    dfa.model <- list(R = R, m = m, A = "zero") ## if we zscore or demean A should go to zero... 
      
    fit = MARSS(plankton_z, model = dfa.model, control = cont_list, 
                  form = "dfa", z.score = FALSE, covariates = d)
    
    
  model.data <- rbind(model.data, data.frame(R = R, m = m,
                      fit$logLik, AICc=fit$AICc, num.param=fit$num.params,
                      stringsAsFactors = FALSE))
  }
}
model.data


```

Diagonal and unequal tests as a better R matrix, but not sure whether we justify trying this based on sample design? We could just show it and then say we stuck with R = diagonal and equal

## Step 3- Testing covariates 

STOP this chunk doesn't fully run. 

Just started to paste some of Lily's code into here in case. But idk might not be worth it. 

```{r}

## trying to figure out an option to loop through different combinations of covariates

RRR <- 'DE' #'UNC'
mmm = 3
cov1 = c(NA,"pH")
cov2 = c(NA, "Temp")
cov3 = c(NA, "TP")
seas = c(NA, "seas")

parameter_options = expand.grid(RRR = RRR, mmm = mmm, cov1=cov1,cov2=cov2, cov3 = cov3, seas=seas)
parameter_options

## This was me just copying and pasting Lily's dfa for loop for testing different covariates. She only tested one and the effect of season so I lost speed once I realized I'd have to figure out how to test 2 more. She also runs it in tmb instead of MARSS which we don't do
for(i in (1:nrow(parameter_options))){
  
  RRR = parameter_options$RRR[i]
  mmm = parameter_options$mmm[i]
  sss = parameter_options$sss[i]
  cov = parameter_options$cov[i]
  
  ## Seasonality 
  seasonality <- list(fixed_factors=ccgen(sss)) # would have to change this 
  
  #create covariate matrix (d) - see function definition in section 3.1 for details
  covariates <- cov.z[[cov]]
  
  cov_and_seas <- rbind(seasonality$fixed_factors, covariates)
  
    #fit model with TMB
  if (is.null(seasonality$fixed_factors) == TRUE & is.null(covariates) == TRUE){
    dfa <- runDFA(obs=dat_z, NumStates=mmm, ErrStruc=RRR,EstCovar=FALSE)
  } else if (is.null(seasonality$fixed_factors) == FALSE & is.null(covariates) == TRUE) {
    dfa <- runDFA(obs=dat_z, NumStates=mmm, ErrStruc=RRR,EstCovar=TRUE, Covars=cov_and_seas)
  } else if (is.null(seasonality$fixed_factors) == TRUE & is.null(covariates) == FALSE) {
    dfa <- runDFA(obs=dat_z, NumStates=mmm, ErrStruc=RRR,EstCovar=TRUE, Covars=cov_and_seas, indivCovar = TRUE)
  } 
  else {
    ## Creates the seasonal data frame
    Dmat.1<- matrix(rep(0, 216),ncol=12,nrow=24)
    Dfac.1<-matrix(1:216,ncol=12,nrow=24)
    
    ## Creates the discharge time series data frame
    Dmat.2<-matrix(0,ncol=24,nrow=24)
    diag(Dmat.2)<-rep(0,24)#rnorm(nrow(Covars),0,1)
    Dfac.2<-matrix(NA,ncol=24,nrow=24)
    diag(Dfac.2)<-seq(from=217, to=240)
    
    ## Binds the two together
    Dmat = cbind(Dmat.1, Dmat.2)
    Dfac = cbind(Dfac.1, Dfac.2) %>% as.factor()
    
    dfa <- runDFA(obs=dat_z, NumStates=mmm, ErrStruc=RRR,EstCovar=TRUE, Covars=cov_and_seas, Dmat=Dmat, Dfac=Dfac)
  }
  
}
```

# Results


# Discussion


# Team contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

