---
title: "Lab 3 - Common trends in plankton data"
subtitle: "Dynamic Factor Analysis (DFA)"
author: "Nick Chambers, Terrance Wang, Zoe Rand"
date: May 2, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
bibliography: references.bib
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

------------------------------------------------------------------------

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(zoo)
```

# Data

We decided to model 4 groups of zooplankton: Cyclops, Diaptomus, Epischura, and Neomysis. We chose to test environemtnal covariates (temp, pH, TP), as well as phytoplankton abundances as covariates(cryptomonas, diatoms, greens, unicells, other.algae) since the zooplankton we chose are copepods and mysids, and therefore may be impacted by the type of phytoplankton present. We used data from the start of the time series (1962) until 1985, because the Neomysis data gets very patchy after the 1980s and we didn't want this to cause issues with model fitting.

## Load the data

```{r load_data}
## load MARSS for data and analyses
library(MARSS)

## load the raw data (there are 3 datasets contained here)
data(lakeWAplankton, package = "MARSS")

## we want `lakeWAplanktonTrans`, which has been transformed
## so the 0's are replaced with NA's and the data z-scored
all_dat <- lakeWAplanktonTrans

```

## Wrangle the data

-   You can choose any taxa to include in your analyses, but make sure there are at least 5 groups. You can mix phytoplankton and zooplankton, if you'd like, but justify any choices you make.

-   You will want to subset the data to a time period with adequate observations for the taxa of interest. Your time window should include at lest 5 years (60 months).

### Zooplankton

```{r plot_data, warning=FALSE}
## plotting the taxa of interest
all_dat %>% as_tibble() %>%
  mutate(Date = as.yearmon(paste(Year, Month), "%Y %m")) %>%
  select(c("Date", "Cyclops", "Diaptomus", "Epischura", "Neomysis")) %>% 
  pivot_longer(-c(Date), names_to = "Species", values_to = "Vals") %>%
  ggplot(aes(x = Date, y = Vals, color = Species)) + geom_point() + 
  geom_line() + facet_wrap(~Species) + scale_x_yearmon(format = "%Y") + 
  theme_minimal() + labs(y = "Abundance Index", x = "Year") + theme(legend.position =  "none")

```

```{r wrangle_data_1}
#using data up to 1985
spp<-c("Cyclops", "Diaptomus", "Epischura", "Neomysis")
dat_1985<-all_dat[all_dat[, "Year"] <= 1985,]
dat_zoo_1985<-dat_1985[, spp]
#head(dat_zoo_1985)
dat_zoo<-t(dat_zoo_1985)

yr_frst <- min(dat_1985[,"Year"])
yr_last <- max(dat_1985[,"Year"])
```

```{r wrangle_data_2}
## get number of time series
N_ts <- dim(dat_zoo)[1]

## get length of time series
TT <- dim(dat_zoo)[2] 

## mean of each taxon
y_bar <- apply(dat_zoo, 1, mean, na.rm = TRUE)

## subtract the means
dat <- dat_zoo - y_bar

## assign new row names
rownames(dat) <- spp
```

### Covariates

```{r wrangle_cov }
#temp, pH, TP, cryptomonas, diatoms, greens, unicells, other.algae 
covs<-colnames(dat_1985)[c(3:8,10:11)]
cov_dat<-t(dat_1985[,covs])
#head(cov_dat)
#dummy seasonal affect
cos_t <- cos(2 * pi * seq(TT) / 12)
sin_t <- sin(2 * pi * seq(TT) / 12)
dd <- rbind(cos_t, sin_t)

```

```{r cov_corr}
#looking at any correlation issues for covariates
pairs(dat_1985[,covs])
library(corrplot)
corrplot(cor(dat_1985[,covs], use = "pairwise.complete.obs"), addCoef.col = "blue")
```

# General tasks

Each group has the same general tasks, but you will adapt them as you work on the data.

1)  Find the most parsimonious model among a set that examines the effects of environmental covariates and/or an indicator of seasonality, varying numbers of trends, and different forms of variance-covariance matrices for the observation errors.

2)  Plot trends and individual loadings for the model identified in task (1) above.

3)  Plot model fits and uncertainty intervals for the model identified in task (1) above.

4)  Describe the effects of environmental or dummy variables on (possibly seasonal) patterns in the data.

# Methods

Please address the following in your methods:

-   Which plankton taxa did you choose and how did you choose them?

-   What time period(s) did you examine and why?

-   What environmental or dummy variables did you include and why?

-   What forms of models did you fit (ie, write them out in matrix form)?

-   What sort of model diagnostics did you use to examine model assumptions?

### Data:

We chose to model four zooplankton taxa :Cyclops, Diaptomus, Epischura, and Neomysis. We chose these because they are similar types of zooplankton (copepods and mysids) which are known to be selective about the types of phytoplankton they consume [@hansen1997]. We used data from 1962-1985, because there is very little Neomysis data after 1985. We tested environmental covariates, including temperature, pH, and phosphorus concentrations as well as a seasonal dummy variable. We were also interested in the affect of phytoplankton abundance on these species, so we tested abundance indexes of diatoms, greens, unicells, and other.algae as potential covariates as well.

### Model:

# Observation Model

## ZZ is loadings matrix
```{r}
ZZ = NULL
# Define different ZZ hypotheses 
# 1 state
Z_vals <- list("z11",
               "z21", 
               "z31",
               "z41")
ZZ[[1]] <- matrix(Z_vals, nrow = N_ts, ncol = 1, byrow = TRUE)
# 2 states
Z_vals <- list("z11",  0  , 
               "z21","z22", 
               "z31","z32",
               "z41","z42")
ZZ[[2]] <- matrix(Z_vals, nrow = N_ts, ncol = 2, byrow = TRUE)
# 3 states
Z_vals <- list("z11",  0  ,  0  ,
               "z21","z22",  0  ,
               "z31","z32","z33",
               "z41","z42","z43")
ZZ[[3]] <- matrix(Z_vals, nrow = N_ts, ncol = 3, byrow = TRUE)

# 4 states
Z_vals <- list("z11",  0  ,  0  , 0,
               "z21","z22",  0  , 0,
               "z31","z32","z33", 0,
               "z41","z42","z43","z44")
ZZ[[4]] <- matrix(Z_vals, nrow = N_ts, ncol = 4, byrow = TRUE)

## 'aa' is the offset/scaling
aa <- "zero"

## 'DD' and 'd' are for covariates
DD <- "zero"  # matrix(0,mm,1)
dd <- "zero"  # matrix(0,1,wk_last)

## 'RR' is var-cov matrix for obs errors
RR <- "diagonal and unequal"
```

# Process Model
```{r}
## number of processes
mm <- 2 # this is just a placeholder, as we loop through # of states we will update this

## 'BB' is identity: 1's along the diagonal & 0's elsewhere

BB <- "identity"  # diag(mm)

## 'uu' is a column vector of 0's
uu <- "zero"  # matrix(0, mm, 1)

## 'CC' and 'cc' are for covariates
CC <- "zero"  # matrix(0, mm, 1)
cc <- "zero"  # matrix(0, 1, wk_last)

## 'QQ' is identity
QQ <- "identity"  # diag(mm)

```


# Fit with MARSS
```{r}
## list with specifications for model vectors/matrices
mod_list <- list(Z = ZZ[[2]], A = aa, D = DD, d = dd, R = RR,
                 B = BB, U = uu, C = CC, c = cc, Q = QQ)
## list with model inits
init_list <- list(x0 = matrix(rep(0, mm), mm, 1))
## list with model control parameters
con_list <- list(maxit = 3000, allow.degen = TRUE)

```


````{r model selection of # states}
Z_AIC <- data.frame(states = NA,AICc=NA)
for (z in 1:length(ZZ)){
## list with specifications for model vectors/matrices
mod_list <- list(Z = ZZ[[z]], A = aa, D = DD, d = dd, R = RR,
                 B = BB, U = uu, C = CC, c = cc, Q = QQ)
mm <- z
## list with model inits
init_list <- list(x0 = matrix(rep(0, mm), mm, 1))
## list with model control parameters
con_list <- list(maxit = 3000, allow.degen = TRUE)

## fit MARSS
dfa <- MARSS(y = dat, model = mod_list, inits = init_list, control = con_list)
# get model results
Z_AIC[z,"states"] = z
Z_AIC[z,"AICc"] = dfa$AICc
}
Z_AIC
```

Best performing based on AICc is the one with 3 states. 

```{r}
## list with specifications for model vectors/matrices
mod_list <- list(Z = ZZ[[3]], A = aa, D = DD, d = dd, R = RR,
                 B = BB, U = uu, C = CC, c = cc, Q = QQ)
mm <- 3
## list with model inits
init_list <- list(x0 = matrix(rep(0, mm), mm, 1))
## list with model control parameters
con_list <- list(maxit = 3000, allow.degen = TRUE)


## fit MARSS
dfa_3 <- MARSS(y = dat, model = mod_list, inits = init_list, control = con_list)
```

#Rotating trends and loadings
```{r}
## get the estimated ZZ
Z_est <- coef(dfa_3, type = "matrix")$Z

## get the inverse of the rotation matrix
H_inv <- varimax(Z_est)$rotmat

#Now rotate both Z and x

## rotate factor loadings
Z_rot = Z_est %*% H_inv   

## rotate processes
proc_rot = solve(H_inv) %*% dfa_1$states
```

#Estimated States and Loadings
```{r}
## plot labels
zooplankton_names = c("Cyclops", "Diaptomus", "Epischura", "Neomysis")
ylbl <- zooplankton_names
w_ts <- seq(dim(dat)[2])

# plot colors
clr <- c("brown", "blue", "darkgreen", "darkred", "purple")

## set up plot area
layout(matrix(c(1,2,3,4,5,6), mm, 2), widths = c(2,1))
par(mai = c(0.5, 0.5, 0.5, 0.1), omi = c(0, 0, 0, 0))

## plot the processes
for(i in 1:mm) {
  ylm <- c(-1, 1) * max(abs(proc_rot[i,]))
  ## set up plot area
    plot(w_ts,proc_rot[i,], type = "n", bty = "L",
         ylim = ylm, xlab = "", ylab = "", xaxt = "n")
    ## draw zero-line
    abline(h = 0, col = "gray")
    ## plot trend line
    lines(w_ts, proc_rot[i,], lwd = 2)
    lines(w_ts, proc_rot[i,], lwd = 2)
    ## add panel labels
    mtext(paste("State",i), side = 3, line = 0.5)
    axis(1, 12 * (0:dim(dat)[2]) + 1, yr_frst + 0:dim(dat)[2])
}
## plot the loadings
minZ <- 0
ylm <- c(-1, 1) * max(abs(Z_rot))
for(i in 1:mm) {
  plot(x = c(1:N_ts)[abs(Z_rot[,i])>minZ],
       y = as.vector(Z_rot[abs(Z_rot[,i])>minZ,i]),
       type = "h",
       lwd = 2, xlab = "", ylab = "", xaxt = "n", ylim = ylm,
       xlim = c(0.5, N_ts + 0.5), col = clr)
    for(j in 1:N_ts) {
      if(Z_rot[j,i] > minZ) {text(j, -0.03, ylbl[j], srt = 90, adj = 1, cex = 1.2, col = clr[j])}
      if(Z_rot[j,i] < -minZ) {text(j, 0.03, ylbl[j], srt = 90, adj = 0, cex = 1.2, col = clr[j])}
      abline(h = 0, lwd = 1.5, col = "gray")
      } 
  mtext(paste("Factor loadings on state", i), side = 3, line = 0.5)
}

```

### Diagnostics:

# Results

# Discussion

# Team contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

# References
