---
title: "Lab 4 - Hidden Markov Models"
author: "Dylan Hub, Eric French, Zoe Rand"
date: May 9, 2023
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

-   Team 1: Dylan Hubl, Eric French, Zoe Rand
-   Team 2: Maria Kuruvilla, Miranda Mudge
-   Team 3: Nick Chambers, Madison Shipley, Karl Veggerby
-   Team 4: Liz Elmstrom, Terrance Wang, Emma Timmins-Schiffman

Make sure to label your final write-up with "final" in the title so we
know which one is the final one.

```{r}
library(tidyverse)
library(depmixS4)
```

# Data

As part of the California Current Integrated Ecosystem Report, NOAA
scientists do annual updates of [stoplight charts for ecosystem
indicators](https://www.fisheries.noaa.gov/west-coast/science-data/ocean-conditions-indicators-trends).

We have included the `stoplight.csv` dataset for this week. One of the
columns divides indicators into groups (e.g. Local Physical, Local
Biological, etc). Please pick a type of indicators, and develop a 2- or
3-state multivariate HMM. A few tips:

Describe what plankton groups and covariates you used and indicate any
temporal subsetting you chose. Also describe any standardization,
centering or scaling that was used.

## Load the data

One of the columns divides indicators into groups (e.g. Local Physical,
Local Biological, etc).

```{r load_data}
stoplight <- read.csv(here::here("Lab-4", "stoplight.csv"))
```

## Wrangle the data

```{r wrangle_data}
## add some code here
# we will look at Copepod_richness, N_copepod, S_copepod

df <- rbind.data.frame(stoplight[stoplight$Ecosystem.Indicators == "Copepod_richness",],stoplight[stoplight$Ecosystem.Indicators == "N_copepod",],stoplight[stoplight$Ecosystem.Indicators == "S_copepod",])
```

```{r plot_data}
newdf<-df%>%pivot_longer(-c(Ecosystem.Indicators, Type), names_to = "Year") %>%
  mutate(Year = as.numeric(gsub("X","",Year)))

ggplot(newdf) + 
  geom_line(aes(x = Year, y = value, color = Ecosystem.Indicators)) + 
  theme_classic()
```
We see that Copepod Richness and the Southern Copepod biomass measures are in phase with one another across the plot.
```{r wrangle_data_mods}
#arranging data for model fits
df2<-df %>% dplyr::select(-Type) %>%
  gather(Year, value, -Ecosystem.Indicators) %>% 
  spread(Ecosystem.Indicators, value) 

df2$Year<-as.numeric(gsub("X","",df2$Year))
head(df2)
```

# General tasks


# Methods

A few tips:

-   Assume all responses are Gaussian.

-   You're welcome to include covariates (year? Climate variables?) --
    but fitting a simple model without covariates is also totally fine

## 2-State HMM

```{r 2_state_HMM}
set.seed(123)
#set up 2-state model
mod<-depmix(list(Copepod_richness ~1, N_copepod ~1, S_copepod ~1), 
           nstates = 2, 
           #assume gaussian errors for all
           family = list(gaussian(), gaussian(), gaussian()), 
           data = df2)
#fit the model
fitmod<-fit(mod)
#model results
summary(fitmod)
```

The model results change slightly depending on the seed, so making sure
it found the best model:

```{r}
iter <-100 #number of times to run the model
seeds<-sample(100:1000, size = iter, replace = F) #get 100 seed values
best <- 1e10 #this is large so first AIC is definitely smaller
best_model <- NA
for(i in 1:iter){
  #set the seed from the ones we picked (makes sure its new each time)
  set.seed(seeds[i]) 
  #fit the model
  fitmod <- fit(mod)
  #is this model better than the last?
  if(AIC(fitmod)< best){
    best_model <- fitmod
    best <- AIC(fitmod)
  }
}

summary(best_model)
```

Plot of states and data from the 2-State HMM

```{r}
prstates<-apply(posterior(fitmod)[,c("S1", "S2")],1, which.max)
mu<-summary(best_model)[,1]
mu2<-summary(best_model)[,3]
mu3<-summary(best_model)[,5]
pred <- tibble("Year" = df2$Year, "Copepod_richness" = mu[prstates], "N_copepod" = mu2[prstates], "S_copepod" = mu3[prstates])
pred2<-pred %>% pivot_longer(-Year, names_to = "Ecosystem.Indicators", values_to = "Fit")
ggplot() + geom_point(data = newdf, aes(x = Year, y = value, color = Ecosystem.Indicators)) + 
  geom_line(data = pred2, aes(x = Year, y = Fit, group = Ecosystem.Indicators, color = Ecosystem.Indicators)) + theme_classic()

?depmix

```

Our best model does converge for the two state model.

We see that the predicted fit lines still follow the general trend of the data. Southern Copepod Biomass is in phase with the Copepod Richness measure. The observations for Northern and Southern Copepod Biomass have much less variance than the Copepod Richness measurement and thus the fitted lines are much closer to the observed points.


## 3-State HMM

```{r 3_state_HMM}
set.seed(123)
#set up 3-state model
mod3<-depmix(list(Copepod_richness ~1, N_copepod ~1, S_copepod ~1), 
           nstates = 3, 
           #assume gaussian errors for all
           family = list(gaussian(), gaussian(), gaussian()), 
           data = df2)
#fit the model
fitmod3<-fit(mod3)
#model results
summary(fitmod3)
```

Again, the model results change slightly depending on the seed, so making sure
it found the best model:

```{r}
iter <-100 #number of times to run the model
seeds<-sample(100:1000, size = iter, replace = F) #get 100 seed values
best <- 1e10 #this is large so first AIC is definitely smaller
best_model <- NA
for(i in 1:iter){
  #set the seed from the ones we picked (makes sure its new each time)
  set.seed(seeds[i]) 
  #fit the model
  fitmod3 <- fit(mod3)
  #is this model better than the last?
  if(AIC(fitmod3)< best){
    best_model <- fitmod3
    best <- AIC(fitmod3)
  }
}

summary(best_model)
```

Plot of states and data from the 3-State HMM

```{r}
prstates<-apply(posterior(fitmod3)[,c("S1", "S2","S3")],1, which.max)
mu<-summary(best_model)[,1]
mu2<-summary(best_model)[,3]
mu3<-summary(best_model)[,5]
pred <- tibble("Year" = df2$Year, "Copepod_richness" = mu[prstates], "N_copepod" = mu2[prstates], "S_copepod" = mu3[prstates])
pred2<-pred %>% pivot_longer(-Year, names_to = "Ecosystem.Indicators", values_to = "Fit")
ggplot() + geom_point(data = newdf, aes(x = Year, y = value, color = Ecosystem.Indicators)) + 
  geom_line(data = pred2, aes(x = Year, y = Fit, group = Ecosystem.Indicators, color = Ecosystem.Indicators)) + theme_classic()

# The 3-State HMM seems to fit the data worse than the 2-State HMM
```
The best model does converge for the 3 state model.

The predicted values do not appear to fit the data well in the three state model. The Northern and Southern Copepod predicted values seem to switch states away from their respective observations in a few places. And the Copepod Richness predicted values do not drop down to the lowest state in a few sections where it seems like it would be appropriate to do so. 

One thing to note is that the transition probabilities state that there is a 0.00 chance of switching from state 2 to state 1. State 2 being the lowest level for each of the fitted lines and State 1 being the highest. We do see that the fitted values do make this transition so this is apparently just a very small number that is left at 0.00 due to rounding in the table.

# Results
```{r}
`3-State` <- AIC(fitmod3)
`2-State` <- AIC(fitmod)
d <- c(a,b)-min(a,b)
library(knitr)
e <- cbind(rbind(`3-State`,`2-State`),d)
kable(e, format = "simple", row.names = TRUE, col.names = c("AIC", "deltaAIC"))

```

The three state model has the lower AIC value despite appearing to not fit the data as well. 
Summarize the model you've created. Specifically,

-   Does it converge?

-   How many states seem to be most supported?

-   Plot the time series of estimated states. What does this mean?

-   What are the transition probabilities?

-   If included, what are the covariate effects? What do these mean?

-   Anything else interesting that you've discovered?

# Discussion
When looking at the three Copepod measurements, both of our models converge. We did not use any covariates in order to keep our models simple. The AIC score indicates that the 3-state model is better supported than the 2-state. However, visually it appears that the two state model fits the data better. Additionally, when thinking a biological reason for selecting a 2 vs 3 state model, it seems to make more sense to go with a two state model as we see on the NOAA website that copepods richness gradually increases as the summer starts and then decreases at the end of the summer. There is a very brief plateau in May but the 3 state model doesn't seem to follow that general pattern, instead it seems to make large random jumps. At one point Copepod richness drops to the lowest state but there is no observed data there. In contrast, the 2 state model seems to follow the general pattern of the observed data quite well.

We see that Southern Copepod Biomass is in phase with the Copepod Richness measurement in both models. This is because there are more species representing the Southern Copepod group than the Northern. It also makes sense that the Northern and Southern groups are out of phase as two different water currents which are opposed to one another carry in the copepods from each group.
# Team contributions

Example: "All team members helped decide on the goal and ran the
analyses for the individual regions. Team members 2 & 3 wrote most of
the code for the analysis. Team member 1 worked on the plotting section
of the report using and adapting code that team member 3 wrote. All team
members helped edit the report and wrote the discussion together."

#Notes (GET RID OF): Tasks: 1) Fit 2 state (Zoe) 2) Fit 3 state (Eric)
3) Results (transition probabilities, time series of states, number of
states) (Dylan) 4) Discussion (Together)
